### 2011-05-04 ctw2011_abstract/CTW2011.tex
added abstract and submitted PDF of this abstract for the CTW'2011 workshop

\documentclass{ctwart}
\usepackage{amssymb}
\usepackage{color}
% \usepackage{dsfont}
\usepackage[numbers,sort&compress]{natbib}

\newcommand{\TODO}[1]{{\color{red}TODO: #1}}
\newcommand{\TODON}[1]{{\color{blue}TODO: #1}}
\newcommand{\COMMENTD}[1]{{\color{green}Dimo: #1}}
\newcommand{\COMMENTN}[1]{{\color{green}Nora: #1}}

\newcommand{\mypara}[1]{{\bfseries #1: }}

\begin{document}

\begin{frontmatter}

\title{An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem}

\author{Nora Touati-Moungla and Dimo Brockhoff}

\address{Laboratoire d'Informatique, \'{E}cole Polytechnique, 91128 Palaiseau Cedex, France}

\begin{keyword}
Evolutionary multiobjective optimization, Transportation of hazardous materials, Risk equity, Shortest Paths. 
\end{keyword}

\end{frontmatter}

\section{Introduction}

The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \cite{CAR08}. The main target of this problem is to select routes from a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized---without producing excessive economic costs. 
%
When solving such a problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions are not affected. In this case, one may wish to distribute the risk in an equitable way over the population and the environment. Several studies consider this additional minimization of the equity risk, but most of them consist of a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{AKG02, CAR08}).
% 
% 
% and can be classified into the \textit{resolution-equity-based methods} and the \textit{model-equity-based methods}. In resolution-equity-based methods, the equity constraints are taken into account only indirectly via a dissimilarity index between paths which is to be maximized \cite{JOH92, LOM93, KUB97, AKG02}. Model-equity-based methods, on the other hand, formalize the risk equity as constraints directly within the model. In \cite{GOP90a, GOP90b}, for example, the authors propose an equity shortest path model that minimizes the total risk of travel while the difference between the risks imposed on any two arbitrary zones does not exceed a given threshold.
%
A more realistic \emph{multi}-commodity flow model was proposed in \citep{CAR08} where each commodity is considered as one hazmat type. The objective function is formulated as the sum of the economical cost and the cost related to the consequences of an incident for each material. To deal with risk equity, the costs are defined as functions of the flow traversing the arcs which imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc.

The majority of all hazmat routing studies deal with a single-objective scenario although the problem itself is multiobjective in nature and it is important to study the trade-offs among the objectives. Evolutionary Multiobjective Optimization (EMO) algorithms are able to compute a set of solutions showing these trade-offs within a single algorithm run which is the reason why we propose to use them for the problem of hazmat routing in this study (Sec.~\ref{S_MEO}). Before, we formalize the routing of hazmat problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{CAR08} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).

% see for example \cite{CAR08, cgr2007, gkbk1990, za2004}, but all problem formulations include the equity risk as constraints and not as an additional objective. \citet{gian1998} on the other hand considers the equity risk as separate objective but uses a weighted sum to aggregate his four objectives into a single-objective problem. The main reason why, to the best of our knowledge, no study considered a fully multiobjective model with the equity risk as objective might be the resulting non-linearity of the problem which make classical optimization algorithms such as branch-and-cut \TODON{equity is an objective in \cite{{gian1998}} ... We solve the broblem with a branch and price algorithm because the model is a multicommodity flow model, and this model is the more adapted one for this problem because it permit to manage simultaneously many hazmat types. The min-max formulation of the objective is easily linearized. I give you the major axes of the work: (1) Solve the presented problem (optimize cost, risk and equity) (2) what are models and resolution methods proposed in the litterature (3) what is the model retained and why (in our case, the multicommodity model) (4) how to handle equity with this model ? (5) how to solve it ? (branch and price is adapted for these models, but can not hadle many objectives $->$ evolutionnary algorithms) } infeasible and led us to our choice of proposing an evolutionary multiobjective algorithm as heuristic here.

%The computation of routes with a fairly distributed risk consists in generating dissimilar origin-destination paths, i.e paths which relatively don't impact the same zones. Solution approaches

%Section~\ref{S_RECRP} defines the problem of transporting hazardous materials with equity risk as a three-objective optimization problem with the minimization of the total routing cost, of the total routing risk {\em and} of the risk equity as the objectives---the latter broadly defined as the risk shared by a set of regions that compose the geographical area under consideration. Based on this formulation, Sec.~\ref{S_MEO} presents an evolutionary multiobjective optimization algorithm before Sec.~\ref{S_FW} concludes the abstract.

%In \cite{CAR08} was proposed a multi-commodity flow model for this problem, where each commodity is considered as one hazmat type. To deal with risk equity, the costs are defined as functions of the flow traversing the arcs, this imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc. The major drawback of this model is that it can lead to the saturation of the arcs of shortest paths and, therefore, does not guarantee the equity target.

%In \cite{BEL10}, the authors consider a similar model but with additional equity constraints and objectives. The objective is the minimization of the maximum risk equity imposed on the population and the environment and the model is solved using a pranch-and-price algorithm.

%As hazmat routing is multiobjective in nature, we introduce in this work a multiobjective version of this problem. Our goal is the minimization of the total routing cost, of the total routing risk {\em and} of the risk equity, the latter broadly defined as the risk shared by a set of regions that compose the geographical area under consideration.

%This paper is organized as follows. We describe the problem in section \ref{S_RECRP}, we give its mathematical formulation in section \ref{S_AOM}, in section \ref{S_MEO} we show how multiobjective evolutionary methods can be applied to this problem and in section \ref{S_FW} we exopse our further work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The multiobjective risk-equity constrained routing problem}
\label{S_RECRP}

Let the transportation network be represented as a directed graph $G = (N, A)$, with $N$ being the set of nodes and $A$ the set of arcs. Let $C$ be the set of commodities, given as a set of point-to-point demands to transport a certain amount of hazmats. For any commodity $c\in C$, let $s^c$ and $t^c$ be the source node and the destination node respectively, and let $V^c$ be the number of available trucks for the transportation of commodity $c$. Each commodity is associated with a unique type of hazmat. We assume that the risk is computed on each arc of the network and is proportional to the number of trucks traversing such an arc. We consider a set $Q$ of regions, and we define $r_{ij}^{cq}$ as the risk imposed on region $q \in Q$ when the arc $(i,j) \in A$ is used by a truck for the transportation of hazmat of type $c$. We remark that we employ a notion of {\em spread risk}, in that an accidental event on arc $(i,j)$ within region $q\in Q$ may strongly affect another region $q'\in Q$. With each arc $(i,j) \in A$ a cost $c_{ij}^c$ is associated, involved by the travel of a truck of commodity $c$ on this arc.

The problem of transporting hazmat is multiobjective in nature: one usually wants to minimize the \textit{total cost} of transportation, the \textit{total risk} of transportation imposed on all regions and the {\em distributed risk}, which can be defined as a measure of risk that is shared among different regions. More specifically, for a given solution, each region $q\in Q$ will be affected by a risk $\omega_q$ which depends on the transportation patterns in all other regions. The third objective will then be $\max_{q\in Q} \omega_q$, and has to be minimized.

We introduce the integer variable $y_{ij}^c$ for the number of trucks that use arc $(i,j)$ for transporting commodity $c$. We assume a fixed number of trucks and that all trucks have the same load. The proposed model is defined as follows:
\begin{eqnarray}
   \min  & \sum_{c \in C} \sum_{(i,j) \in A} c_{ij}^{c} y_{ij}^c & \label{eq:obj1}\\
   \min  & \sum_{c \in C} \sum_{q\in Q} \sum_{(i,j) \in A} r_{ij}^{cq} y_{ij}^c & \label{eq:obj2}\\
   \min  & \max_{q\in Q} \big\{ \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} y_{ij}^c \big\} & \label{eq:obj3}\\
    s.t. & \sum_{j \in \delta^+(i)} y_{ij}^c -  \sum_{j \in \delta^-(i)} y_{ji}^c = q_i^c\quad         & \forall i \in N,  c \in  C \label{eq:flow} \\
         & y_{ij}^c \in \{0, 1, 2, \ldots\}                                                            & \forall (i,j)\in A, c\in C\label{eq:integerY}.
\end{eqnarray}

The first objective in (\ref{eq:obj1}) is a cost function and is to be minimized. The second objective, given by (\ref{eq:obj2}), minimizes the total risk on all regions and objective (\ref{eq:obj3}) minimizes the maximum risk imposed on all regions. Constraints (\ref{eq:flow}) are conservation constraints, where $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$ are the direct successors and predecessors of node $i$, and $q_i^c=V^c$ if  $i=s^c$, $q_i^c=-V^c$ if  $i=t^c$ and $q_i^c=0$ otherwise.

\section{Evolutionary multiobjective optimization}
\label{S_MEO}
Evolutionary algorithms (EAs) and Evolutionary Multiobjective Optimization (EMO) algorithms in particular are general-purpose randomized search heuristics and as such well suited for problems where the objective function(s) can be highly non-linear, noisy, or even given only implicitly, e.g., by expensive simulations \citep{deb2001a,cvl2007a}. Since the third objective in the above problem formulation is nonlinear, we propose to use an EMO algorithm for the multiobjective risk-equity constrained routing problem here. Our EMO algorithm follows the standard iterative/generational cycle of an EA of mating selection, variation, objective function evaluation, and environmental selection and is build upon the state-of-the-art selection scheme in HypE \citep{bz2011a} as implemented in the PISA platform \citep{bltz2003a}. The variation operators as well as the representation of the solutions, however, have to be adapted to the problem at hand in the following way in order to fulfill the problem's constraints at all times.

\mypara{Representation}
We choose a variable length representation as it has been theoretically shown to be a good choice for multiobjective shortest paths problems \citep{horo2009a}: A solution is thereby represented by a list of paths of variable lengths with one path per truck. For the moment, we consider a fixed amount of trucks for each commodity and therefore a fixed number of paths through the network. In order to have every variable length path represent an uninterrupted path from source to destination at any time (see the constraints in (\ref{eq:flow})), we ensure all paths to always start with the source $s^c$ for the corresponding commodity $c$, ensure with the variation operator that all neighbored vertices in the path are connected by an arc, and complete each path by the shortest path between the path's actual end node and the commodity's destination node $t^c$.

\mypara{Initialization}
Initially, we start with a single solution where the paths $p$ for all trucks are empty ($p=(s^c)$). This corresponds to the situation where all trucks choose the shortest $s^c$--$t^c$ path for their assigned commodity---implying the smallest possible overall cost but a high risk along the used route(s). Nevertheless, the initial solution is already Pareto-optimal and is expected to be a good starting point for the algorithm.

\mypara{Variation}
As mutation operator, we suggest to shorten or lengthen the path of one or several trucks. In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as in \citep{horo2009a}:
\begin{itemize}
	\item if $b = 0$ and $l=:\textnormal{length}(p)\geq 2$, set $p' = (s^c, \ldots, v_{l-1})$	
	\item if $b = 1$ and $|V_{\textnormal{\scriptsize rem}}=\{v \in V \,|\, (v_{l}, v) \in A\}| \neq\emptyset$, choose $v_{l+1}$ from $V_{\textnormal{\scriptsize rem}}$ uniformly at random and set $p' = (v_0, \ldots, v_l, v_{l+1})$.
	\item otherwise, use the same path $p$ also in the new solution $s'$.
\end{itemize}

\section{Conclusions}
\label{S_FW}
The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest. Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary algorithm to cope with the non-linear equity risk objective.

The third objective function of our problem can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$. Although this equivalent formulation makes the problem linear (with additional linear constraints), classical algorithms are expected to have difficulties with this formulation as well and our algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints. Note that, for the moment, the proposed EMO algorithm exists on paper only and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to generate solutions of sufficient quality and whether adaptively changing the number and capacity of trucks is beneficial.

% The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest.

% In this study, we formalize transportation of hazmats problem as the minimization of three objectives and propose to use an evolutionary algorithm since many exact optimization procedures have difficulties with the non-linear equity risk objective. Note that our problem formulation can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$ \COMMENTD{this has to be adapted according to the new problem formulation}. Although this equivalent formulation makes the problem linear (with additional non-linear constraints), classical algorithms are expected to have difficulties with this formulation as well and the EMO algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints.
% 
% Note also that, for the moment, the proposed EMO algorithm only exists as a gedankenexperiment and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to improve the quality of the found solutions to a sufficient level.

% 
 \bibliographystyle{plainnat}
 \footnotesize
% \bibliography{hazmat}

\begin{thebibliography}{00}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Akgun et~al.(2003)Akgun, Erkut, and Batta]{AKG02}
V. Akgun, E. Erkut and R. Batta, 
On finding dissimilar paths, 
European Journal of Operational Research 121(2):232-246, 2000.

\bibitem[Bader and Zitzler(2011)]{bz2011a}
J.~Bader and E.~Zitzler.
\newblock Hype: An algorithm for fast hypervolume-based many-objective
  optimization.
\newblock \emph{Evolutionary Computation}, 19\penalty0 (1):\penalty0 45–--76,
  2011.

\bibitem[Bleuler et~al.(2003)Bleuler, Laumanns, Thiele, and Zitzler]{bltz2003a}
S.~Bleuler, M.~Laumanns, L.~Thiele, and E.~Zitzler.
\newblock PISA---a platform and programming language independent interface for
  search algorithms.
\newblock In \emph{Evolutionary
  Multi-Criterion Optimization {(EMO~2003)}}, pages
  494--508, 2003. Springer.

\bibitem[Caramia et~al.(2008)Caramia and Dell'Omo]{CAR08}
M. Caramia and P. Dell'Olmo, 
Multiobjective management in freight logistics: increasing capacity, service level and safety with optimization algorithms,
Springer London Ltd, 2008.

\bibitem[{Coello Coello} et~al.(2007){Coello Coello}, {Lamont}, and {Van
  Veldhuizen}]{cvl2007a}
C.~A. {Coello Coello}, G.~B. {Lamont}, and D.~A. {Van Veldhuizen}.
\newblock \emph{Evolutionary Algorithms for Solving Multi-Objective Problems}.
\newblock Springer, 2007.

\bibitem[Deb(2001)]{deb2001a}
K.~Deb.
\newblock \emph{Multi-Objective Optimization Using Evolutionary Algorithms}.
\newblock Wiley, Chichester, UK, 2001.

\bibitem[Horoba(2009)]{horo2009a}
C.~Horoba.
\newblock Analysis of a simple evolutionary algorithm for the multiobjective
  shortest path problem.
\newblock In \emph{Foundations of Genetic Algorithms {(FOGA 2009)}}, pages
  113--120. ACM, 2009.

% \bibitem{GOP90b}
% R. Gopalan, K.S. Kolluri, R. Batta and M.H. Karwan, 
% Modeling equity of risk in the transportation of hazardous materials, 
% Operations Research, 38(6):961-973, 1990.
% 
% \bibitem{KUB97}
% M. Kuby, X. Zhongyi and X. Xiaodong, 
% A minimax method for finding the $k$ best differentiated paths, 
% Geographical Analysis 29(4):298-313, 1997.

% \bibitem{LOM93}
% K. Lombard and R.L. Church, 
% The gateway shortest path problem: Generating alternative routes for a corridor location problem, 
% Geographical Systems, 1:25-45, 1993.

\end{thebibliography}

\end{document}

### 2011-10-17 trunk/code/variatorHazmatJava/src/graph/Commodity.java
The code is modified to take into account the new format of data

@@ -6,21 +6,28 @@ public class Commodity {
 	private int source;
 	private int dest;
 	private int demand;
-	private int tCap;
-	private int nbTrucks;
+	//private int tCap;
+	//private int nbTrucks;
 	
-	public Commodity(int n, int s, int des, int dem, int t, int nb){
+	//public Commodity(int n, int s, int des, int dem, int t, int nb){
+	//this.num = n;
+	//this.source = s;
+	//this.dest = des;
+	//this.demand = dem;
+	//this.tCap = t;
+	//this.nbTrucks = nb;
+	//}
+	
+	public Commodity(int n, int s, int des, int dem){
 		this.num = n;
 		this.source = s;
 		this.dest = des;
 		this.demand = dem;
-		this.tCap = t;
-		this.nbTrucks = nb;
 	}
 	
-	public int getNbTrucks() {
-		return this.nbTrucks;
-	}
+	//public int getNbTrucks() {
+	//return this.nbTrucks;
+	//}
 	
 	public int getSource() {
 		return this.source;
@@ -38,8 +45,12 @@ public class Commodity {
 	 * Returns a (deep) copy of itself.
 	 * @return deep copy
 	 */
+	//public Commodity getCopy() {
+	//	return new Commodity(this.num, this.source, this.dest, this.demand, this.tCap, this.nbTrucks);
+	//}
+	
 	public Commodity getCopy() {
-		return new Commodity(this.num, this.source, this.dest, this.demand, this.tCap, this.nbTrucks);
+		return new Commodity(this.num, this.source, this.dest, this.demand);
 	}
 }
 

### 2011-10-17 trunk/code/variatorHazmatJava/src/graph/Graph.java

@@ -6,11 +6,12 @@ import general.Variator;
 import java.util.*;
 import java.io.*;
 import java.lang.Integer;
+import java.lang.Runtime;
 import java.util.Scanner;
 import java.util.regex.Pattern;
 import population.*;
 import java.io.*;
-
+import java.lang.*;
 
 
 
@@ -20,7 +21,7 @@ public class Graph {
 	    public int nbNodes; 
 		public int nbArcs; 
 		public int nbCom; 
-		public int nbReg; 
+		//public int nbReg; 
 		
 		public ArrayList<Node> vectNodes; 
 		public ArrayList<Arc> vectArcs;   
@@ -51,23 +52,30 @@ public class Graph {
    				
    				nbCom = sc.nextInt();
    				System.out.println("c = " + nbCom);
-   				nbReg = sc.nextInt();
-   				System.out.println("r = " + nbReg);
+   				//nbReg = sc.nextInt();
+   				//System.out.println("r = " + nbReg);
    				
-   				sc.next();
-   				sc.next();
+   				//sc.next();
+   				sc.next();sc.next();sc.next();sc.next();
    				
    				// Read Arcs
    				int i = 0;
    				int j = 0;
    				int nbA = 0;   // The number of arcs 
    				while (sc.hasNextInt()){
+   					System.out.println("New arc \n");
    					//buffer = sc.next();
    					i = sc.nextInt();
+   					System.out.println("Noeud origine "+i);
    					Node ni = vectNodes.get(i);
+   					System.out.println("Noeud origine2222 "+ni.get_numero());
    					j = sc.nextInt();
+   					System.out.println("Noeud dest "+j);
    					Node nj = vectNodes.get(j);
-   					Double cost = Double.parseDouble(sc.next()); 
+   					System.out.println("Noeud dest2222 "+nj.get_numero());
+   					String ss = sc.next();
+   					Double cost = Double.parseDouble(ss); 
+   					System.out.println("Cout "+cost);
    					// Add the new arc
    					nbA++;
    					Arc arc = new Arc(ni, nj);
@@ -76,59 +84,38 @@ public class Graph {
    					arc.setCost(cost);
    					ni.add_out_arc(arc);
    					ni.add_in_arc(arc);
-   					vectArcs.add(arc);   					
+   					
+   					// add the risk to the arc
+   					for (Integer k = 0; k < nbCom; k++){
[…]

### 2011-10-17 code/variatorHazmatJava/src/graph/Arc.java
The code is modified to take into account the new format of data

@@ -10,20 +10,22 @@ public class Arc {
 	
 	private Integer num; 				// numero of the arc
 	private double cost; 			// cost of the arc
-	private double rCost; 			// reduced cost of the arc
-	private double totRisk; 		// The total risk on the arc (for all the regions)
+	//private double rCost; 			// reduced cost of the arc
+	//private double totRisk; 		// The total risk on the arc (for all the regions)
 	private Node origNode; 			// origin node of the arc
 	private Node destNode; 			// destination node of the arc
-	private Vector comRegRisk; 				// com --> Reg--> risk
+	//private Vector comRegRisk; 				// com --> Reg--> risk
+	
 	// TODO: why do we need a list of successor and predecessor nodes for each arc here? (Dimo)
 	private ArrayList<Node> succNodes;		// Successor Nodes of the arc
 	private ArrayList<Node> predNodes;		// predecessor Nodes of the arc
-	private ArrayList<ArrayList<Integer>>  mapRisk; // commodity - region - risk 
+	//private ArrayList<ArrayList<Integer>>  mapRisk; // commodity - region - risk 
+	private ArrayList<ArrayList<Double>>  comRisk; // commodity - risk 
 	
 	public Arc(){
 		this.succNodes = new ArrayList<Node>();
 		this.predNodes = new ArrayList<Node>();
-		this.mapRisk = new ArrayList<ArrayList<Integer>>();
+		this.comRisk = new ArrayList<ArrayList<Double>>();
 	}
 
 	public Arc(Node v1, Node v2){
@@ -31,7 +33,7 @@ public class Arc {
 	  this.destNode = v2;
 	  this.succNodes = new ArrayList<Node>();
 	  this.predNodes = new ArrayList<Node>();
-	  this.mapRisk = new ArrayList<ArrayList<Integer>>();
+	  this.comRisk = new ArrayList<ArrayList<Double>>();
 	}
 	
 	public Node returnOrigNode(){
@@ -54,13 +56,13 @@ public class Arc {
 	  return num;
 	}
 
-	public double returnReducedCost() {
-	  return rCost;
-	}
+	//public double returnReducedCost() {
+	// return rCost;
+	//}
 
-	public void setReducedCost(double cr) {
-		this.rCost = cr;
-	}
[…]

### 2011-10-17 code/variatorHazmatJava/src/mainTest.java

@@ -10,17 +10,19 @@ public class mainTest {
 		   System.out.println("1. Lecture du fichier de donnee");		   
 		   Graph G = new Graph();
 		   G.readInstance(args[0]);
+		   System.out.println("Fin de lecture de l'instance");	
 		   Commodity c = G.returnListCommodities().get(0);
 		   int indexO = c.getSource();
 		   Node o = G.returnNode(indexO);
 		   int indexD = c.getDest();
 		   Node d = G.returnNode(indexD);
+		   System.out.println("Commodite 0: Origine = "+o.get_numero()+", Dest = "+d.get_numero());	
 		   ArrayList<Node> sPath= G.shortestPath(o,d);
-		   
-		   // Test		   
-		   Arc arc = G.getArc(3, 6);		      
-		   double risk = arc.getRisk(1, 1);
-		   System.out.println("3  6  1  1 = " + risk);	
+		   // Test	
+		   System.out.println("Je suis laaaaaaa11111111111");	
+		   Arc arc = G.getArc(39, 38);		      
+		   double risk = arc.getRisk(0);
+		   System.out.println("r(39, 38) = " + risk);	
 		   
 		   G.graphViz(args[0]);
 	   }

### 2011-10-17 trunk/code/variatorHazmatJava/src/graph/Graph.java
remove some prints

@@ -63,19 +63,19 @@ public class Graph {
    				int j = 0;
    				int nbA = 0;   // The number of arcs 
    				while (sc.hasNextInt()){
-   					System.out.println("New arc \n");
+   					//System.out.println("New arc \n");
    					//buffer = sc.next();
    					i = sc.nextInt();
-   					System.out.println("Noeud origine "+i);
+   					//System.out.println("Noeud origine "+i);
    					Node ni = vectNodes.get(i);
-   					System.out.println("Noeud origine2222 "+ni.get_numero());
+   					//System.out.println("Noeud origine2222 "+ni.get_numero());
    					j = sc.nextInt();
-   					System.out.println("Noeud dest "+j);
+   					//System.out.println("Noeud dest "+j);
    					Node nj = vectNodes.get(j);
-   					System.out.println("Noeud dest2222 "+nj.get_numero());
+   					//System.out.println("Noeud dest2222 "+nj.get_numero());
    					String ss = sc.next();
    					Double cost = Double.parseDouble(ss); 
-   					System.out.println("Cout "+cost);
+   					//System.out.println("Cout "+cost);
    					// Add the new arc
    					nbA++;
    					Arc arc = new Arc(ni, nj);
@@ -90,7 +90,7 @@ public class Graph {
    						Double risk = Double.parseDouble(sc.next()); 
    						Double t = k.doubleValue();
    						arc.setRisk(t, risk);	
-   						System.out.println("risk com"+k+" = "+ risk);
+   						//System.out.println("risk com"+k+" = "+ risk);
    					}
    					
    					vectArcs.add(arc);   
@@ -98,7 +98,7 @@ public class Graph {
    					
    				}
    				nbArcs = nbA;
-   				System.out.println("number of arcs = "+nbA);
+   				//System.out.println("number of arcs = "+nbA);
    				String s;
    				// Read: EndArcs source	 dest	demand 
    				for (int k = 0; k < 4; k++){
@@ -110,12 +110,12 @@ public class Graph {
    					int source =  sc.nextInt();   					   					
    					int dest =  sc.nextInt();   					   					
    					int demand =  sc.nextInt(); 
-   					System.out.println("Commodity "+k+": Source = "+source+", dest = "+ dest+" and demand = "+demand);
[…]

### 2011-10-17 trunk/code/variatorHazmatJava/src/mainTest.java

@@ -19,7 +19,7 @@ public class mainTest {
 		   System.out.println("Commodite 0: Origine = "+o.get_numero()+", Dest = "+d.get_numero());	
 		   ArrayList<Node> sPath= G.shortestPath(o,d);
 		   // Test	
-		   System.out.println("Je suis laaaaaaa11111111111");	
+		   //System.out.println("Je suis laaaaaaa11111111111");	
 		   Arc arc = G.getArc(39, 38);		      
 		   double risk = arc.getRisk(0);
 		   System.out.println("r(39, 38) = " + risk);	

### 2011-10-18 code/variatorHazmatJava/src/graph/Commodity.java
add the number of trucks to the file data and the code

@@ -7,7 +7,7 @@ public class Commodity {
 	private int dest;
 	private int demand;
 	//private int tCap;
-	//private int nbTrucks;
+	private int nbTrucks;
 	
 	//public Commodity(int n, int s, int des, int dem, int t, int nb){
 	//this.num = n;
@@ -18,16 +18,17 @@ public class Commodity {
 	//this.nbTrucks = nb;
 	//}
 	
-	public Commodity(int n, int s, int des, int dem){
+	public Commodity(int n, int s, int des, int dem, int nb){
 		this.num = n;
 		this.source = s;
 		this.dest = des;
 		this.demand = dem;
+		this.nbTrucks = nb;
 	}
 	
-	//public int getNbTrucks() {
-	//return this.nbTrucks;
-	//}
+	public int getNbTrucks() {
+	return this.nbTrucks;
+	}
 	
 	public int getSource() {
 		return this.source;
@@ -50,7 +51,7 @@ public class Commodity {
 	//}
 	
 	public Commodity getCopy() {
-		return new Commodity(this.num, this.source, this.dest, this.demand);
+		return new Commodity(this.num, this.source, this.dest, this.demand, this.nbTrucks);
 	}
 }
 

### 2011-10-18 code/variatorHazmatJava/src/graph/Graph.java

@@ -101,7 +101,7 @@ public class Graph {
    				//System.out.println("number of arcs = "+nbA);
    				String s;
    				// Read: EndArcs source	 dest	demand 
-   				for (int k = 0; k < 4; k++){
+   				for (int k = 0; k < 5; k++){
    					s = sc.next();   					
    				}
    				
@@ -110,8 +110,9 @@ public class Graph {
    					int source =  sc.nextInt();   					   					
    					int dest =  sc.nextInt();   					   					
    					int demand =  sc.nextInt(); 
-   					//System.out.println("Commodity "+k+": Source = "+source+", dest = "+ dest+" and demand = "+demand);
-   					Commodity comm = new Commodity(k, source, dest, demand);
+   					int nbTrucks = sc.nextInt();
+   					//System.out.println("Commodity "+k+": Source = "+source+", dest = "+ dest+" and demand = "+demand+" and NbTrucks = "+nbTrucks);
+   					Commodity comm = new Commodity(k, source, dest, demand, nbTrucks);
    					this.listCom.add(comm); 
    				}   				
 	   			sc.close(); 

### 2011-10-18 code/variatorHazmatJava/src/population/IndividualHazmat.java
rewrote eval() --- implementing the new model of (1) min costs (2) min total risk (3) min max_{a \in arcs} risks associated with a and summed over all commodities, [no testing done yet though]

@@ -176,10 +176,10 @@ public class IndividualHazmat extends IndividualAbstract {
 			this.objectiveSpace[0] = 0;
 			this.objectiveSpace[1] = 0;
 			this.objectiveSpace[2] = Double.NEGATIVE_INFINITY; // 2nd obj. is a maximum of risks
-			// store preliminary sums of the r^{cq}_{ij} y^c_{ij} for each region:
-			ArrayList<Double> risksumsPerRegion = new ArrayList<Double>();
-			for (int r=0; r < PopulationHazmat.mygraph.nbReg; r++) {
-				risksumsPerRegion.add(r, 0.0);
+			// store risks summed over all commodities for each arc:
+			ArrayList<Double> risksumsPerArc = new ArrayList<Double>();
+			for (int arcRisk=0; arcRisk < PopulationHazmat.mygraph.nbArcs; arcRisk++) {
+				risksumsPerArc.add(arcRisk, 0.0);
 			}
 			
 			for (LinkedList<Node> path : completedTruckPaths) {
@@ -191,18 +191,17 @@ public class IndividualHazmat extends IndividualAbstract {
 					Arc currArc = PopulationHazmat.mygraph.getArc(a.get_numero(), b.get_numero());
 					// update 1st objective for first arc
 					this.objectiveSpace[0] += currArc.returnCost();
-					// for update of 2nd and 3rd objective, update the risk per region
-					for (int r=0; r < risksumsPerRegion.size(); r++) {
-						risksumsPerRegion.set(r, risksumsPerRegion.get(r)
-								+ currArc.getRisk(this.associatedCommodities.get(i).getNum(), r));
-					}
+					// for update of 2nd and 3rd objective, update the risk per arc
+					int currArcID = currArc.returnNum();
+					risksumsPerArc.set(currArcID, risksumsPerArc.get(currArcID)
+								+ currArc.getRisk(this.associatedCommodities.get(i).getNum()));
 				}
 				i++;
 			} // now, 1st objective already computed, 2nd, and 3rd still missing
 			
 			// go through risksumsPerRegion and compute remaining objectives as sum and max resp.
-			for (int r=0; r < risksumsPerRegion.size(); r++) {
-				double currRisk = risksumsPerRegion.get(r);	
+			for (int arcID=0; arcID < risksumsPerArc.size(); arcID++) {
+				double currRisk = risksumsPerArc.get(arcID);	
[…]

### 2011-10-18 code/variatorHazmatJava/src/graph/Arc.java
Correct the function returnNum()

@@ -28,9 +28,10 @@ public class Arc {
 		this.comRisk = new ArrayList<ArrayList<Double>>();
 	}
 
-	public Arc(Node v1, Node v2){
+	public Arc(Node v1, Node v2, int n){
 	  this.origNode = v1;
 	  this.destNode = v2;
+	  this.num = n;
 	  this.succNodes = new ArrayList<Node>();
 	  this.predNodes = new ArrayList<Node>();
 	  this.comRisk = new ArrayList<ArrayList<Double>>();
@@ -54,6 +55,10 @@ public class Arc {
 	
 	public int returnNum() {
 	  return num;
+	}
+	
+	public void setNum(int n) {
+		num = n;
 	}
 
 	//public double returnReducedCost() {

### 2011-10-18 code/variatorHazmatJava/src/graph/Graph.java

@@ -77,8 +77,7 @@ public class Graph {
    					Double cost = Double.parseDouble(ss); 
    					//System.out.println("Cout "+cost);
    					// Add the new arc
-   					nbA++;
-   					Arc arc = new Arc(ni, nj);
+   					Arc arc = new Arc(ni, nj, nbA);
    					arc.addSuccNode(nj);
    					arc.addPredNode(ni);
    					arc.setCost(cost);
@@ -94,7 +93,7 @@ public class Graph {
    					}
    					
    					vectArcs.add(arc);   
-   					
+   					nbA++;
    					
    				}
    				nbArcs = nbA;

### 2011-10-18 trunk/code/variatorHazmatJava/src/graph/Graph.java
introduced Hashtable for shortest paths to decrease run time of the algorithm (for one and the same s-t node pair, the shortest path is only computed once now and not every time, it is needed)

@@ -26,11 +26,17 @@ public class Graph {
 		public ArrayList<Node> vectNodes; 
 		public ArrayList<Arc> vectArcs;   
 		public ArrayList<Commodity> listCom; 
+		
+		// List of hashtables for storing already computed shortest paths:
+		// shortestPathStorage[s] contains all shortest paths from source s in a Hashtable
+		// with the destination node d as key and the corresponding shortes s-d path as value
+		private ArrayList<Hashtable<Node, ArrayList<Node>>> shortestPathStorage;
 	  	   
 	   public Graph() {
 		   this.vectNodes = new ArrayList<Node>();
 		   this.vectArcs = new ArrayList<Arc>();
 		   this.listCom = new ArrayList<Commodity>();
+		   this.shortestPathStorage = new ArrayList<Hashtable<Node, ArrayList<Node>>>();
 	   }	
 	   
 	   public void readInstance(String fileName) {
@@ -120,6 +126,12 @@ public class Graph {
 	   		catch (Exception e){
 	   			System.out.println(e.toString());
 	   		}
+	   		
+	   		// don't forget to finish initialization of this.shortestPathStorage:
+	   		for (int i=0; i< this.nbNodes; i++) {
+	   			this.shortestPathStorage.add(new Hashtable<Node, ArrayList<Node>>());
+	   		}
+	   		
        }
 
 		 public int returnNbNodes() {
@@ -174,159 +186,172 @@ public class Graph {
 		 
 		 public ArrayList<Node> shortestPath(Node o, Node d) {
 			 
-			 //System.out.println("shortestPath from "+ o.get_numero() + " to "+d.get_numero());
-			 
-			 // The path: List of nodes
-			 ArrayList<Integer> sPath = new ArrayList<Integer>();
-			 ArrayList<Node> sPathSol = new ArrayList<Node>();
-			 
-			 // For each node (numero), we associate its weight
-			 HashMap<Integer, Double> weight = new HashMap<Integer, Double>();
-			 
-			 // For each node (numero), we associate its predecessor node in the shortest path
-			 HashMap<Integer, Integer> pred = new HashMap<Integer, Integer>();
-			 
-			 // List of definively treated nodes
-			 ArrayList<Node> treatedNodes = new ArrayList<Node>(0);
-			 			 
-			 // List of temporary treated nodes
-			 ArrayList<Node> visitedNodes = new ArrayList<Node>(0);
-						 
-			 // Initialize weight and pred
[…]

### 2011-10-18 trunk/code/variatorHazmatJava/src/population/IndividualHazmat.java

@@ -114,10 +114,10 @@ public class IndividualHazmat extends IndividualAbstract {
 		this.alreadyEvaluated = true;
 		
 		
-		System.out.println("---");
-		System.out.println(this.truckPaths);
-		System.out.println(this.associatedCommodities);
-		System.out.println(this.objectiveSpace);
+		//System.out.println("---");
+		//System.out.println(this.truckPaths);
+		//System.out.println(this.associatedCommodities);
+		//System.out.println(this.objectiveSpace);
 		
 		
 	}
@@ -125,6 +125,9 @@ public class IndividualHazmat extends IndividualAbstract {
 	/** Calculates the objective space values of this individual. The two objectives are number of
 	 * leading ones and number of trailing zeros. Both objectives are converted such that they have to
 	 * be minimized (by subtracting them from the total number of decision variables). */
+	/* (non-Javadoc)
+	 * @see population.IndividualAbstract#eval()
+	 */
 	public void eval() {
 		
 		alreadyEvaluated = false; // for the moment to ensure a new evaluation each time (testing) 
@@ -162,14 +165,14 @@ public class IndividualHazmat extends IndividualAbstract {
 				i++;
 			}
 			
-			String s = "";
-			for (LinkedList<Node> tp: completedTruckPaths) {
-				s += "path: ";
-				for (Node n: tp) {
-					s += n.get_numero() + " ";
-				}
-			}
-			System.out.println(s);
+			//String s = "";
+			//for (LinkedList<Node> tp: completedTruckPaths) {
+			//	s += "path: ";
+			//	for (Node n: tp) {
+			//		s += n.get_numero() + " ";
+			//	}
+			//}
+			//System.out.println(s);
 			
 			/* Second, go through all paths and compute the objective functions */
 			i = 0; // as above, an index to get the corresponding commodity
@@ -211,11 +214,11 @@ public class IndividualHazmat extends IndividualAbstract {
 			this.alreadyEvaluated = true;
 		}
 		
-		System.out.println(this);
-		System.out.println(this.objectiveSpace[0]);
-		System.out.println(this.objectiveSpace[1]);
-		System.out.println(this.objectiveSpace[2]);
-		System.out.println();
+		//System.out.println(this);
+		//System.out.println(this.objectiveSpace[0]);
+		//System.out.println(this.objectiveSpace[1]);
+		//System.out.println(this.objectiveSpace[2]);
+		//System.out.println();
 	}
 
 	/** Returns a (deep) copy of this individual.

### 2011-11-09 code/variatorHazmatJava/src/population/PopulationHazmat.java
implemented new random initialization

@@ -57,6 +57,15 @@ public class PopulationHazmat extends PopulationAbstract {
 	/** The probability during mutation of an individual that the current bit is flipped. */
 	double bitFlipProbability = 0.1;
 	
+	/** Integer value indicating the type of initialization:
+	 *    0: all individuals are created with the trucks taking the shortest (cost) path
+	 *    1: all individuals are created randomly in the sense that a random node is chosen
+	 *       as end node of the representation and the shortest path between the source node
+	 *       and this randomly chosen node resembles the individual's representation. */
+	public static int INITTYPE_COST = 0;
+	public static int INITTYPE_RAND = 1; 
+	int initializationType = INITTYPE_COST;
+	
 	/** File name where graph instance is defined */
 	String graph_definition_file = "graph_ns2_1_New.dat";
 	
@@ -73,6 +82,13 @@ public class PopulationHazmat extends PopulationAbstract {
 		IndividualHazmat newIndividual;
 		for (int i = 0; i < alpha; i++) {
 			newIndividual = new IndividualHazmat();
+			
+			if (initializationType == INITTYPE_RAND) {
+				newIndividual.initRandom();
+			}
+			else {
+				newIndividual.initCost();
+			}
 			globalPopulation.add(newIndividual);
 		}
 		Variator.debugPrint("Initial population constructed.");
@@ -195,6 +211,20 @@ public class PopulationHazmat extends PopulationAbstract {
 		else if (paramName.equals("bitflip_probability")){
 			try {
 				bitFlipProbability = Double.parseDouble(paramValue);
+				return true;
+			}
+			catch (Exception ex) {
+				return false;
+			}
+		}
+		else if (paramName.equals("initialization_type")){
+			try {
+				if (paramValue.equals("rand") || paramValue.equals("random")) {
+					initializationType = INITTYPE_RAND;
+				}
+				else {
+					initializationType = INITTYPE_COST;
+				}
 				return true;
 			}
 			catch (Exception ex) {

### 2011-11-09 trunk/code/variatorHazmatJava/src/population/IndividualHazmat.java
implemented new random initialization

@@ -48,11 +48,12 @@ public class IndividualHazmat extends IndividualAbstract {
 	Vector<Commodity> associatedCommodities; // list of associated commodities of each truck
 	boolean alreadyEvaluated;
 	
-	/** Standard class constructor, initializes the decision space representation randomly
-	 * based on the graph instance in PopulationHazmat.mygraph.
+	/** Standard class constructor
 	 * 
-	 * In a first implementation, a fixed number of trucks per commodity is send through
-	 * the network.
+	 * Note that no initialization of the truck paths is taking place here!
+	 * 
+	 * Before using the individual further, the initialization should be performed via the
+	 * methods 'initCost' and 'initRandom'.
 	 * 
 	 */
 	public IndividualHazmat(){
@@ -60,19 +61,6 @@ public class IndividualHazmat extends IndividualAbstract {
 		this.associatedCommodities = new Vector<Commodity>();
 		this.alreadyEvaluated = false;
 		
-		for (Commodity commodity: PopulationHazmat.mygraph.listCom) {
-			/* create "empty" path, consisting of only the source node for this commodity */
-			LinkedList<Node> emptyPath = new LinkedList<Node>();
-			emptyPath.add(PopulationHazmat.mygraph.returnNode(commodity.getSource()));
-
-			for (int i=0; i < commodity.getNbTrucks(); i++) {
-				this.truckPaths.add(emptyPath);
-				this.associatedCommodities.add(commodity);
-			}
-		}
-		
-		this.objectiveSpace = new double[3];
-		this.eval();
 	}
 	
 	/** Class constructor which initializes the decision and objective space representations with the given values.
@@ -120,6 +108,62 @@ public class IndividualHazmat extends IndividualAbstract {
 		//System.out.println(this.objectiveSpace);
 		
 		
+	}
+
+	/** Initializes the individual with all truck paths containing only the source node of
+	 * the corresponding commodity. */
+	public void initCost() {
+		for (Commodity commodity: PopulationHazmat.mygraph.listCom) {
+			/* create "empty" path, consisting of only the source node for this commodity */
+			LinkedList<Node> emptyPath = new LinkedList<Node>();
+			emptyPath.add(PopulationHazmat.mygraph.returnNode(commodity.getSource()));
+
+			for (int i=0; i < commodity.getNbTrucks(); i++) {
+				this.truckPaths.add(emptyPath);
+				this.associatedCommodities.add(commodity);
+			}
+		}
+		
[…]

### 2011-11-09 DAW_hazmatHypE/trunk/experiments/randVsCost/interactivemonitor/intmon.py
added results of first official runs comparing the random and the shortest path initialization (until generation 100 on graph_ns2_1_New.dat instance)

@@ -1,0 +1,318 @@
+"""Interactive PISA monitor.
+
+    Calls various selector and variator modules to perform PISA algorithm runs.
+    Is going to provide interactive optimization features such as plotting.
+    
+    Usage: python intmon.py PV CV PS CS PM OM poll
+    
+    where       PV: parameter file of variator
+                CV: communication filename base of variator
+                PS: paramameter file of selector
+                CS: communication filename base of selector
+                PM: parameter file of monitor 
+                OM: filename base of monitor output files
+                poll: polling interval
+
+    """
+
+
+# ========================================================================
+#  PISA  (www.tik.ee.ethz.ch/pisa/)
+# 
+#  ========================================================================
+#  Computer Engineering Lab (TIK)
+#  ETH Zurich
+# 
+#  ========================================================================
+#  Interactive Monitor 
+#  
+#  Main file
+#  
+#  python implementation
+#  
+#  file: intmon.py
+#  author: Dimo Brockhoff brockho@lix.polytechnique.fr
+#
+#  last change: April 2011
+#  
+#  ========================================================================
+
+
+
+import sys
+import getopt
+from intmon_internal import *
+import random
+import time
+from collections import deque
+import interaction
+
+
+currentGeneration = 0
+currentRun = 0
+
+
+
+
+def wait(timeInSeconds):
+    assert(float(timeInSeconds) >= 0.001)
+    time.sleep(float(timeInSeconds))
+    
+# end of wait(timeInSeconds)    
+    
+    
+def writeNewSeed(filename, updateType, reset):
+    """ replaces old seed in file 'filename' by a new seed iff a line
+        beginning with 'seed' exists within the file 'filename'.
+        The seed is chosen either randomly (updateType == 'random'), linearly
+        with the number of the current run (updateType == 'linearlyWithRuns'),
+        or not changed at all (updateType == 'const').
+        If reset==1 and updateType == 'linearlyWithRuns', the seed is reset
+        by writing the seed '10001' such that the seed is always '10000+#runs'.
+    """
+    
+    lines = deque() # to store file contents, use a queue
+    
+    # first read file fully
+    try:
+        file = open(filename, 'r')
+        for line in file:
+            lines.append(line)
+        file.close()
+    except IOError:
+        print 'Problem reading file ' + filename
+
+    # now write file back with new seed
+    try:
+        file = open(filename, 'w')
+        for line in lines:
+            if 'seed' in line:
+                newseed = 0
[…]

### 2011-11-09 DAW_hazmatHypE/trunk/experiments/randVsCost/interactivemonitor/onClick.py

@@ -1,0 +1,129 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from comparativePrefStatements import *
+
+class onClick:
+    def __init__(self, ov, xtol=None, ytol=None):
+        self.toggle = 1
+        self.ps = ''
+        self.objectiveVectors = ov
+        if (len(ov[0]) == 2):
+            if xtol is None:
+                xtol = ((max(ov[:,0]) - min(ov[:,0]))/float(len(ov)))/2
+            if ytol is None:
+                ytol = ((max(ov[:,1]) - min(ov[:,1]))/float(len(ov)))/2
+        else:
+            xtol = 0.0
+            ytol = (np.max(ov) - np.min(ov)) / float(len(ov)) / 4
+            
+        print ov
+            
+        self.xtol = xtol
+        self.ytol = ytol
+        self.lastChosenIndex = -2
+    
+    def __call__(self, event):
+        clickX = event.xdata
+        clickY = event.ydata
+        # needed for detecting clicks inside top left box:
+        ymin, ymax = plt.ylim()
+        
+        # if left mouse click: remove plot window and continue with optimization
+        # where the closest solution to the mouse click is extracted:
+        if (event.button == 1) and event.inaxes:
+            if (len(self.objectiveVectors[0]) == 2):
+                for i in range(len(self.objectiveVectors)):
+                    if clickX-self.xtol < self.objectiveVectors[i,0] < clickX+self.xtol and  clickY-self.ytol < self.objectiveVectors[i,1] < clickY+self.ytol :
+                        self.lastChosenIndex = i
+                        plt.close()
+            else:
+                # sort objective vectors for easy selecting them on the right
+                I = np.argsort(self.objectiveVectors[:,self.objectiveVectors.shape[1]-1])
+                objVectorsSorted = self.objectiveVectors[I,:]
+                # add (kind of) id of solution as last column like in plot:
+                maximum = objVectorsSorted.max(0).max(0)
+                minimum = objVectorsSorted.min(0).min(0)
+                
+                objVectorsSorted = np.column_stack((objVectorsSorted,
+                                    np.linspace(minimum, maximum,
+                                    objVectorsSorted.shape[0]).transpose()))
+                
+                # detect click only when on the right:
+                if (clickX > objVectorsSorted.shape[1]-2 + 0.25):
+                    # find correct objective vector among the sorted ones:
+                    for i in range(len(objVectorsSorted)):
[…]

### 2011-11-09 trunk/DAW_hazmatHypE/trunk/experiments/randVsCost/interactivemonitor/AnnoteFinder.py
added results of first official runs comparing the random and the shortest path initialization (until generation 100 on graph_ns2_1_New.dat instance)

@@ -1,0 +1,79 @@
+""" This code is copied from the matplotlib Cookbook:
+    http://www.scipy.org/Cookbook/Matplotlib/Interactive_Plotting
+    and slightly adapted.
+"""
+
+import math
+import pylab
+import matplotlib
+
+class AnnoteFinder:
+    """
+    callback for matplotlib to display an annotation when points are clicked on.  The
+    point which is closest to the click and within xtol and ytol is identified.
+
+    Register this function like this:
+    
+    scatter(xdata, ydata)
+    af = AnnoteFinder(xdata, ydata, annotes)
+    connect('button_press_event', af)
+    """
+ 
+    def __init__(self, xdata, ydata, annotes, axis=None, xtol=None, ytol=None):
+        self.data = zip(xdata, ydata, annotes)
+        if xtol is None:
+            xtol = ((max(xdata) - min(xdata))/float(len(xdata)))/2
+        if ytol is None:
+            ytol = ((max(ydata) - min(ydata))/float(len(ydata)))/2
+        self.xtol = xtol
+        self.ytol = ytol
+        if axis is None:
+            self.axis = pylab.gca()
+        else:
+            self.axis= axis
+        self.drawnAnnotations = {}
+        self.links = []
+
+    def distance(self, x1, x2, y1, y2):
+        """
+        return the distance between two points
+        """
+        return math.hypot(x1 - x2, y1 - y2)
+  
+    def __call__(self, event):
+        # button==3 corresponds to right click:
+        if (event.button == 3) and event.inaxes:
+            clickX = event.xdata
+            clickY = event.ydata
+            if self.axis is None or self.axis==event.inaxes:
+                annotes = []
+                for x,y,a in self.data:
+                    if  clickX-self.xtol < x < clickX+self.xtol and  clickY-self.ytol < y < clickY+self.ytol :
+                        annotes.append((self.distance(x,clickX,y,clickY),x,y, a) )
+                if annotes:
+                    annotes.sort()
+                    distance, x, y, annote = annotes[0]
+                    self.drawAnnote(event.inaxes, x, y, annote)
+                    for l in self.links:
+                        l.drawSpecificAnnote(annote)
+                        
+
+    def drawAnnote(self, axis, x, y, annote):
+        """
+        Draw the annotation on the plot
+        """
+        if (x,y) in self.drawnAnnotations:
[…]

### 2011-11-10 DAW_hazmatHypE/trunk/experiments/randVsCost/plot3d_after1000gen.m
added new results comparing random and shortest path initialization (1000 generations, 10 runs with HypE), added first plot script as well

@@ -1,0 +1,39 @@
+% plot resulting populations after 1000 generations
+
+C = dlmread('results/graph_ns2_1_New_costInit.1000');
+R = dlmread('results/graph_ns2_1_New_randInit.1000');
+
+subplot(2,2,1)
+plot3(C(:,1),C(:,2),C(:,3), 'x');
+hold all; grid on;
+plot3(R(:,1),R(:,2),R(:,3), 'o');
+xlabel('cost')
+ylabel('total risk')
+zlabel('equity risk')
+
+
+subplot(2,2,2)
+plot(C(:,1),C(:,2), 'x');
+hold all; grid on;
+plot(R(:,1),R(:,2), 'o');
+legend('shortest path initialization', 'random initialization');
+xlabel('cost')
+ylabel('total risk')
+
+
+subplot(2,2,3)
+plot(C(:,1),C(:,3), 'x');
+hold all; grid on;
+plot(R(:,1),R(:,3), 'o');
+legend('shortest path initialization', 'random initialization');
+xlabel('cost')
+ylabel('equity risk')
+
+
+subplot(2,2,4)
+plot(C(:,2),C(:,3), 'x');
+hold all; grid on;
+plot(R(:,2),R(:,3), 'o');
+legend('shortest path initialization', 'random initialization');
+xlabel('total risk')
+ylabel('equity risk')

### 2011-11-10 trunk/DAW_hazmatHypE/trunk/experiments/randVsCost/mysubplot.m
updated plot of population after 1000 generations of HypE with both initialization schemes

@@ -1,0 +1,56 @@
+function [] = mysubplot(C, R, C_label, R_label, titlename, filename)
+figure()
+
+subplot(2,2,1)
+plot3(C(:,1),C(:,2),C(:,3), 'x');
+hold all; grid on;
+plot3(R(:,1),R(:,2),R(:,3), 'o');
+xlabel('cost')
+ylabel('total risk')
+zlabel('equity risk')
+
+
+subplot(2,2,2)
+plot(C(:,1),C(:,2), 'x');
+hold all; grid on;
+plot(R(:,1),R(:,2), 'o');
+ylim([4,6]*1e6);
+h_legend = legend(C_label, R_label);
+set(h_legend,'FontSize',7);
+xlabel('cost')
+ylabel('total risk')
+
+
+subplot(2,2,3)
+plot(C(:,1),C(:,3), 'x');
+hold all; grid on;
+plot(R(:,1),R(:,3), 'o');
+ylim([2,5]*1e5);
+h_legend = legend(C_label, R_label);
+set(h_legend,'FontSize',7);
+xlabel('cost')
+ylabel('equity risk')
+
+
+subplot(2,2,4)
+plot(C(:,2),C(:,3), 'x');
+hold all; grid on;
+plot(R(:,2),R(:,3), 'o');
+ylim([2,5]*1e5);
+h_legend = legend(C_label, R_label);
+set(h_legend,'FontSize',7);
+xlabel('total risk')
+ylabel('equity risk')
+
+% save plot as eps:
+exportfig(gcf, filename);
+
+% add title above all for plot on screen:
+set(gcf,'NextPlot','add');
+axes;
+h = title(titlename);
+set(gca,'Visible','off');
+set(h,'Visible','on'); 
+
+
+end

### 2011-11-10 DAW_hazmatHypE/trunk/experiments/randVsCost/plot3d_after1000gen.m
updated plot of population after 1000 generations of HypE with both initialization schemes

@@ -1,39 +1,25 @@
 % plot resulting populations after 1000 generations
 
 C = dlmread('results/graph_ns2_1_New_costInit.1000');
+C_nondom = C(find(paretoFront(C(:,1:3))),:);
+%plot3(C(:,1),C(:,2),C(:,3), 'x');
+%hold all; grid on;
+%plot3(C_nondom(:,1),C_nondom(:,2),C_nondom(:,3), 'o');
+
+
 R = dlmread('results/graph_ns2_1_New_randInit.1000');
-
-subplot(2,2,1)
-plot3(C(:,1),C(:,2),C(:,3), 'x');
-hold all; grid on;
-plot3(R(:,1),R(:,2),R(:,3), 'o');
-xlabel('cost')
-ylabel('total risk')
-zlabel('equity risk')
+R_nondom = R(find(paretoFront(R(:,1:3))),:);
+%figure()
+%plot3(R(:,1),R(:,2),R(:,3), 'x');
+%hold all; grid on;
+%plot3(R_nondom(:,1),R_nondom(:,2),R_nondom(:,3), 'o');
 
 
-subplot(2,2,2)
-plot(C(:,1),C(:,2), 'x');
-hold all; grid on;
-plot(R(:,1),R(:,2), 'o');
-legend('shortest path initialization', 'random initialization');
-xlabel('cost')
-ylabel('total risk')
 
+%mysubplot(C,C_nondom, 'C', 'C_nondom', 'C only', 'filename');
+%mysubplot(R,R_nondom, 'R', 'R_nondom', 'R only', 'filename');
 
-subplot(2,2,3)
-plot(C(:,1),C(:,3), 'x');
-hold all; grid on;
-plot(R(:,1),R(:,3), 'o');
-legend('shortest path initialization', 'random initialization');
-xlabel('cost')
-ylabel('equity risk')
+mysubplot(C,R, 'shortest path initialization', 'random initialization', 'all solutions', 'allsolutions.eps')
+mysubplot(C_nondom,R_nondom, 'shortest path initialization', 'random initialization', 'nondominated', 'onlynondominated.eps')
 
-
-subplot(2,2,4)
-plot(C(:,2),C(:,3), 'x');
-hold all; grid on;
-plot(R(:,2),R(:,3), 'o');
-legend('shortest path initialization', 'random initialization');
-xlabel('total risk')
-ylabel('equity risk')
+hold off;

### 2011-11-18 trunk/DAW_hazmatHypE/trunk/experiments/randVsCost/plotCompareRandAndCost.m
added results of last runs, added also MATLAB scripts to plot them

@@ -1,0 +1,36 @@
+% plot resulting populations of rand and cost initialization
+% after certain number of generations
+
+%%%%%%%%%%%%%%%%%%%%%%%
+generation = 2000;
+numOfCommodities = 2;
+instance = 2;
+%%%%%%%%%%%%%%%%%%%%%%%
+
+filename = ['results/graph_ns' int2str(numOfCommodities) '_' int2str(instance) '_New_costInit.' int2str(generation)];
+C = dlmread(filename);
+C_nondom = C(find(paretofront(C(:,1:3))),:);
+figure(77);
+%plot3(C(:,1),C(:,2),C(:,3), 'x');
+plot3(C_nondom(:,1),C_nondom(:,2),C_nondom(:,3), 'o');
+hold all; grid on;
+
+filename = ['results/graph_ns' int2str(numOfCommodities) '_' int2str(instance) '_New_randInit.' int2str(generation)];
+R = dlmread(filename);
+R_nondom = R(find(paretofront(R(:,1:3))),:);
+%figure()
+%plot3(R(:,1),R(:,2),R(:,3), 'x');
+%hold all; grid on;
+plot3(R_nondom(:,1),R_nondom(:,2),R_nondom(:,3), 'o');
+hold off;
+
+
+%mysubplot(C,C_nondom, 'C', 'C_nondom', 'C only', 'filename');
+%mysubplot(R,R_nondom, 'R', 'R_nondom', 'R only', 'filename');
+
+title = ['all solutions after ' int2str(generation) ' generations'];
+mysubplot(C,R, 'shortest path initialization', 'random initialization', title, 'allsolutions.eps');
+title = ['nondominated solutions after ' int2str(generation) ' generations'];
+mysubplot(C_nondom,R_nondom, 'shortest path initialization', 'random initialization', title, 'onlynondominated.eps');
+
+hold off;

### 2011-11-18 DAW_hazmatHypE/trunk/experiments/randVsCost/mysubplot3.m
added results of last runs, added also MATLAB scripts to plot them

@@ -1,43 +1,47 @@
-function [] = mysubplot(C, R, C_label, R_label, titlename, filename)
+function [] = mysubplot3(A, B, C, A_label, B_label, C_label, titlename, filename)
 figure()
 
 subplot(2,2,1)
-plot3(C(:,1),C(:,2),C(:,3), 'x');
+plot3(A(:,1),A(:,2),A(:,3), 'x');
 hold all; grid on;
-plot3(R(:,1),R(:,2),R(:,3), 'o');
+plot3(B(:,1),B(:,2),B(:,3), 'o');
+plot3(C(:,1),C(:,2),C(:,3), 's');
 xlabel('cost')
 ylabel('total risk')
 zlabel('equity risk')
 
 
 subplot(2,2,2)
-plot(C(:,1),C(:,2), 'x');
+plot(A(:,1),A(:,2), 'x');
 hold all; grid on;
-plot(R(:,1),R(:,2), 'o');
+plot(B(:,1),B(:,2), 'o');
+plot(C(:,1),C(:,2), 's');
 %ylim([4,6]*1e6);
-h_legend = legend(C_label, R_label);
+h_legend = legend(A_label, B_label, C_label);
 set(h_legend,'FontSize',7);
 xlabel('cost')
 ylabel('total risk')
 
 
 subplot(2,2,3)
-plot(C(:,1),C(:,3), 'x');
+plot(A(:,1),A(:,3), 'x');
 hold all; grid on;
-plot(R(:,1),R(:,3), 'o');
+plot(B(:,1),B(:,3), 'o');
+plot(C(:,1),C(:,3), 's');
 %ylim([2,5]*1e5);
-h_legend = legend(C_label, R_label);
+h_legend = legend(A_label, B_label, C_label);
 set(h_legend,'FontSize',7);
 xlabel('cost')
 ylabel('equity risk')
 
 
 subplot(2,2,4)
-plot(C(:,2),C(:,3), 'x');
+plot(A(:,2),A(:,3), 'x');
 hold all; grid on;
-plot(R(:,2),R(:,3), 'o');
+plot(B(:,2),B(:,3), 'o');
+plot(C(:,2),C(:,3), 's');
 %ylim([2,5]*1e5);
-h_legend = legend(C_label, R_label);
+h_legend = legend(A_label, B_label, C_label);
 set(h_legend,'FontSize',7);
 xlabel('total risk')
 ylabel('equity risk')

### 2011-11-28 trunk/DAW_hazmatHypE/trunk/experiments/randVsCost/hypervolumes/hypervolumeAlgoComparisons.m
added box plots comparing HypE with popsize 50 and 100 over time

@@ -1,0 +1,30 @@
+addpath('../');
+
+hypervolumes = zeros(10, 5);
+
+% HypE popsize 25
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize25_randInit.4000';
+hypervolumes(:,1) = -1 * dlmread(filename);
+% HypE popsize 50
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize50_randInit.2000';
+hypervolumes(:,2) = -1 * dlmread(filename);
+% HypE popsize 100
+filename = 'HYP_ns4_1_OriginalCosts_randInit.1000';
+hypervolumes(:,3) = -1 * dlmread(filename);
+% HypE popsize 200
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize200_randInit.500';
+hypervolumes(:,4) = -1 * dlmread(filename);
+% NSGA-II
+filename = 'HYP_ns4_1_OriginalCosts_nsga2_randInit.1000';
+hypervolumes(:,5) = -1 * dlmread(filename);
+
+
+
+
+boxplot(hypervolumes, 'Labels', {'HypE_25'; 'HypE_50'; 'HypE_100'; 'HypE_200'; 'NSGA-II'});
+
+
+
+% save plot as eps:
+exportfig(gcf, 'algoComparison.eps');
+

### 2011-11-28 DAW_hazmatHypE/trunk/experiments/randVsCost/hypervolumes/hypervolumeOverTime.m
added box plots comparing HypE with popsize 50 and 100 over time

@@ -1,0 +1,24 @@
+
+generations = [0 250 500 750 1000 1250 1500 1750 2000];
+
+hypervolumes = zeros(10, size(generations, 2));
+
+i = 1;
+for g = generations
+    filename = ['HYP_ns4_1_OriginalCosts_randInit.' int2str(g)];
+    hypervolumes(:,i) = -1 * dlmread(filename);
+    i = i + 1;
+end
+
+
+%boxplot(hypervolumes, 'Labels', {'0'; '250'; '500'; '750'; '1000'; '1250'; '1500'; '1750'; '2000'});
+
+figure(2)
+boxplot(hypervolumes(:,2:end), 'Labels', {'250'; '500'; '750'; '1000'; '1250'; '1500'; '1750'; '2000'});
+
+ylim([9.72e26 9.745e26]);
+xlabel('generations');
+ylabel('hypervolume');
+
+% save plot as eps:
+exportfig(gcf, 'hypervolumeOverTime_100.eps');

### 2011-11-28 DAW_hazmatHypE/trunk/experiments/randVsCost/hypervolumes/hypervolumeOverTime_popsize50.m

@@ -1,0 +1,25 @@
+
+generations = [0 250 500 750 1000 1250 1500 1750 2000];
+
+hypervolumes = zeros(10, size(generations, 2));
+
+i = 1;
+for g = generations
+    filename = ['HYP_ns4_1_OriginalCosts_HypE_popsize50_randInit.' int2str(g)];
+    hypervolumes(:,i) = -1 * dlmread(filename);
+    i = i + 1;
+end
+
+
+%boxplot(hypervolumes, 'Labels', {'0'; '250'; '500'; '750'; '1000'; '1250'; '1500'; '1750'; '2000'});
+
+figure(3)
+boxplot(hypervolumes(:,2:end), 'Labels', {'250'; '500'; '750'; '1000'; '1250'; '1500'; '1750'; '2000'});
+
+ylim([9.72e26 9.745e26]);
+xlabel('generations');
+ylabel('hypervolume');
+
+% save plot as eps:
+exportfig(gcf, 'hypervolumeOverTime_50.eps');
+

### 2011-11-28 DAW_hazmatHypE/trunk/experiments/randVsCost/mysubplot3.m

@@ -9,7 +9,7 @@ plot3(C(:,1),C(:,2),C(:,3), 's');
 xlabel('cost')
 ylabel('total risk')
 zlabel('equity risk')
-
+view(45,30);
 
 subplot(2,2,2)
 plot(A(:,1),A(:,2), 'x');

### 2011-11-29 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added updated all.bib with all old publications and use it in the text, added parts of PISA description (not done yet)

%% This is file `elsarticle-template-1a-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
%%
\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% our packages and definitions: %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{url} 
\usepackage{color}
% \usepackage{dsfont}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{multirow}

\newcommand{\TODO}[1]{{\color{red}TODO: #1}}
\newcommand{\TODON}[1]{{\color{blue}TODO: #1}}
\newcommand{\COMMENTD}[1]{{\color{green}Dimo: #1}}
\newcommand{\COMMENTN}[1]{{\color{green}Nora: #1}}

\journal{DAM}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author{Dimo Brockhoff}
\address{INRIA Lille - Nord Europe, DOLPHIN team, 59650 Villeneuve d'Ascq, France\\
{\upshape \url{dimo.brockhoff@inria.fr}}}

\author{Nora Touati-Moungla}
\address{Laboratoire d'Informatique, \'{E}cole Polytechnique, 91128 Palaiseau Cedex, France\\
{\upshape \url{touati@lix.polytechnique.fr}}}

\begin{abstract}
tbw
\end{abstract}

\begin{keyword}
Evolutionary multiobjective optimization, Transportation of hazardous materials, Risk equity, Shortest Paths. 
\end{keyword}

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \cite{cd2008a}. The main target of this problem is to select routes from a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized---without producing excessive economic costs. 
%
When solving such a problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions are not affected. In this case, one may wish to distribute the risk in an equitable way over the population and the environment. Several studies consider this additional minimization of the equity risk, but most of them consist of a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{aeb2000a, cd2008a}).
% 
% 
% and can be classified into the \textit{resolution-equity-based methods} and the \textit{model-equity-based methods}. In resolution-equity-based methods, the equity constraints are taken into account only indirectly via a dissimilarity index between paths which is to be maximized \cite{JOH92, LOM93, KUB97, aeb2000a}. Model-equity-based methods, on the other hand, formalize the risk equity as constraints directly within the model. In \cite{GOP90a, GOP90b}, for example, the authors propose an equity shortest path model that minimizes the total risk of travel while the difference between the risks imposed on any two arbitrary zones does not exceed a given threshold.
%
A more realistic \emph{multi}-commodity flow model was proposed in \citep{cd2008a} where each commodity is considered as one hazmat type. The objective function is formulated as the sum of the economical cost and the cost related to the consequences of an incident for each material. To deal with risk equity, the costs are defined as functions of the flow traversing the arcs which imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc.

The majority of all hazmat routing studies deal with a single-objective scenario although the problem itself is multiobjective in nature and it is important to study the trade-offs among the objectives. Evolutionary Multiobjective Optimization (EMO) algorithms are able to compute a set of solutions showing these trade-offs within a single algorithm run which is the reason why we propose to use them for the problem of hazmat routing in this study (Sec.~\ref{S_MEO}). Before, we formalize the routing of hazmat problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).

% see for example \cite{cd2008a, cgr2007, gkbk1990, za2004}, but all problem formulations include the equity risk as constraints and not as an additional objective. \citet{gian1998} on the other hand considers the equity risk as separate objective but uses a weighted sum to aggregate his four objectives into a single-objective problem. The main reason why, to the best of our knowledge, no study considered a fully multiobjective model with the equity risk as objective might be the resulting non-linearity of the problem which make classical optimization algorithms such as branch-and-cut \TODON{equity is an objective in \cite{{gian1998}} ... We solve the broblem with a branch and price algorithm because the model is a multicommodity flow model, and this model is the more adapted one for this problem because it permit to manage simultaneously many hazmat types. The min-max formulation of the objective is easily linearized. I give you the major axes of the work: (1) Solve the presented problem (optimize cost, risk and equity) (2) what are models and resolution methods proposed in the litterature (3) what is the model retained and why (in our case, the multicommodity model) (4) how to handle equity with this model ? (5) how to solve it ? (branch and price is adapted for these models, but can not hadle many objectives $->$ evolutionnary algorithms) } infeasible and led us to our choice of proposing an evolutionary multiobjective algorithm as heuristic here.

%The computation of routes with a fairly distributed risk consists in generating dissimilar origin-destination paths, i.e paths which relatively don't impact the same zones. Solution approaches

%Section~\ref{S_RECRP} defines the problem of transporting hazardous materials with equity risk as a three-objective optimization problem with the minimization of the total routing cost, of the total routing risk {\em and} of the risk equity as the objectives---the latter broadly defined as the risk shared by a set of regions that compose the geographical area under consideration. Based on this formulation, Sec.~\ref{S_MEO} presents an evolutionary multiobjective optimization algorithm before Sec.~\ref{S_FW} concludes the abstract.

%In \cite{cd2008a} was proposed a multi-commodity flow model for this problem, where each commodity is considered as one hazmat type. To deal with risk equity, the costs are defined as functions of the flow traversing the arcs, this imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc. The major drawback of this model is that it can lead to the saturation of the arcs of shortest paths and, therefore, does not guarantee the equity target.

%In \cite{BEL10}, the authors consider a similar model but with additional equity constraints and objectives. The objective is the minimization of the maximum risk equity imposed on the population and the environment and the model is solved using a pranch-and-price algorithm.

%As hazmat routing is multiobjective in nature, we introduce in this work a multiobjective version of this problem. Our goal is the minimization of the total routing cost, of the total routing risk {\em and} of the risk equity, the latter broadly defined as the risk shared by a set of regions that compose the geographical area under consideration.

%This paper is organized as follows. We describe the problem in section \ref{S_RECRP}, we give its mathematical formulation in section \ref{S_AOM}, in section \ref{S_MEO} we show how multiobjective evolutionary methods can be applied to this problem and in section \ref{S_FW} we exopse our further work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Multiobjective Risk-Equity Constrained Routing Problem} \label{S_RECRP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Let the transportation network be represented as a directed graph $G = (N, A)$, with $N$ being the set of nodes and $A$ the set of arcs. Let $C$ be the set of commodities, given as a set of point-to-point demands to transport a certain amount of hazmats. For any commodity $c\in C$, let $s^c$ and $t^c$ be the source node and the destination node respectively, and let $V^c$ be the number of available trucks for the transportation of commodity $c$. Each commodity is associated with a unique type of hazmat. We assume that the risk is computed on each arc of the network and is proportional to the number of trucks traversing such an arc. Although such a proportionality is not always given in pracice, our assumption is a good approximation if the trucks are of the same type and/or the load is very similar as it is the case for standardized castor containers for radioactive material\footnote{Note that changing the proposed model towards different types of trucks with different types of loads and therefore different resulting risks is straightforward and is expected to cause no additional difficulties for the proposed evolutionary algorithm.}. With each arc $(i,j) \in A$, we also associate a cost $c_{ij}$, induced by the travel of a truck on this arc.

\subsection{Multiple objective functions} \label{SS_MOF}
The problem of transporting hazmat is multiobjective in nature: one usually wants to minimize the \textit{total cost} of transportation, the \textit{total risk} of transportation and the {\em distributed risk}, which can be defined as a measure of risk that is shared among different arcs or different paths. The three objective functions are typically conflicting pairwisely. For example if all trucks take the shortest path from the source to the destination of their associated commodity, the cost objective is optimized but at the same time, the distributed risk is definitely suboptimal---spreading the risk over separate routes, on the other hand, will result in higher costs due to longer routes for some of the trucks.

\subsection{The optimization model} \label{SS_AOM}
We introduce the integer variable $y_{ij}^c$ for the number of trucks that use arc $(i,j)\in A$ for transporting commodity $c$. As mentioned above, we assume a fixed number of trucks for each commodity and that all trucks associated to a certain commodity have the same load. The proposed model is defined as follows:
\begin{eqnarray}
   \min  & \sum_{c \in C}\limits \sum_{(i,j) \in A}\limits c_{ij} y_{ij}^c & \label{eq:obj1}\\
   \min  & \sum_{c \in C}\limits \sum_{(i,j) \in A}\limits r_{ij}^{c} y_{ij}^c & \label{eq:obj2}\\
   \min  & \max_{(i,j)\in A}\limits \big\{ \sum_{c \in C}\limits r_{ij}^{c} y_{ij}^c \big\} & \label{eq:obj3}\\
    s.t. & \sum_{j \in \delta^+(i)}\limits y_{ij}^c -  \sum_{j \in \delta^-(i)}\limits y_{ji}^c = q_i^c\quad         & \forall i \in N,  c \in  C \label{eq:flow} \\
         & y_{ij}^c \in \{0, 1, 2, \ldots\}                                                            & \forall (i,j)\in A, c\in C\label{eq:integerY}.
\end{eqnarray}

The first objective in (\ref{eq:obj1}) is a cost function and is to be minimized. The second objective, given by (\ref{eq:obj2}), minimizes the total risk over all arcs and objective (\ref{eq:obj3}) minimizes the maximum risk imposed on all arcs. The constraints in (\ref{eq:flow}) are conservation constraints, where $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$ are the direct successors and predecessors of node $i$, and $q_i^c=V^c$ if  $i=s^c$, $q_i^c=-V^c$ if  $i=t^c$ and $q_i^c=0$ otherwise. Note that while the first two objectives are linear in the variables, the maximum in the third objective makes it highly non-linear---the main reason why we propose to use an evolutionary algorithm to deal with the above hazmat routing problem.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{An Evolutionary Multiobjective Optimization Algorithm for Hazmat Routing} \label{S_MEO}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Evolutionary algorithms (EAs) and Evolutionary Multiobjective Optimization (EMO) algorithms in particular are general-purpose randomized search heuristics and as such well suited for problems where the objective function(s) can be highly non-linear, noisy, or even given only implicitly, e.g., by expensive simulations \citep{deb2001a,cvl2007a}. Since the third objective in the above problem formulation is nonlinear, we propose to use an EMO algorithm for the multiobjective risk-equity constrained routing problem here. Our EMO algorithm follows the standard iterative/generational cycle of an EA of mating selection, variation, objective function evaluation, and environmental selection and is build upon the state-of-the-art selection scheme in HypE \citep{bz2011a} as implemented in the PISA platform \citep{bltz2003a}. The variation operators as well as the representation of the solutions, however, have to be adapted to the problem at hand in the following way in order to fulfill the problem's constraints at all times\footnote{The source code of the corresponding PISA hazmat variator is going to be available on the official PISA web page \url{http://www.tik.ee.ethz.ch/pisa/} and until then can be downloaded on the preliminary web page \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip}}. Moreover, two different initialization schemes are proposed to deal with the specific variable length representation.

\subsection{Representation}
We choose a variable length representation as it has been theoretically shown to be a good choice for multiobjective shortest paths problems \citep{horo2009a}: A solution is thereby represented by a list of paths of variable lengths with one path per truck. We decided to consider a fixed amount of trucks for each commodity and therefore a fixed number of paths through the network although, in principle, the number of trucks could change during the algorithm run if an appropriate mutation operator is designed. The left hand side of Fig.~\ref{fig:paths} shows an exemplary solution with two trucks for commodity 1 and one truck for commodity 2.

In order to have every variable length path represent an uninterrupted path from source to destination at any time (see the constraints in (\ref{eq:flow})), we ensure all paths to always start with the source $s^c$ for the corresponding commodity $c$, ensure during initialization and variation that all neighbored vertices in the path are connected by an arc (see Sec.~\ref{sec:init} and \ref{sec:variation}), and complete each path by the shortest path between the path's actual end node and the commodity's destination node $t^c$. The right hand side of Fig.~\ref{fig:paths} shows the entire source--destination paths corresponding to the representation in the left part of the figure.

\begin{figure}%
\includegraphics[width=\columnwidth]{figs/truckpaths}%
\caption{\label{fig:paths} Illustration of the variable-length representation of a solution (left) with its two paths for commodity 1 and one path for commodity 2 in opposition to the resulting full source--destination paths where the last part is the shortest path between the variable-length path's end node and the destination node for this commodity (right). The source and destination nodes for commodity 1 are 0 and 6 and for the second commodity the source is node 1 and the destination is node 7. The smaller numbers along the arcs of the graph in the middle depict the costs of the instance.}
\end{figure}

\subsection{Initialization}\label{sec:init}
Often, the population of a (multiobjective) evolutionary algorithm is initialized with solutions that are drawn uniformly at random within the search space. Although it is not easy to define a uniformity measure in our case of variable-length representation, we still propose a random initialization scheme for our evolutionary algorithm. In addition, the specific problem formulation allows another initialization where a Pareto-optimal solution is inserted in the initial population. In the following, we present both approaches in more detail and see later on how they compare with respect to the overall performance of the algorithm.

\subsubsection{A Simple Random Initialization}
Based on the idea of representing a solution by the beginning of the truck paths which are completed by the shortest path to the destination nodes, we propose the following random initialization:
\begin{enumerate}
	\item Choose for each truck a node in the network uniformly at random among all nodes which are not the source.
	\item Initialize the truck paths with the shortest path from the associated source node to the randomly selected node.
\end{enumerate}
Note that this initialization scheme is, though random, does not produce fully random paths. The initialization results in truck paths which are constructed by concatenating two shortest paths---the shortest path from the source node to the one randomly selected and the shortest path from the random node to the destination, which is added due to the representation described above.

\subsubsection{A Cost-Optimal Initialization}
The chosen representation allows us another initialization which starts each algorithm run from a population consisting of copies of a single cost-optimal solution. To this end, we initialize the paths $p$ for all trucks as empty, containing only the corresponding source node ($p=(s^c)$). With the above proposed representation, this corresponds to the situation where all trucks choose the shortest $s^c$--$t^c$ path for their assigned commodity---implying the smallest possible overall cost but a possibly high risk along the used route(s). Nevertheless, since no other solution can have a smaller cost, the initial solution is already (weakly) Pareto-optimal and is expected to be a good starting point for the algorithm. In the following experiments, we will see when each of the two initialization schemes is favourable.

\subsection{Variation}\label{sec:variation}
As mutation operator, we suggest to shorten or lengthen the path of one or several trucks. In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as in \citep{horo2009a}:
\begin{itemize}
	\item if $b = 0$ and $l=:\textnormal{length}(p)\geq 2$, set $p' = (s^c, \ldots, v_{l-1})$	
	\item if $b = 1$ and $|V_{\textnormal{\scriptsize rem}}=\{v \in V \,|\, (v_{l}, v) \in A\}| \neq\emptyset$, choose $v_{l+1}$ from $V_{\textnormal{\scriptsize rem}}$ uniformly at random and set $p' = (v_0, \ldots, v_l, v_{l+1})$.
	\item otherwise, use the same path $p$ also in the new solution $s'$.
\end{itemize}

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}. More precisely, the implementation of an algorithm/problem combination in PISA is divided into the problem-specific part (called \emph{variator module}) consisting of representation, initialization, and variation operators and the problem-independent part (called the \emph{selector module}). As selection module, any of the state-of-the-art selectors in PISA can be used off-the-shelf. In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{}, and W-HypE \citep{} with a focus on the first one.

After the initialization of the population with $\alpha$ individuals, those solutions are evaluated and their objective function values transferred to the selector module which then \TODO{write}

$\lambda$ individuals are chosen for variation and undergo the above described mutation operator creating the $\lambda$ so-called offspring.

 The objective vectors of the current population and these new solutions are then handed over to the selector module which decides which solutions to

\TODO{describe how PISA works}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In the remainder of this paper, we investigate how the above proposed evolutionary multiobjective optimization algorithm can cope with realistic instances of the multiobjective risk-equity constrained routing problem. The experimental analyses have two main goals. On the one hand, we show that the algorithm is efficient and can find multiple non-dominated solutions on practical examples in reasonable time. On the other hand, we investigate which of the two proposed initialization schemes is favourable and when this is the case. In addition, a few more experiments are conducted to show that the standard parameters of the algorithm behave as with other test problems. To this end, we vary the population size of the algorithm as well as its environmental selection scheme.

As a test case for all the experiments, we chose to use a real-world case study based on the work of \citet{bcg2009a}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome. We distinguish between instances with a different number of commodities (between 2 and 4) and with different numbers of trucks, see Table~\ref{tab:instances}. While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bcg2009a} and Table~\ref{tab:instances}. Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but with randomly permutated costs where each occuring cost value has been divided by a random value between 0 and 5000.

\begin{table}
\caption{\label{tab:instances} Parameters of the instances used in the experimental analyses with $|C|$ being the number of different commodities.}\vspace{1em}
\begin{tabular}{lllll}
	\toprule
	name & $|C|$ & \#trucks & cost range & risk range \\ \midrule
	lazio\_2 & 2 & \small 25+16         & \small [0.25, 41] & \begin{minipage}[c]{3.8cm} \small [762, 274628]\\ $\times$ [610.18, 219702.6]\end{minipage}\\
	\midrule
	lazio\_3 & 3 & \small 25+16+28      & \small [0.25, 41] & \begin{minipage}[c]{3.8cm} \small [762, 274628]\\ $\times$ [610.18, 219702.6]\\ $\times$ [1120.352, 439405.2]\end{minipage}\\
	\midrule
	lazio\_4\_1 & 4 & \small 21+3+30+5     &  & \multirow{4}{*}{\begin{minipage}[c]{3.8cm} \hspace{1pt}\\ \\ \small [762, 274628]\\ $\times$ [610.18, 219702.6]\\ $\times$ [1120.352, 439405.2]\\ $\times$ [762, 274628]\end{minipage}}\\
	\cmidrule{1 - 3}
	lazio\_4\_2 & 4 & \small 21+3+30+6     & \small [0.25, 41] & \\
	\cmidrule{1 - 3}
	lazio\_4\_3 & 4 & \small 7+1+10+2     &  & \\
	\cmidrule{1 - 4}
	uncorrelated\_4 & 4 & \small 21+3+30+5     & \begin{minipage}[t]{2cm} \small [0.0109239, 3.93701]  \end{minipage} & \\
	\bottomrule
\end{tabular}
\end{table}

%We concentrated our analysis on a real-world case study \citep{bcg2009a}. We considered the road network of the Lazio region (located in the middle of Italy), and, in particular, its main transport roads. The network is composed of 311 nodes and 441 links. The risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ are evaluated as the societal risk computed as the number of people living inside the exposure zone around link $(i,j)$ (whose size depends on the hazmat type $c$ times the accident probability involving the vehicle), for details see  \citep{bcg2009a}. We considered from 2 to 10 origin-destination pairs on the network, each one associated with a number of commodities ranging from 1 to 3, for an overall number of shipments (commodities) between 2 and 30 (we recall that in our model each commodity is associated with one origin-destination pair). With each one of the latter scenarios we associated 10 instances; each instance has been generated by assigning uniformly at random a demand from 100 to 1000 tons to each shipment.

\subsection{Experimental Settings}
Except otherwise stated, we used the following parameter setting. We perform 10 independent runs of HypE for 2000 generations with a population size of 50, i.e., each algorithm run has a budget of 100,000 function evaluations. The reference point for the hypervolume indicator was chosen as $(10^8, 10^8, 10^8)$ for the instances with 2 and 3 commodities and as $(10^9, 10^9, 10^9)$ for all instances with 4 commodities. The reference point has been chosen according to the objective function values of the random initial populations, such that all initial solutions have a positive contribution to the hypervolume indicator. Note that the reason for using the same values in all objectives is the used PISA implementation of HypE. In all cases, the hypervolume computation is exact.

\begin{table}
\centering
\caption{\label{tab:referencepoints} Used reference points within the environmental selection of HypE}
\vspace{1em}
\begin{tabular}{lr}
	\toprule
	instance & reference point\\ \midrule
	\# commodities: 2 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 3 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 4 & $(10^9, 10^9, 10^9)$\\
	\bottomrule
\end{tabular}
\end{table}

\subsection{Differences Between Instances}
The basis for our experiments are three instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1) which differ also differ in the number of trucks for the commodities. Whereas the number of trucks for lazio\_3 and lazio\_4\_1 have been chosen randomly, lazio\_2 has the same number of trucks than the lazio\_3 instance for the first two commodities.

\TODO{show results and talk about them}

\begin{figure}%
	\centering
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
\end{figure}

\subsection{Random Initialization Versus Cost-Optimal Initialization}

\begin{figure}%
	\includegraphics[width=\columnwidth]{../experiments/randVsCost/allsolutions.eps}%
	\caption{\label{fig:allsolutions} Comparison between the random initialization (circles) and the cost-optimal one (crosses). Shown are the populations of 10 independent HypE runs after 1000 generations for all three objectives (top left) and each combination of two objectives.}
\end{figure}

\begin{figure}%
	\includegraphics[width=\columnwidth]{../experiments/randVsCost/onlynondominated.eps}%
	\caption{\label{fig:onlynondominated} Comparison between the random initialization (circles) and the cost-optimal one (crosses). Shown are only the non-dominated solutions in the final populations of 10 independent HypE runs after 1000 generations.}
\end{figure}

\subsection{Comparison With An Exact Algorithm for the Two Linear Objectives}
\TODO{add plots and describe what we compare (efficient cost/risk paths multiplied by the number of trucks); result: EA is better}

\subsection{Changing the Population Size}
In a second experiment, we show that changing the population size of the algorithm in standard ranges does not change the algorithm outcome drastically. To this end, Fig.~\ref{fig:popsizes} shows the nondominated solutions of 10 independent runs after 100,000 function evaluations for three different versions of HypE where the population size changes from 50 over 100 to 200 on the largest instance lazio\_4\_1 with 4 commodities. From the visual inspection of the resulting solutions, we cannot make a statement about which population size to favour. When comparing the hypervolume values of the resulting populations, however, an effect is visible. Figure~\ref{fig:algoComparison} shows the corresponding box plots of the hypervolume values and one-sided Wilcoxon rank sum tests on the populations' hypervolume indicator values reveil statistical significant differences between two of the three possible algorithm pairs. With a $p$-value of $0.02163$, the null hypothesis of no differences between a population size of 50 and a population size of 100 has to be rejected in favor of stating that the algorithm variant with population size 50 is superior to the one with population size 100. For the comparison between population size 50 and population size 200, the $p$-value equals $0.0144$ with the population size 50 being better whereas the comparison between population size 100 and population size 200 does not reveal a statistical difference (the $p$-values of the one-sided Wilcoxon tests are $0.2179$ and $0.8035$ respectively).

Although these results indicate that smaller population sizes might be always favorable, this is not the case as additional experiments with a population size of 25 show (see the box plot in Fig.~\ref{fig:algoComparison}). The Wilcoxon rank sum test comparing population sizes 25 and 50 results in a $p$-value of $0.004465$ whereas the other comparisons are not statistically significantly different. To conclude, among the tested population sizes, 50 seems to be the best choice for budgets around 100,000 function evaluations. Also with longer runs, we recommend to keep the population size of the algorithm around 50 as the HypE implementation in PISA becomes too costly if the population size is much larger. Ten runs with 2000 iterations and population size 200, for example, need already around 20 hours to complete on an Intel Core2 Duo with 2.8GHz.

\begin{figure}%
	\centering
	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/diffPopsizes}%
	\caption{\label{fig:popsizes} Nondominated solutions of 10 independent runs after 100,000 function evaluations for HypE with a population size of 50 ($\times$), 100 ($\circ$), and 200 ($\square$) respectively.}
\end{figure}

\begin{figure}%
	\centering
	\includegraphics[width=0.5\columnwidth]{../experiments/randVsCost/hypervolumes/algoComparison}%
	\caption{\label{fig:algoComparison} Boxplots of the hypervolume values of 10 independent runs after 100,000 function evaluations of HypE with a population size of 25, 50, 100, and 200, as well as for NSGA-II.}
\end{figure}

\subsection{Changing the Environmental Selection}
With a last experiment, we show the influence of the algorithm's environmental selection scheme on the performance and in particular how this can be used to steer the search towards specific regions of the Pareto front.
\TODO{write}

\subsection{The Feasibility of the EMO Approach}
Due to their heuristic nature, evolutionary multiobjective optimization algorithms like the one presented in this study cannot guarantee that the solutions found are actually Pareto-optimal. However, as we have seen above, the proposed algorithm is able to produce a set of non-dominated solutions in reasonable time which are close to Pareto-optimal solutions found by exact algorithms. Such sets of non-dominated solutions can be a starting point for a decision maker to investigate the trade-offs between the objectives and to decide in which objective space region interesting Pareto-optimal solutions are which can be explored further with more involved algorithms---be it exact algorithms or specialized evolutionary algorithms such as the mentioned W-HypE variant. In the above example instances, it turned out that total risk and cost are indeed more or less fully correlated in the original problem formulation or only loosely (with a larger trade-off between cost and total risk) when the individual arcs' risks and costs are.

Regarding the two initialization schemes, it turned out that in combination with the (local and simple) variation operator, the cost-efficient initialization does not allow the algorithm to find additional solutions besides the initial one and that, in particular if the cost and total risk objectives are correlated, the random initialization is beneficial. Using other initializations as well as additional variation operators from the literature is easily possible with the proposed algorithm but their investigation is left for future research\footnote{For example, the

On the other hand, the proposed variation operator can be changed towards a larger neighborhood by allowing two or more nodes to be added or deleted from each truck path at a time---resulting in more exploration than exploitation when compared to the current variation operator. Also well-designed crossover operators are expected to be beneficial to the problem when, for example, good sub paths can be exchanged between solution paris.
}.

\begin{itemize}

	\item approach is very general and can easily deal with slightly modified problems: e.g. different truck loads (only the objective function code needs to be changed)
	\item it needs to be mentioned that for the tested instances with 311 nodes and 441 arcs, the variation operator of the evolutionary algorithm was able to construct the shortest paths from source to destination in some of the runs for some of the truck paths within the first 100,000 function evaluations. This corresponds to performing random walks of lengths around 20 nodes.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions} \label{S_FW}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest. Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary algorithm to cope with the non-linear equity risk objective.

The third objective function of our problem can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$. Although this equivalent formulation makes the problem linear (with additional linear constraints), classical algorithms are expected to have difficulties with this formulation as well and our algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints. Note that, for the moment, the proposed EMO algorithm exists on paper only and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to generate solutions of sufficient quality and whether adaptively changing the number and capacity of trucks is beneficial.

In future work, one might want to look further into adding (simple) heuristics within the evoluionary algorithm in order to improve the algorithm. For example, one can observe that the solutions produced by the algorithm sometimes uses one and the same arc within a truck path, hence introducing cylces. With the three objectives cost, total and equity risk, such cycles can be easily prevented within the variation operator which might eventually result in a faster convergence of the algorithm to the Pareto front.

% The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest.

% In this study, we formalize transportation of hazmats problem as the minimization of three objectives and propose to use an evolutionary algorithm since many exact optimization procedures have difficulties with the non-linear equity risk objective. Note that our problem formulation can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$ \COMMENTD{this has to be adapted according to the new problem formulation}. Although this equivalent formulation makes the problem linear (with additional non-linear constraints), classical algorithms are expected to have difficulties with this formulation as well and the EMO algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints.
% 
% Note also that, for the moment, the proposed EMO algorithm only exists as a gedankenexperiment and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to improve the quality of the found solutions to a sufficient level.

\section{Acknowledgments}
The authors would like to thank Stefano Giordani for providing us with the instances of the Lazio network. Moreover, Dimo Brockhoff would like to acknowledge support from the CNRS-Microsoft chair "`Optimization for Sustainable Development"' during his time at Ecole Polytechnique, Palaiseau, France.

% 
 \bibliographystyle{plainnat}
 \footnotesize
 \bibliography{all}

\end{document}

%%
%% End of file `elsarticle-template-1a-num.tex'.

### 2011-11-29 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
a bit more text (incrementally)

--- 2011-11-29 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
@@
 As selection module, any of the state-of-the-art selectors in PISA can be used off-the-shelf.
-In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{}, and W-HypE \citep{} with a focus on the first one.
+In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{} with a focus on the first one.
 
@@
 With a last experiment, we show the influence of the algorithm's environmental selection scheme on the performance and in particular how this can be used to steer the search towards specific regions of the Pareto front.
+
+Instead of the environmental selection of HypE \citep{bz2011a}, other selection schemes can be easily used with the PISA framework and we show the resutls of two other algorithms exemplary.
+On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}.
+The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations) are shown in Fig.~\ref{fig:algoComparison}.
+\TODO{What can be seen is that both?
+algorithms show a reduced performance regarding the standard hypervolume value.
+However, W-HypE was run in order to show how the search can be steered towards regions of the objective space that have been specified by the user.
+In our case, we defined the weight of the underlying weighted hypervolume indicator \citep{zbt2007a} as a box with corners XXX such that the weighted hypervolume indicator only influences BLABLABLA}
+
 \TODO{write}

### 2011-11-29 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
new figure for comparing W-Hype with HypE and corresponding (preliminary) text

@@
 
-\begin{table}
-\centering
-\caption{\label{tab:referencepoints} Used reference points within the environmental selection of HypE}
-\vspace{1em}
-\begin{tabular}{lr}
-	\toprule
-	instance & reference point\\ \midrule
-	\# commodities: 2 & $(10^8, 10^8, 10^8)$\\
-	\# commodities: 3 & $(10^8, 10^8, 10^8)$\\
-	\# commodities: 4 & $(10^9, 10^9, 10^9)$\\
-	\bottomrule
-\end{tabular}
-\end{table}
-
 \subsection{Differences Between Instances}
@@
 
-\begin{figure}%
+\begin{figure}
 	\centering
@@
 
-\begin{figure}%
-	\includegraphics[width=\columnwidth]{../experiments/randVsCost/allsolutions.eps}%
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/allsolutions.eps}%
 	\caption{\label{fig:allsolutions} Comparison between the random initialization (circles) and the cost-optimal one (crosses).
@@
 
-\begin{figure}%
-	\includegraphics[width=\columnwidth]{../experiments/randVsCost/onlynondominated.eps}%
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/onlynondominated.eps}%
 	\caption{\label{fig:onlynondominated} Comparison between the random initialization (circles) and the cost-optimal one (crosses).
@@
 With a last experiment, we show the influence of the algorithm's environmental selection scheme on the performance and in particular how this can be used to steer the search towards specific regions of the Pareto front.
+
+Instead of the environmental selection of HypE \citep{bz2011a}, other selection schemes can be easily used with the PISA framework and we show the resutls of two other algorithms exemplary.
+On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}.
+The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations) are shown in Fig.~\ref{fig:algoComparison}.
+\TODO{What can be seen is that both?
+algorithms show a reduced performance regarding the standard hypervolume value.
+For NSGA-II, this is expected as its environmental selection schemes becomes less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}.
+This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing non-compliant quality measures such as NSGA-II.
+
+The reason for the reduced performance in the standard hypervolume indicator for W-Hype, however, is a different one.
+W-HypE was developed to steer the search towards regions of the objective space that have been specified by the user.
+To this end, it does not optimize the standard hypervolume indicator but a more generalized version of it, the weighted hypervolume indicator \citep{zbt2007a}.
+The idea behind W-HypE is that solutions in objective space regions with a higher weight will get a higher fitness due to the \emph{weighted} hypervolume they dominate.
+In order to be efficient also for a large number of objectives, W-HypE does not calculate the weighted hypervolume indicator values exactly (which is not possible for most of the weight functions) but instead samples them.
+Here, 10,000 samples are used in each iteration of the algorithm.
+
+To show the principles of W-HypE in the application of our hazmat routing problem, we defined the weight of W-HypE's underlying weighted hypervolume indicator as constant in a box with corners $[7\cdot 10^3, 3\cdot 10^7, 5\cdot 10^5]$ and $[10^4, 4\cdot 10^7, 1.1\cdot 10^6]$ and zero otherwise.
+Instead of optimizing the standard hypervolume indicator as HypE, W-HypE now optimizes this weighted hypervolume indicator, resulting in better, i.e., dominating solutions in regions in or close to the box.
+Figure~\ref{fig:whype} shows the resulting non-dominated solutions of both algorithms with population size 25 after 10 independent runs of 4,000 generations each as well as the sampling box.}
+
+This advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations, cp.\ Fig.~\ref{fig:whype}.
+
 \TODO{write}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/WHypEVsHypE_ns4_1_OriginalCosts2.eps}%
+	\caption{\label{fig:whype} Comparison between the standard HypE selection ($\circ$) with the one of W-HypE ($\times$) when the weighted hypervolume indicator is only sampled in the gray shaded box $[7\cdot 10^3, 10^4]\times[3\cdot 10^7, 4\cdot 10^7]\times[5\cdot 10^5, 1.1\cdot 10^6]$.
+Shown are all non-dominated solutions for each algorithm after 10 runs with 4,000 generations and population size 25.}
+\end{figure}
 

### 2011-11-29 trunk/DAW_hazmatHypE/trunk/experiments/randVsCost/hypervolumes/plotBoxplotRandAndCost.m
added new boxplots comparing initialization schemes

@@ -1,0 +1,45 @@
+% plot resulting populations of HypE with rand and cost initialization
+% after 100,000 funevals to compare initialization schemes
+
+addpath('../');
+
+hypervolumes = zeros(10, 8);
+labels = {'rand_250'; 'rand_2000'; 'cost_250'; 'cost_2000'; 'rand_250'; 'rand_2000'; 'cost_250'; 'cost_2000'};
+
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize50_randInit.250';
+hypervolumes(:,1) = -1 * dlmread(filename);
+
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize50_randInit.2000';
+hypervolumes(:,2) = -1 * dlmread(filename);
+
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize50_costInit.250';
+hypervolumes(:,3) = -1 * dlmread(filename);
+
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize50_costInit.2000';
+hypervolumes(:,4) = -1 * dlmread(filename);
+
+filename = 'HYP_ns4_1_New_HypE_popsize50_randInit.250';
+hypervolumes(:,5) = -1 * dlmread(filename);
+
+filename = 'HYP_ns4_1_New_HypE_popsize50_randInit.2000';
+hypervolumes(:,6) = -1 * dlmread(filename);
+
+filename = 'HYP_ns4_1_New_HypE_popsize50_costInit.250';
+hypervolumes(:,7) = -1 * dlmread(filename);
+
+filename = 'HYP_ns4_1_New_HypE_popsize50_costInit.2000';
+hypervolumes(:,8) = -1 * dlmread(filename);
+
+
+
+boxplot(hypervolumes(:,1:4), 'Labels', labels(1:4));
+% save plot as eps:
+exportfig(gcf, 'initBoxPlots_OriginalCosts.eps');
+
+
+figure;
+boxplot(hypervolumes(:,5:8), 'Labels', labels(5:8));
+% save plot as eps:
+exportfig(gcf, 'initBoxPlots_New.eps');
+
+

### 2011-11-29 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added new plots comparing the initialization schemes into paper

@@
 \newcommand{\COMMENTN}[1]{{\color{green}Nora: #1}}
+
+\renewcommand{\topfraction}{.70}
+\renewcommand{\textfraction}{.20}
 
@@
 	\centering
-	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/allsolutions.eps}%
-	\caption{\label{fig:allsolutions} Comparison between the random initialization (circles) and the cost-optimal one (crosses).
-Shown are the populations of 10 independent HypE runs after 1000 generations for all three objectives (top left) and each combination of two objectives.}
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/randVsCost_ns4_1_OriginalCost_2000gens.eps}
+	\vspace{-1em}
+	\caption{\label{fig:allsolutions} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$).
+Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations for all three objectives (top left) and each combination of two objectives on the lazio\_4\_1 instance.}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/onlynondominated.eps}%
-	\caption{\label{fig:onlynondominated} Comparison between the random initialization (circles) and the cost-optimal one (crosses).
-Shown are only the non-dominated solutions in the final populations of 10 independent HypE runs after 1000 generations.}
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/randVsCost_ns4_1_New_2000gens.eps}
+	\vspace{-1em}
+	\caption{\label{fig:onlynondominated} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$).
+Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations on the uncorrelated\_4 instance.}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/initBoxPlots_OriginalCosts.eps}
+	\hfill
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/initBoxPlots_New.eps}
+	\vspace{-1em}
+	\caption{\label{fig:costVsRandBoxPlots} Comparison between the random initialization and the cost-optimal one.
+	Shown are boxplots of hypervolume values for HypE with population size 50 on the lazio\_4 instance (left) and on the uncorrelated\_4 instance (right) after both 250 and 2000 generations.}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/diffPopsizes}%
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/diffPopsizes}%
+	\vspace{-1em}
 	\caption{\label{fig:popsizes} Nondominated solutions of 10 independent runs after 100,000 function evaluations for HypE with a population size of 50 ($\times$), 100 ($\circ$), and 200 ($\square$) respectively.}
@@
 	\centering
-	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/WHypEVsHypE_ns4_1_OriginalCosts2.eps}%
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/WHypEVsHypE_ns4_1_OriginalCosts2.eps}
+	\vspace{-1em}
 	\caption{\label{fig:whype} Comparison between the standard HypE selection ($\circ$) with the one of W-HypE ($\times$) when the weighted hypervolume indicator is only sampled in the gray shaded box $[7\cdot 10^3, 10^4]\times[3\cdot 10^7, 4\cdot 10^7]\times[5\cdot 10^5, 1.1\cdot 10^6]$.

### 2011-11-29 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
minor changes in description of instances

@@
 \subsection{Differences Between Instances}
-The basis for our experiments are three instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1) which differ also differ in the number of trucks for the commodities.
+The basis for our experiments are 5 instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1, lazio\_4\_2, and lazio\_4\_3) which differ also in the number of trucks for the commodities.
 Whereas the number of trucks for lazio\_3 and lazio\_4\_1 have been chosen randomly, lazio\_2 has the same number of trucks than the lazio\_3 instance for the first two commodities.
+The lazio\_4\_2 instance has only 1 truck more for commodity 4 than lazio\_4\_1 and overall three times the number of trucks for each commodity than the instance lazio\_4\_3.
 

### 2011-11-29 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added first version of abstract, deleted planned section on comparison with exact approaches due to missing results

@@
 \begin{abstract}
-tbw
+Nowadays, various hazardous materials, or \emph{hazmats}, are transported on our streets, via trains, planes or ships.
+In paricular when routing them over the street network, where typically several possible routes exist, one would like to minimize both the costs and the total risks.
+However, sometimes one aims at minimizing an additional type of risk, which in most optimization models is resulting in a non-linear objective: the equity risk, or the \emph{maximum} risk over all regions, streets, etc.
+In other words, one aims at distributing the overall risks as much as possible while the costs and the total risk should be still minimized simultaneously.
+In this paper, this three-objective optimization problem is tackled with a novel randomized algorithm, based on a state-of-the-art Evolutionary Multiobjective Optimization algorithm and taylored to the needs of finding Pareto-optimal truck routes in a given network by proposing a variable-length representation and problem-specific operators.
+Experimental results on realistic instances for the street network of Italy's Lazio region show that the proposed approach efficiently finds a variety of good solutions in reasonable time.
+The experiments also show the influence of the algorithm's different building blocks and the instances' parameters on the solution quality.
 \end{abstract}
@@
 \begin{keyword}
-Evolutionary multiobjective optimization, Transportation of hazardous materials, Risk equity, Shortest Paths. 
+Evolutionary Multiobjective Optimization, Transportation of hazardous materials, Risk equity, Shortest Paths. 
 \end{keyword}
@@
 
-\subsection{Comparison With An Exact Algorithm for the Two Linear Objectives}
-\TODO{add plots and describe what we compare (efficient cost/risk paths multiplied by the number of trucks); result: EA is better}
-
 \subsection{Changing the Population Size}

### 2011-11-30 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
finished description of PISA part

@@
 In order to have every variable length path represent an uninterrupted path from source to destination at any time (see the constraints in (\ref{eq:flow})), we ensure all paths to always start with the source $s^c$ for the corresponding commodity $c$, ensure during initialization and variation that all neighbored vertices in the path are connected by an arc (see Sec.~\ref{sec:init} and \ref{sec:variation}), and complete each path by the shortest path between the path's actual end node and the commodity's destination node $t^c$.
+The shortest path is thereby calculated exactly with respect to the given costs of the arcs.
 The right hand side of Fig.~\ref{fig:paths} shows the entire source--destination paths corresponding to the representation in the left part of the figure.
@@
 \subsection{Initialization}\label{sec:init}
-Often, the population of a (multiobjective) evolutionary algorithm is initialized with solutions that are drawn uniformly at random within the search space.
-Although it is not easy to define a uniformity measure in our case of variable-length representation, we still propose a random initialization scheme for our evolutionary algorithm.
+Often, the population of a (multiobjective) evolutionary algorithm is initialized with solutions that are drawn uniformly at random within the search space \citep{deb2001a}.
+Although it is not easy to define a uniformity measure in our case of a variable-length representation, we still propose a random initialization scheme for our evolutionary algorithm.
 In addition, the specific problem formulation allows another initialization where a Pareto-optimal solution is inserted in the initial population.
@@
 \end{enumerate}
-Note that this initialization scheme is, though random, does not produce fully random paths.
-The initialization results in truck paths which are constructed by concatenating two shortest paths---the shortest path from the source node to the one randomly selected and the shortest path from the random node to the destination, which is added due to the representation described above.
+Note that this initialization scheme, though random, does not produce fully random paths.
+Instead, the initialization results in truck paths which are constructed by concatenating two shortest paths---the shortest path from the source node to the one randomly selected and the shortest path from the random node to the destination, which is added due to the representation described above.
 
@@
 As mutation operator, we suggest to shorten or lengthen the path of one or several trucks.
-In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as in \citep{horo2009a}:
+In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as suggested and analyzed in \citep{horo2009a}:
 \begin{itemize}
@@
 More precisely, the implementation of an algorithm/problem combination in PISA is divided into the problem-specific part (called \emph{variator module}) consisting of representation, initialization, and variation operators and the problem-independent part (called the \emph{selector module}).
-As selection module, any of the state-of-the-art selectors in PISA can be used off-the-shelf.
-In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{} with a focus on the first one.
+As selection module, any of the state-of-the-art selectors provided by PISA can be used off-the-shelf.
+In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{abbz2009b,bbtz2011a} with a focus on the first one.
 
-After the initialization of the population with $\alpha$ individuals, those solutions are evaluated and their objective function values transferred to the selector module which then \TODO{write}
+After the initialization of the population with $\alpha$ individuals by the variator, their evaluation and sending the objective values to the selector module, a standard PISA run always continues in the same way.
+The selector decides which $\mu$ solutions to variate and the variator uses those $\mu$ solutions as starting points for the generation of $\lambda$ new search points which are again evaluated and send to the selector.
+In addition, the selector can decide whether solutions are stored in an archive or whether they are deleted.
+In the case of the selectors used in this study, no archive is maintained.
+The search stops after a predefined number of variator/selector iterations, also called generations, are performed.
+In our case, $\alpha=\mu=\lambda$, which will be also called population size hereafter.
+For more details about PISA, we refer to \citep{bltz2003a}.
 
-$\lambda$ individuals are chosen for variation and undergo the above described mutation operator creating the $\lambda$ so-called offspring.
-
- The objective vectors of the current population and these new solutions are then handed over to the selector module which decides which solutions to
-
-\TODO{describe how PISA works}
+\subsection{The HypE and NSGA-II Selectors of PISA}
+\TODO{write}
 

### 2011-11-30 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
wrote HypE part

@@
 
-\subsection{The HypE and NSGA-II Selectors of PISA}
-\TODO{write}
+\subsection{Hypervolume Estimation Algorithm for Multiobjective Optimization: The HypE Selector of PISA}
+In order to fully describe the proposed EMO algorithm for the multiobjective risk-equity constrained routing problem, we briefly mention the working principles of the HypE selection.
+The main purpose of an EMO algorithm's environmental selection is to ensure convergence of the population towards the Pareto front as well as a good spread of it along the front \citep{deb2001a}.
+HypE persues these two goals by a two-step approach based on the hypervolume indicator \citep[see][]{ztlf2003a}.
+
+Given the $\mu$ solutions of the old population and the $\lambda$ new ones, the best $\mu$ of them are chosen in the following way.
+First, a non-dominated sorting \citep{gold1989a} is performed---identifying sets of non-dominated fronts with different ranks.
+The solutions that are not dominated in the entire set get a rank of 1, the solutions of rank 2 are those which are non-dominated among the remaining solutions, and so forth.
+As long as the number of solutions is not exceeding $\mu$, entire fronts are chosen for the next iteration starting from the front with rank 1.
+For the first front for which the overall number of solutions exceeds $\mu$, a second criterion is used to determine the solutions to be kept for the next iteration.
+This decision is made according to the (expected) loss in hypervolume, which is a criterion that is refining the underlying Pareto dominance relation \citep{ztb2010a}---the main reason why many recent EMO algorithms employ the hypervolume indicator.
 

### 2011-11-30 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added description of Sec. 4.2 about the differences between the instances (including new plot)

@@
 
-\TODO{show results and talk about them}
+Figure~\ref{fig:compareAll} shows the non-dominated solutions of 10 independent HypE runs for each instance.
+What can be observed are three main statements.
+First of all, cost and risks of the found non-dominated solutions are highly correlated, as they are for each single arc in the problem formulations; therefore, the problem is not truly three-objective but rather bi-objective.
+Secondly, the range of the objective function values for the found non-dominated solutions follow the number of trucks in the problem instance, i.e., the more trucks a problem instance has the more costs, total and equity risk can be expected in the solutions found.
+However and thirdly, this relationship is not linear in the plots of Fig.~\ref{fig:compareAll} as can be seen, for example, between the lazio\_4\_2 and the lazio\_4\_3 instance.
+Though the number of trucks per commodity differs by a factor of 3 between them, the objective function values of the resulting non-dominated solutions after 1,000 generations are slightly further away.
+The reason for that is the much smaller search space for the lazio\_4\_3 instance which can be searched more exhaustively by the algorithm with the same number of function evalutions---resulting in a better convergence towards the Pareto front.
 
@@
 	\centering
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
-	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/compAll_2000gens.eps}
+	\caption{\label{fig:compareAll} Comparison between the different instances lazio\_2 (squares), lazio\_3 (circles), lazio\_4\_1 (diamonds), lazio\_4\_2 (crosses), and lazio\_4\_3 (stars).
+Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations.}
 \end{figure}
@@
 Ten runs with 2000 iterations and population size 200, for example, need already around 20 hours to complete on an Intel Core2 Duo with 2.8GHz.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
+	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
+\end{figure}
 

### 2011-11-30 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added some more text about performance assessment

@@
 
+In terms of performance assessment, we present both visual comparisons of the final populations as well as statistical results.
+For the statistical results, the hypervolume indicator was also used as quality measure of the final populations as it is the only known unary quality indicator  which is a refinement of the Pareto dominance relation and therefore has important theoretical properties: whenever a set of solutions dominates another set, also the hypervolume value is higher; on the contrary, if the hypervolume indicator value of a set A is worse than for set B, we know that B cannot dominate A \citep{ztlf2003a,ztb2010a}.
+In order to compare algorithms with respect to their hypervolume indicator values, we use both box plots\footnote{As implemented in MATLAB R2008b.} and the Wilcoxon rank sum test\footnote{As implemented in the statistical programming language R, version 2.1.3.0.}.
+
 \subsection{Differences Between Instances}
@@
 \subsection{Random Initialization Versus Cost-Optimal Initialization}
+In order to see the impact of the two different initialization schemes \emph{random} and \emph{cost-optimal}, we performed 10 runs of HypE with both initialization schemes for 2,000 generations and a population size of 50 as above.
+Figure~\ref{fig:randVsCostOriginal} shows the resulting non-dominated solutions of the final populations.
+Surprisingly, the cost-optimal initialization could not produce other solutions than the cost-optimal one itself.
+\TODO{However, what can be seen is that the ...}
+
+Boxplots can be found in Fig.~\ref{fig:costVsRandBoxPlots} ...
 
@@
 	\vspace{-1em}
-	\caption{\label{fig:allsolutions} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$).
+	\caption{\label{fig:randVsCostOriginal} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$).
 Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations for all three objectives (top left) and each combination of two objectives on the lazio\_4\_1 instance.}
@@
 	\vspace{-1em}
-	\caption{\label{fig:onlynondominated} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$).
+	\caption{\label{fig:randVsCostNew} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$).
 Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations on the uncorrelated\_4 instance.}

### 2011-11-30 DAW_hazmatHypE/trunk/experiments/randVsCost/plotCompareRandAndCost.m
added missing results of final runs (W-HypE, graph_ns3_1)

@@ -3,34 +3,47 @@
 
 %%%%%%%%%%%%%%%%%%%%%%%
 generation = 2000;
-numOfCommodities = 2;
-instance = 2;
 %%%%%%%%%%%%%%%%%%%%%%%
 
-filename = ['results/graph_ns' int2str(numOfCommodities) '_' int2str(instance) '_New_costInit.' int2str(generation)];
+% first ns4_1_New:
+%%%%%%%%%%%%%%%%%%
+filename = ['results/graph_ns4_1_New_HypE_popsize50_costInit.' int2str(generation)];
 C = dlmread(filename);
 C_nondom = C(find(paretofront(C(:,1:3))),:);
 figure(77);
-%plot3(C(:,1),C(:,2),C(:,3), 'x');
 plot3(C_nondom(:,1),C_nondom(:,2),C_nondom(:,3), 'o');
 hold all; grid on;
 
-filename = ['results/graph_ns' int2str(numOfCommodities) '_' int2str(instance) '_New_randInit.' int2str(generation)];
+filename = ['results/graph_ns4_1_New_HypE_popsize50_randInit.' int2str(generation)];
 R = dlmread(filename);
 R_nondom = R(find(paretofront(R(:,1:3))),:);
-%figure()
-%plot3(R(:,1),R(:,2),R(:,3), 'x');
-%hold all; grid on;
 plot3(R_nondom(:,1),R_nondom(:,2),R_nondom(:,3), 'o');
 hold off;
 
-
-%mysubplot(C,C_nondom, 'C', 'C_nondom', 'C only', 'filename');
-%mysubplot(R,R_nondom, 'R', 'R_nondom', 'R only', 'filename');
-
-title = ['all solutions after ' int2str(generation) ' generations'];
-mysubplot(C,R, 'shortest path initialization', 'random initialization', title, 'allsolutions.eps');
 title = ['nondominated solutions after ' int2str(generation) ' generations'];
-mysubplot(C_nondom,R_nondom, 'shortest path initialization', 'random initialization', title, 'onlynondominated.eps');
+filename = ['randVsCost_ns4_1_New_' int2str(generation) 'gens.eps'];
+mysubplotRandVsCost(R_nondom, C_nondom, 'random', 'shortest path', title, filename);
+
+hold off;
+
+
+% now ns4_1_OriginalCost:
+%%%%%%%%%%%%%%%%%%%%%%%%%%
[…]

### 2011-11-30 DAW_hazmatHypE/trunk/experiments/randVsCost/plotCompareWHypE.m

@@ -1,0 +1,37 @@
+% plot resulting populations of W-HypE vs. HypE
+% after certain number of generations
+
+%%%%%%%%%%%%%%%%%%%%%%%
+generation = 4000;
+%%%%%%%%%%%%%%%%%%%%%%%
+
+% sampling box:
+Supperlower = [8e3 3e7 0.5e6; 1e4 4e7 1e6];
+
+figure(1234);
+S = UniformSampling(10000,Supperlower(1,:), Supperlower(2,:));
+plot3(S(:,1),S(:,2),S(:,3), 'y.', 'MarkerSize', 2);
+hold all; grid on;
+
+% first W-HypE:
+%%%%%%%%%%%%%%%
+filename = ['results/graph_ns4_1_OriginalCosts_WHypE_box_popsize25_randInit.' int2str(generation)];
+W = dlmread(filename);
+W_nondom = W(find(paretofront(W(:,1:3))),:);
+plot3(W_nondom(:,1),W_nondom(:,2),W_nondom(:,3), 'o');
+
+filename = ['results/graph_ns4_1_OriginalCosts_HypE_popsize25_randInit.' int2str(generation)];
+H = dlmread(filename);
+H_nondom = H(find(paretofront(H(:,1:3))),:);
+plot3(H_nondom(:,1),H_nondom(:,2),H_nondom(:,3), 'o');
+
+all = [W_nondom; H_nondom];
+all = all(find(paretofront(all(:,1:3))),:);
+plot3(all(:,1),all(:,2),all(:,3), 'sr');
+hold off;
+
+
+title = ['nondominated solutions after ' int2str(generation) ' generations'];
+mysubplot3sampling(W_nondom, H_nondom, Supperlower, 'W-HypE', 'HypE', 'sample box', title, 'WHypEVsHypE_ns4_1_OriginalCosts.eps');
+
+hold off;

### 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added missing results of final runs (W-HypE, graph_ns3_1)

@@
 
+\renewcommand{\topfraction}{.70}
+\renewcommand{\textfraction}{.20}
+
 \journal{DAM}
@@
 More precisely, the implementation of an algorithm/problem combination in PISA is divided into the problem-specific part (called \emph{variator module}) consisting of representation, initialization, and variation operators and the problem-independent part (called the \emph{selector module}).
-As selection module, any of the state-of-the-art selectors in PISA can be used off-the-shelf.
-In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{} with a focus on the first one.
+As selection module, any of the state-of-the-art selectors provided by PISA can be used off-the-shelf.
+In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{abbz2009b,bbtz2011a} with a focus on the first one.
 
-After the initialization of the population with $\alpha$ individuals, those solutions are evaluated and their objective function values transferred to the selector module which then \TODO{write}
+After the initialization of the population with $\alpha$ individuals by the variator, their evaluation and sending the objective values to the selector module, a standard PISA run always continues in the same way.
+The selector decides which $\mu$ solutions to variate and the variator uses those $\mu$ solutions as starting points for the generation of $\lambda$ new search points which are again evaluated and send to the selector.
+In addition, the selector can decide whether solutions are stored in an archive or whether they are deleted.
+In the case of the selectors used in this study, no archive is maintained.
+The search stops after a predefined number of variator/selector iterations, also called generations, are performed.
+In our case, $\alpha=\mu=\lambda$, which will be also called population size hereafter.
+For more details about PISA, we refer to \citep{bltz2003a}.
 
-$\lambda$ individuals are chosen for variation and undergo the above described mutation operator creating the $\lambda$ so-called offspring.
+\subsection{Hypervolume Estimation Algorithm for Multiobjective Optimization: The HypE Selector of PISA}
+In order to fully describe the proposed EMO algorithm for the multiobjective risk-equity constrained routing problem, we briefly mention the working principles of the HypE selection.
+The main purpose of an EMO algorithm's environmental selection is to ensure convergence of the population towards the Pareto front as well as a good spread of it along the front \citep{deb2001a}.
+HypE persues these two goals by a two-step approach based on the hypervolume indicator \citep[see][]{ztlf2003a}.
 
- The objective vectors of the current population and these new solutions are then handed over to the selector module which decides which solutions to
-
-\TODO{describe how PISA works}
+Given the $\mu$ solutions of the old population and the $\lambda$ new ones, the best $\mu$ of them are chosen in the following way.
+First, a non-dominated sorting \citep{gold1989a} is performed---identifying sets of non-dominated fronts with different ranks.
+The solutions that are not dominated in the entire set get a rank of 1, the solutions of rank 2 are those which are non-dominated among the remaining solutions, and so forth.
+As long as the number of solutions is not exceeding $\mu$, entire fronts are chosen for the next iteration starting from the front with rank 1.
+For the first front for which the overall number of solutions exceeds $\mu$, a second criterion is used to determine the solutions to be kept for the next iteration.
+This decision is made according to the (expected) loss in hypervolume, which is a criterion that is refining the underlying Pareto dominance relation \citep{ztb2010a}---the main reason why many recent EMO algorithms employ the hypervolume indicator.
 
@@
 
-\begin{table}
-\centering
-\caption{\label{tab:referencepoints} Used reference points within the environmental selection of HypE}
-\vspace{1em}
-\begin{tabular}{lr}
-	\toprule
-	instance & reference point\\ \midrule
-	\# commodities: 2 & $(10^8, 10^8, 10^8)$\\
-	\# commodities: 3 & $(10^8, 10^8, 10^8)$\\
-	\# commodities: 4 & $(10^9, 10^9, 10^9)$\\
-	\bottomrule
-\end{tabular}
-\end{table}
+In terms of performance assessment, we present both visual comparisons of the final populations as well as statistical results.
+For the statistical results, the hypervolume indicator was also used as quality measure of the final populations as it is the only known unary quality indicator  which is a refinement of the Pareto dominance relation and therefore has important theoretical properties: whenever a set of solutions dominates another set, also the hypervolume value is higher; on the contrary, if the hypervolume indicator value of a set A is worse than for set B, we know that B cannot dominate A \citep{ztlf2003a,ztb2010a}.
+In order to compare algorithms with respect to their hypervolume indicator values, we use both box plots\footnote{As implemented in MATLAB R2008b.} and the Wilcoxon rank sum test\footnote{As implemented in the statistical programming language R, version 2.1.3.0.}.
 
 \subsection{Differences Between Instances}
-The basis for our experiments are three instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1) which differ also differ in the number of trucks for the commodities.
+The basis for our experiments are 5 instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1, lazio\_4\_2, and lazio\_4\_3) which differ also in the number of trucks for the commodities.
 Whereas the number of trucks for lazio\_3 and lazio\_4\_1 have been chosen randomly, lazio\_2 has the same number of trucks than the lazio\_3 instance for the first two commodities.
+The lazio\_4\_2 instance has only 1 truck more for commodity 4 than lazio\_4\_1 and overall three times the number of trucks for each commodity than the instance lazio\_4\_3.
 
-\TODO{show results and talk about them}
+Figure~\ref{fig:compareAll} shows the non-dominated solutions of 10 independent HypE runs for each instance.
+What can be observed are three main statements.
+First of all, cost and risks of the found non-dominated solutions are highly correlated, as they are for each single arc in the problem formulations; therefore, the problem is not truly three-objective but rather bi-objective.
+Secondly, the range of the objective function values for the found non-dominated solutions follow the number of trucks in the problem instance, i.e., the more trucks a problem instance has the more costs, total and equity risk can be expected in the solutions found.
+However and thirdly, this relationship is not linear in the plots of Fig.~\ref{fig:compareAll} as can be seen, for example, between the lazio\_4\_2 and the lazio\_4\_3 instance.
+Though the number of trucks per commodity differs by a factor of 3 between them, the objective function values of the resulting non-dominated solutions after 1,000 generations are slightly further away.
+The reason for that is the much smaller search space for the lazio\_4\_3 instance which can be searched more exhaustively by the algorithm with the same number of function evalutions---resulting in a better convergence towards the Pareto front.
 
-\begin{figure}%
+\begin{figure}
 	\centering
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
-	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/compAll_2000gens.eps}
+	\caption{\label{fig:compareAll} Comparison between the different instances lazio\_2 (squares), lazio\_3 (circles), lazio\_4\_1 (diamonds), lazio\_4\_2 (crosses), and lazio\_4\_3 (stars).
+Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations.}
 \end{figure}
@@
 \subsection{Random Initialization Versus Cost-Optimal Initialization}
+In order to see the impact of the two different initialization schemes \emph{random} and \emph{cost-optimal}, we performed 10 runs of HypE with both initialization schemes for 2,000 generations and a population size of 50 as above.
+Figure~\ref{fig:randVsCostOriginal} shows the resulting non-dominated solutions of the final populations.
+Surprisingly, the cost-optimal initialization could not produce other solutions than the cost-optimal one itself.
+\TODO{However, what can be seen is that the ...}
 
-\begin{figure}%
-	\includegraphics[width=\columnwidth]{../experiments/randVsCost/allsolutions.eps}%
-	\caption{\label{fig:allsolutions} Comparison between the random initialization (circles) and the cost-optimal one (crosses).
-Shown are the populations of 10 independent HypE runs after 1000 generations for all three objectives (top left) and each combination of two objectives.}
+Boxplots can be found in Fig.~\ref{fig:costVsRandBoxPlots} ...
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/randVsCost_ns4_1_OriginalCost_2000gens.eps}
+	\vspace{-1em}
+	\caption{\label{fig:randVsCostOriginal} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$).
+Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations for all three objectives (top left) and each combination of two objectives on the lazio\_4\_1 instance.}
 \end{figure}
 
-\begin{figure}%
-	\includegraphics[width=\columnwidth]{../experiments/randVsCost/onlynondominated.eps}%
-	\caption{\label{fig:onlynondominated} Comparison between the random initialization (circles) and the cost-optimal one (crosses).
-Shown are only the non-dominated solutions in the final populations of 10 independent HypE runs after 1000 generations.}
+\begin{figure}
+	\centering
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/randVsCost_ns4_1_New_2000gens.eps}
+	\vspace{-1em}
+	\caption{\label{fig:randVsCostNew} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$).
+Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations on the uncorrelated\_4 instance.}
 \end{figure}
 
-\subsection{Comparison With An Exact Algorithm for the Two Linear Objectives}
-\TODO{add plots and describe what we compare (efficient cost/risk paths multiplied by the number of trucks); result: EA is better}
+\begin{figure}
+	\centering
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/initBoxPlots_OriginalCosts.eps}
+	\hfill
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/initBoxPlots_New.eps}
+	\vspace{-1em}
+	\caption{\label{fig:costVsRandBoxPlots} Comparison between the random initialization and the cost-optimal one.
+	Shown are boxplots of hypervolume values for HypE with population size 50 on the lazio\_4 instance (left) and on the uncorrelated\_4 instance (right) after both 250 and 2000 generations.}
+\end{figure}
 
@@
 
+\begin{figure}
+	\centering
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
+	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
+\end{figure}
+
 \begin{figure}%
 	\centering
-	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/diffPopsizes}%
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/diffPopsizes}%
+	\vspace{-1em}
 	\caption{\label{fig:popsizes} Nondominated solutions of 10 independent runs after 100,000 function evaluations for HypE with a population size of 50 ($\times$), 100 ($\circ$), and 200 ($\square$) respectively.}
@@
 algorithms show a reduced performance regarding the standard hypervolume value.
-However, W-HypE was run in order to show how the search can be steered towards regions of the objective space that have been specified by the user.
-In our case, we defined the weight of the underlying weighted hypervolume indicator \citep{zbt2007a} as a box with corners XXX such that the weighted hypervolume indicator only influences BLABLABLA}
+For NSGA-II, this is expected as its environmental selection schemes becomes less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}.
+This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing non-compliant quality measures such as NSGA-II.
+
+The reason for the reduced performance in the standard hypervolume indicator for W-Hype, however, is a different one.
+W-HypE was developed to steer the search towards regions of the objective space that have been specified by the user.
+To this end, it does not optimize the standard hypervolume indicator but a more generalized version of it, the weighted hypervolume indicator \citep{zbt2007a}.
+The idea behind W-HypE is that solutions in objective space regions with a higher weight will get a higher fitness due to the \emph{weighted} hypervolume they dominate.
+In order to be efficient also for a large number of objectives, W-HypE does not calculate the weighted hypervolume indicator values exactly (which is not possible for most of the weight functions) but instead samples them.
+Here, 10,000 samples are used in each iteration of the algorithm.
+
+To show the principles of W-HypE in the application of our hazmat routing problem, we defined the weight of W-HypE's underlying weighted hypervolume indicator as constant in a box with corners $[7\cdot 10^3, 3\cdot 10^7, 5\cdot 10^5]$ and $[10^4, 4\cdot 10^7, 1.1\cdot 10^6]$ and zero otherwise.
+Instead of optimizing the standard hypervolume indicator as HypE, W-HypE now optimizes this weighted hypervolume indicator, resulting in better, i.e., dominating solutions in regions in or close to the box.
+Figure~\ref{fig:whype} shows the resulting non-dominated solutions of both algorithms with population size 25 after 10 independent runs of 4,000 generations each as well as the sampling box.}
+
+This advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations, cp.\ Fig.~\ref{fig:whype}.
 
 \TODO{write}
+
+\TODO{W-HypE: number of non-dominated solutions in the sampling box: 14 for W-HypE, 6 for HypE; when overall non-dominated solutions are computed, W-HypE has 13, HypE only 1}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/WHypEVsHypE_ns4_1_OriginalCosts2.eps}
+	\vspace{-1em}
+	\caption{\label{fig:whype} Comparison between the standard HypE selection ($\circ$) with the one of W-HypE ($\times$) when the weighted hypervolume indicator is only sampled in the gray shaded box $[7\cdot 10^3, 10^4]\times[3\cdot 10^7, 4\cdot 10^7]\times[5\cdot 10^5, 1.1\cdot 10^6]$.
+Shown are all non-dominated solutions for each algorithm after 10 runs with 4,000 generations and population size 25.}
+\end{figure}
 

### 2011-11-30 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added text of W-HypE experiments

@@
 On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}.
-The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations) are shown in Fig.~\ref{fig:algoComparison}.
-\TODO{What can be seen is that both?
-algorithms show a reduced performance regarding the standard hypervolume value.
-For NSGA-II, this is expected as its environmental selection schemes becomes less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}.
-This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing non-compliant quality measures such as NSGA-II.
+The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations for NSGA-II and population size of 25 and 4,000 generations for W-HypE) are shown in Fig.~\ref{fig:algoComparison}.
+What can be seen is that both algorithms show a reduced performance regarding the standard hypervolume value with W-HypE being much worse than NSGA-II.
+For NSGA-II, this decrease  in hypervolume is expected as its environmental selection schemes does not optimize the hypervolume indicator as does HypE.
+Moreover, NSGA-II is known to become less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}.
+This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing quality measures that do not comply with the Pareto dominance relation such as the crwoding distance of NSGA-II.
 
-The reason for the reduced performance in the standard hypervolume indicator for W-Hype, however, is a different one.
-W-HypE was developed to steer the search towards regions of the objective space that have been specified by the user.
-To this end, it does not optimize the standard hypervolume indicator but a more generalized version of it, the weighted hypervolume indicator \citep{zbt2007a}.
+The reason for the reduced performance in the standard hypervolume indicator for W-HypE, on the other hand, can be explained in a different way.
+W-HypE and the underlying weighted hypervolume indicator was developed to steer the search towards regions of the objective space that have been specified by the user.
+To this end, instead of the standard hypervolume indicator, the weighted hypervolume indicator of \citep{zbt2007a} is optimized.
 The idea behind W-HypE is that solutions in objective space regions with a higher weight will get a higher fitness due to the \emph{weighted} hypervolume they dominate.
-In order to be efficient also for a large number of objectives, W-HypE does not calculate the weighted hypervolume indicator values exactly (which is not possible for most of the weight functions) but instead samples them.
+Furthermore, in order to be efficient also for a large number of objectives, W-HypE does not calculate the weighted hypervolume indicator values exactly but instead samples them.
 Here, 10,000 samples are used in each iteration of the algorithm.
+For details on the algorithm, we refer to the original publication \citep{bz2011a}.
 
 To show the principles of W-HypE in the application of our hazmat routing problem, we defined the weight of W-HypE's underlying weighted hypervolume indicator as constant in a box with corners $[7\cdot 10^3, 3\cdot 10^7, 5\cdot 10^5]$ and $[10^4, 4\cdot 10^7, 1.1\cdot 10^6]$ and zero otherwise.
-Instead of optimizing the standard hypervolume indicator as HypE, W-HypE now optimizes this weighted hypervolume indicator, resulting in better, i.e., dominating solutions in regions in or close to the box.
-Figure~\ref{fig:whype} shows the resulting non-dominated solutions of both algorithms with population size 25 after 10 independent runs of 4,000 generations each as well as the sampling box.}
-
-This advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations, cp.\ Fig.~\ref{fig:whype}.
-
-\TODO{write}
+Instead of optimizing the standard hypervolume indicator as HypE does, W-HypE now optimizes this weighted hypervolume indicator as a secondary selection criterion after employing a non-dominated sorting.
+As a result, it is finding better, i.e., dominating solutions in regions in or close to the box.
+Figure~\ref{fig:whype} shows the resulting non-dominated solutions of both algorithms W-HypE and HypE with population size 25 after 10 independent runs of 4,000 generations together with the sampling box.
+While HypE originally finds 6 non-dominated solutions within the box, W-HypE finds 14 within the same budget of function evaluations.
+When both algorithm outcomes are combined, it turns out that 13 of W-HypE's solutions are still non-dominated while 5 out of the 6 solutions found by HypE are dominated by W-HypE solutions.
+It has to be noted that the advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations (results not shown here).
 

### 2011-11-30 DAW_hazmatHypE/trunk/experiments/randVsCost/hypervolumes/hypervolumeAlgoComparisons.m

@@ -1,6 +1,6 @@
 addpath('../');
 
-hypervolumes = zeros(10, 5);
+hypervolumes = zeros(10, 6);
 
 % HypE popsize 25
 filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize25_randInit.4000';
@@ -17,11 +17,14 @@ hypervolumes(:,4) = -1 * dlmread(filename);
 % NSGA-II
 filename = 'HYP_ns4_1_OriginalCosts_nsga2_randInit.1000';
 hypervolumes(:,5) = -1 * dlmread(filename);
+% W-HypE
+filename = 'HYP_ns4_1_OriginalCosts_whype_box2_popsize25_randInit.4000';
+hypervolumes(:,6) = -1 * dlmread(filename);
 
 
 
 
-boxplot(hypervolumes, 'Labels', {'HypE_25'; 'HypE_50'; 'HypE_100'; 'HypE_200'; 'NSGA-II'});
+boxplot(hypervolumes, 'Labels', {'HypE_25'; 'HypE_50'; 'HypE_100'; 'HypE_200'; 'NSGA-II'; 'W-HypE'});
 
 
 

### 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
finished first version of "feasibility of the approach" section

@@
 On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}.
-The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations) are shown in Fig.~\ref{fig:algoComparison}.
-\TODO{What can be seen is that both?
-algorithms show a reduced performance regarding the standard hypervolume value.
-For NSGA-II, this is expected as its environmental selection schemes becomes less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}.
-This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing non-compliant quality measures such as NSGA-II.
+The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations for NSGA-II and population size of 25 and 4,000 generations for W-HypE) are shown in Fig.~\ref{fig:algoComparison}.
+What can be seen is that both algorithms show a reduced performance regarding the standard hypervolume value with W-HypE being much worse than NSGA-II.
+For NSGA-II, this decrease  in hypervolume is expected as its environmental selection schemes does not optimize the hypervolume indicator as does HypE.
+Moreover, NSGA-II is known to become less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}.
+This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing quality measures that do not comply with the Pareto dominance relation such as the crwoding distance of NSGA-II.
 
-The reason for the reduced performance in the standard hypervolume indicator for W-Hype, however, is a different one.
-W-HypE was developed to steer the search towards regions of the objective space that have been specified by the user.
-To this end, it does not optimize the standard hypervolume indicator but a more generalized version of it, the weighted hypervolume indicator \citep{zbt2007a}.
+The reason for the reduced performance in the standard hypervolume indicator for W-HypE, on the other hand, can be explained in a different way.
+W-HypE and the underlying weighted hypervolume indicator was developed to steer the search towards regions of the objective space that have been specified by the user.
+To this end, instead of the standard hypervolume indicator, the weighted hypervolume indicator of \citep{zbt2007a} is optimized.
 The idea behind W-HypE is that solutions in objective space regions with a higher weight will get a higher fitness due to the \emph{weighted} hypervolume they dominate.
-In order to be efficient also for a large number of objectives, W-HypE does not calculate the weighted hypervolume indicator values exactly (which is not possible for most of the weight functions) but instead samples them.
+Furthermore, in order to be efficient also for a large number of objectives, W-HypE does not calculate the weighted hypervolume indicator values exactly but instead samples them.
 Here, 10,000 samples are used in each iteration of the algorithm.
+For details on the algorithm, we refer to the original publication \citep{bz2011a}.
 
 To show the principles of W-HypE in the application of our hazmat routing problem, we defined the weight of W-HypE's underlying weighted hypervolume indicator as constant in a box with corners $[7\cdot 10^3, 3\cdot 10^7, 5\cdot 10^5]$ and $[10^4, 4\cdot 10^7, 1.1\cdot 10^6]$ and zero otherwise.
-Instead of optimizing the standard hypervolume indicator as HypE, W-HypE now optimizes this weighted hypervolume indicator, resulting in better, i.e., dominating solutions in regions in or close to the box.
-Figure~\ref{fig:whype} shows the resulting non-dominated solutions of both algorithms with population size 25 after 10 independent runs of 4,000 generations each as well as the sampling box.}
-
-This advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations, cp.\ Fig.~\ref{fig:whype}.
-
-\TODO{write}
-
-\TODO{W-HypE: number of non-dominated solutions in the sampling box: 14 for W-HypE, 6 for HypE; when overall non-dominated solutions are computed, W-HypE has 13, HypE only 1}
+Instead of optimizing the standard hypervolume indicator as HypE does, W-HypE now optimizes this weighted hypervolume indicator as a secondary selection criterion after employing a non-dominated sorting.
+As a result, it is finding better, i.e., dominating solutions in regions in or close to the box.
+Figure~\ref{fig:whype} shows the resulting non-dominated solutions of both algorithms W-HypE and HypE with population size 25 after 10 independent runs of 4,000 generations together with the sampling box.
+While HypE originally finds 6 non-dominated solutions within the box, W-HypE finds 14 within the same budget of function evaluations.
+When both algorithm outcomes are combined, it turns out that 13 of W-HypE's solutions are still non-dominated while 5 out of the 6 solutions found by HypE are dominated by W-HypE solutions.
+It has to be noted that the advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations (results not shown here).
 
@@
 Due to their heuristic nature, evolutionary multiobjective optimization algorithms like the one presented in this study cannot guarantee that the solutions found are actually Pareto-optimal.
-However, as we have seen above, the proposed algorithm is able to produce a set of non-dominated solutions in reasonable time which are close to Pareto-optimal solutions found by exact algorithms.
-Such sets of non-dominated solutions can be a starting point for a decision maker to investigate the trade-offs between the objectives and to decide in which objective space region interesting Pareto-optimal solutions are which can be explored further with more involved algorithms---be it exact algorithms or specialized evolutionary algorithms such as the mentioned W-HypE variant.
-In the above example instances, it turned out that total risk and cost are indeed more or less fully correlated in the original problem formulation or only loosely (with a larger trade-off between cost and total risk) when the individual arcs' risks and costs are.
+However, as we have seen above, the proposed algorithm is able to produce a set of non-dominated solutions in reasonable time.
+In particular the comparison with the cost-optimal solution shows that the randomized algorithm can find solutions close to the Pareto front.
+The found sets of non-dominated solutions can be a starting point for a decision maker to investigate the trade-offs between the objectives and to decide in which objective space region interesting Pareto-optimal solutions should be searched for with more involved algorithms---be it exact algorithms or specialized evolutionary algorithms such as the mentioned W-HypE variant.
 
-Regarding the two initialization schemes, it turned out that in combination with the (local and simple) variation operator, the cost-efficient initialization does not allow the algorithm to find additional solutions besides the initial one and that, in particular if the cost and total risk objectives are correlated, the random initialization is beneficial.
-Using other initializations as well as additional variation operators from the literature is easily possible with the proposed algorithm but their investigation is left for future research\footnote{For example, the
+Regarding the two initialization schemes, it turned out that in combination with the (local and quite simple) variation operator, the cost-optimal initialization does not allow the algorithm to find additional solutions besides the initial one.
+However, if the cost and total risk objectives are less correlated, this cost-optimal initialization has advantages over the random initialization in the early stage of the search when compared with respect to the hypervolume indicator.
+In general, however, we suggest to use the random initialization if the number of function evaluations is not a limiting factor.
 
-On the other hand, the proposed variation operator can be changed towards a larger neighborhood by allowing two or more nodes to be added or deleted from each truck path at a time---resulting in more exploration than exploitation when compared to the current variation operator.
-Also well-designed crossover operators are expected to be beneficial to the problem when, for example, good sub paths can be exchanged between solution paris.
-}.
+Using other initializations as well as additional variation operators from the literature is easily possible with the proposed algorithm but their investigation is left for future research.
+For example, one can think of a ``more random'' initialization where the initial truck paths are generated by iteratively applying the mutation operator on the empty path and also the multiple application of the described mutation operator within one single mutation might be beneficial as the explored space of neighbored paths is larger in comparison to the current mutation.
+Also well-designed crossover operators are expected to be beneficial to the problem when, for example, good sub paths can be exchanged between solution pairs.
 
-\begin{itemize}
-
-	\item approach is very general and can easily deal with slightly modified problems: e.g. different truck loads (only the objective function code needs to be changed)
-	\item it needs to be mentioned that for the tested instances with 311 nodes and 441 arcs, the variation operator of the evolutionary algorithm was able to construct the shortest paths from source to destination in some of the runs for some of the truck paths within the first 100,000 function evaluations.
-This corresponds to performing random walks of lengths around 20 nodes.
-\end{itemize}
+Overall, it is to say that the approach is very general and can easily be adapted if the problem formulationi is (slightly) changed as it is often the case in practice, e.g., if different numbers of trucks are available.
+Also in case of different truck loads, it is only the objective function code that needs to be changed---a big advantage of blackbox optimizers such as the proposed EMO algorithm.
 
@@
 Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary algorithm to cope with the non-linear equity risk objective.
+
+\TODO{ In the above example instances, it turned out that total risk and cost are indeed more or less fully correlated in the original problem formulation or only loosely (with a larger trade-off between cost and total risk) when the individual arcs' risks and costs are.
+}
 

### 2011-11-30 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
rewrote conclusions

@@
 The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest.
-Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary algorithm to cope with the non-linear equity risk objective.
+Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary multiobjective optimization (EMO) algorithm to cope with the non-linear equity risk objective.
 
-The third objective function of our problem can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$.
-Although this equivalent formulation makes the problem linear (with additional linear constraints), classical algorithms are expected to have difficulties with this formulation as well and our algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints.
-Note that, for the moment, the proposed EMO algorithm exists on paper only and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to generate solutions of sufficient quality and whether adaptively changing the number and capacity of trucks is beneficial.
+The algorithm, based on the state-of-the-art environmental selection scheme of HypE, employs a problem-specific variable-length representation and a mutation operator which are known to be well-suited for shortest path problems from theoretical studies.
+With its random initialization, the algorithm allows to find sets of non-dominated solutions in reasonable time which are close to the Pareto-optimal front.
+For the tested real-world instances with the road network of the Lazio region and dozens of trucks for up to 4 hazmat commodities, the current Java implementation allows to finish the 100,000 function evaluations of our experiments within less than 15 minutes on a standard Intel Core i5 laptop.
 
-In future work, one might want to look further into adding (simple) heuristics within the evoluionary algorithm in order to improve the algorithm.
+In future work, one might want to look further into adding (simple) heuristics within the EMO algorithm in order to improve the convergence of the algorithm.
 For example, one can observe that the solutions produced by the algorithm sometimes uses one and the same arc within a truck path, hence introducing cylces.
 With the three objectives cost, total and equity risk, such cycles can be easily prevented within the variation operator which might eventually result in a faster convergence of the algorithm to the Pareto front.
-
-% The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest.
-
-% In this study, we formalize transportation of hazmats problem as the minimization of three objectives and propose to use an evolutionary algorithm since many exact optimization procedures have difficulties with the non-linear equity risk objective.
-Note that our problem formulation can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$ \COMMENTD{this has to be adapted according to the new problem formulation}.
-Although this equivalent formulation makes the problem linear (with additional non-linear constraints), classical algorithms are expected to have difficulties with this formulation as well and the EMO algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints.
-% 
-% Note also that, for the moment, the proposed EMO algorithm only exists as a gedankenexperiment and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to improve the quality of the found solutions to a sufficient level.
 

### 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
wrote about cost vs. random initialization

@@
 While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bcg2009a} and Table~\ref{tab:instances}.
-Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but with randomly permutated costs where each occuring cost value has been divided by a random value between 0 and 5000.
+Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but where the costs of each arc are the reciprocal of the risk of the first commodity.
+Figure~\ref{fig:correlation} shows those differences between the costs and risks for the instances lazio\_4\_1 and uncorrelated\_4.
 
@@
 \end{table}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.48\columnwidth]{figs/correlationLazio41.eps}
+	\hfill
+	\includegraphics[width=0.48\columnwidth]{figs/correlationUncorrelated4.eps}
+	\caption{\label{fig:correlation} Correlation between the costs and the risks of the first commodity for both the lazio\_4\_1 instance (left) and the uncorrelated\_4 instance (right).
+A single point corresponds to a single arc.}
+\end{figure}
 
@@
 
+\subsection{Convergence Towards the Pareto Front}
+Figure~\ref{fig:hypervolumeOverTime} shows the boxplots of the hypervolume values for HypE with population size 50 and 100 for different number of generations.
+What can be seen is that, in general, the hypervolume indicator is mazimized over time and that for the algorithm with population size 50, the hypervolume value starts to stagnate starting from around generation 1500 with the mean still going up but the variance staying high.
+The fact that all runs found a similar hypervolume value which is significantly higher than the one of the random population in the beginning (around $9.62\cdot 10^26$ for both algorithms) indicates that the algorithm found solution sets which are close to the Pareto front.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}
+	\hfill
+	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}
+	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
+\end{figure}
+
 \subsection{Random Initialization Versus Cost-Optimal Initialization}
@@
 Surprisingly, the cost-optimal initialization could not produce other solutions than the cost-optimal one itself.
-\TODO{However, what can be seen is that the ...}
+As we will see below, the reason is a high correlation between the risk and costs of each arc, which seemingly result in the fact that no non-dominated solution can be reached from the initial solution by the applied mutation operator.
+However, the comparison between the both initialization schemes on the lazio\_4\_1 instance shows that also the algorithm employing the random initialization can find solutions which are close to the Pareto front---improving over the hypervolume of the single cost-optimal solution in later generations, see the left-hand plot of Fig.~\ref{fig:costVsRandBoxPlots}.
 
-Boxplots can be found in Fig.~\ref{fig:costVsRandBoxPlots} ...
+To show that the cost-optimal initialization is not useless, we run the algorithms again on the instance uncorrelated\_4 where the cost and risk values are fully uncorrelated.
+In this case, there are more local changes of the truck paths that result in non-dominated solutions and eventually in a larger variaty of solutions found by the algorithm employing the cost-optimal initialization (see Fig.~\ref{fig:randVsCostNew}).
+Due to the fact that the cost-optimal initialization starts already on the Pareto-front, it also results in higher hypervolume values than for the random initialization variant---both for a low and a high number of generations, see the right-hand side plot of Fig.~\ref{fig:costVsRandBoxPlots}.
 
@@
 \subsection{Changing the Population Size}
-In a second experiment, we show that changing the population size of the algorithm in standard ranges does not change the algorithm outcome drastically.
+In the next experiment, we aim at showing that changing the population size of the algorithm in standard ranges does not change the algorithm outcome drastically.
 To this end, Fig.~\ref{fig:popsizes} shows the nondominated solutions of 10 independent runs after 100,000 function evaluations for three different versions of HypE where the population size changes from 50 over 100 to 200 on the largest instance lazio\_4\_1 with 4 commodities.
@@
 Ten runs with 2000 iterations and population size 200, for example, need already around 20 hours to complete on an Intel Core2 Duo with 2.8GHz.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
-	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
-\end{figure}
 
@@
 The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest.
-Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary algorithm to cope with the non-linear equity risk objective.
+Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary multiobjective optimization (EMO) algorithm to cope with the non-linear equity risk objective.
 
-\TODO{ In the above example instances, it turned out that total risk and cost are indeed more or less fully correlated in the original problem formulation or only loosely (with a larger trade-off between cost and total risk) when the individual arcs' risks and costs are.
-}
+The algorithm, based on the state-of-the-art environmental selection scheme of HypE, employs a problem-specific variable-length representation and a mutation operator which are known to be well-suited for shortest path problems from theoretical studies.
+With its random initialization, the algorithm allows to find sets of non-dominated solutions in reasonable time which are close to the Pareto-optimal front.
+For the tested real-world instances with the road network of the Lazio region and dozens of trucks for up to 4 hazmat commodities, the current Java implementation allows to finish the 100,000 function evaluations of our experiments within less than 15 minutes on a standard Intel Core i5 laptop.
 
-The third objective function of our problem can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$.
-Although this equivalent formulation makes the problem linear (with additional linear constraints), classical algorithms are expected to have difficulties with this formulation as well and our algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints.
-Note that, for the moment, the proposed EMO algorithm exists on paper only and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to generate solutions of sufficient quality and whether adaptively changing the number and capacity of trucks is beneficial.
-
-In future work, one might want to look further into adding (simple) heuristics within the evoluionary algorithm in order to improve the algorithm.
+In future work, one might want to look further into adding (simple) heuristics within the EMO algorithm in order to improve the convergence of the algorithm.
 For example, one can observe that the solutions produced by the algorithm sometimes uses one and the same arc within a truck path, hence introducing cylces.
 With the three objectives cost, total and equity risk, such cycles can be easily prevented within the variation operator which might eventually result in a faster convergence of the algorithm to the Pareto front.
-
-% The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest.
-
-% In this study, we formalize transportation of hazmats problem as the minimization of three objectives and propose to use an evolutionary algorithm since many exact optimization procedures have difficulties with the non-linear equity risk objective.
-Note that our problem formulation can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$ \COMMENTD{this has to be adapted according to the new problem formulation}.
-Although this equivalent formulation makes the problem linear (with additional non-linear constraints), classical algorithms are expected to have difficulties with this formulation as well and the EMO algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints.
-% 
-% Note also that, for the moment, the proposed EMO algorithm only exists as a gedankenexperiment and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to improve the quality of the found solutions to a sufficient level.
 

### 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
rewrote intro (including some new references that have to be put into the BibteX file)

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \cite{cd2008a}.
+The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \citep{etv2007a,cd2008a}.
 The main target of this problem is to select routes from a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized---without producing excessive economic costs.
-%
-When solving such a problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions are not affected.
+
+More specifically, one can distinguish between two main problems of different scale \citep{etv2007a}.
+The \emph{local} hazmat routing problem aims at finding the route for a single shipment from a source to a destination node in a given network; the \emph{global} hazmat routing problem, on the other hand, looks as the macro level of shipping several hazmat types simultaneously from multiple sources to multiple destinations---with much less studies on the latter \citep{etc2007a}.
+When solving such a global hazmat routing problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions might not be affected at all.
 In this case, one may wish to distribute the risk in an equitable way over the population and the environment.
-Several studies consider this additional minimization of the equity risk, but most of them consist of a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{aeb2000a, cd2008a}).
-% 
-% 
-% and can be classified into the \textit{resolution-equity-based methods} and the \textit{model-equity-based methods}.
-In resolution-equity-based methods, the equity constraints are taken into account only indirectly via a dissimilarity index between paths which is to be maximized \cite{JOH92, LOM93, KUB97, aeb2000a}.
-Model-equity-based methods, on the other hand, formalize the risk equity as constraints directly within the model.
-In \cite{GOP90a, GOP90b}, for example, the authors propose an equity shortest path model that minimizes the total risk of travel while the difference between the risks imposed on any two arbitrary zones does not exceed a given threshold.
-%
-A more realistic \emph{multi}-commodity flow model was proposed in \citep{cd2008a} where each commodity is considered as one hazmat type.
-The objective function is formulated as the sum of the economical cost and the cost related to the consequences of an incident for each material.
-To deal with risk equity, the costs are defined as functions of the flow traversing the arcs which imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc.
+Although several studies consider this additional minimization of the equity risk, most of them restrict the model to a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{aeb2000a}).
+More realistic \emph{multi}-commodity flow model have been proposed where each commodity is considered as one hazmat type \citep{cd2008a} but the majority of all hazmat routing studies do not consider the multi-objective nature of the hazmat routing problem but optimize only the risk \cite{gkbk1990a} or an aggregation or scalarization of the objective functions instead \citep{cd2008a}.
+An exception is \citep{dgs2005a} where the focus lies on finding a set of dissimilar paths that are Pareto-optimal with respect to cost and risk instead of solving the multiobjective hazmat routing problem directly.
+Moreover, only one origin-destination pair is considered.
 
-The majority of all hazmat routing studies deal with a single-objective scenario although the problem itself is multiobjective in nature and it is important to study the trade-offs among the objectives.
-Evolutionary Multiobjective Optimization (EMO) algorithms are able to compute a set of solutions showing these trade-offs within a single algorithm run which is the reason why we propose to use them for the problem of hazmat routing in this study (Sec.~\ref{S_MEO}).
-Before, we formalize the routing of hazmat problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).
+As it is important for a decision maker to study the trade-offs among the objectives, we treat the global hazmat routing problem of finding a set of routes minimizing total costs, total risk, and equity risk, from a multiobjective perspective.
+Evolutionary Multiobjective Optimization (EMO) algorithms have been shown to be able to compute a set of solutions showing the trade-offs between multiple objectives within a single algorithm run in particular when other (exact) algorithms have difficulties to approach the Pareto front, e.g, if the front is too large.
+As general black-box algorithms, they also allow to cope with changes in the problem formulation which occurs frequently in real-world applications.
 
-% see for example \cite{cd2008a, cgr2007, gkbk1990, za2004}, but all problem formulations include the equity risk as constraints and not as an additional objective.
-\citet{gian1998} on the other hand considers the equity risk as separate objective but uses a weighted sum to aggregate his four objectives into a single-objective problem.
-The main reason why, to the best of our knowledge, no study considered a fully multiobjective model with the equity risk as objective might be the resulting non-linearity of the problem which make classical optimization algorithms such as branch-and-cut \TODON{equity is an objective in \cite{{gian1998}} ...
-We solve the broblem with a branch and price algorithm because the model is a multicommodity flow model, and this model is the more adapted one for this problem because it permit to manage simultaneously many hazmat types.
-The min-max formulation of the objective is easily linearized.
-I give you the major axes of the work: (1) Solve the presented problem (optimize cost, risk and equity) (2) what are models and resolution methods proposed in the litterature (3) what is the model retained and why (in our case, the multicommodity model) (4) how to handle equity with this model ?
-(5) how to solve it ?
-(branch and price is adapted for these models, but can not hadle many objectives $->$ evolutionnary algorithms) } infeasible and led us to our choice of proposing an evolutionary multiobjective algorithm as heuristic here.
-
-%The computation of routes with a fairly distributed risk consists in generating dissimilar origin-destination paths, i.e paths which relatively don't impact the same zones. Solution approaches
-
-%Section~\ref{S_RECRP} defines the problem of transporting hazardous materials with equity risk as a three-objective optimization problem with the minimization of the total routing cost, of the total routing risk {\em and} of the risk equity as the objectives---the latter broadly defined as the risk shared by a set of regions that compose the geographical area under consideration.
-Based on this formulation, Sec.~\ref{S_MEO} presents an evolutionary multiobjective optimization algorithm before Sec.~\ref{S_FW} concludes the abstract.
-
-%In \cite{cd2008a} was proposed a multi-commodity flow model for this problem, where each commodity is considered as one hazmat type.
-To deal with risk equity, the costs are defined as functions of the flow traversing the arcs, this imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc.
-The major drawback of this model is that it can lead to the saturation of the arcs of shortest paths and, therefore, does not guarantee the equity target.
-
-%In \cite{BEL10}, the authors consider a similar model but with additional equity constraints and objectives.
-The objective is the minimization of the maximum risk equity imposed on the population and the environment and the model is solved using a pranch-and-price algorithm.
-
-%As hazmat routing is multiobjective in nature, we introduce in this work a multiobjective version of this problem.
-Our goal is the minimization of the total routing cost, of the total routing risk {\em and} of the risk equity, the latter broadly defined as the risk shared by a set of regions that compose the geographical area under consideration.
-
-%This paper is organized as follows.
-We describe the problem in section \ref{S_RECRP}, we give its mathematical formulation in section \ref{S_AOM}, in section \ref{S_MEO} we show how multiobjective evolutionary methods can be applied to this problem and in section \ref{S_FW} we exopse our further work.
+In this study, we therefore propose an EMO algorithm for the global hazmat routing problem (Sec.~\ref{S_MEO}).
+Before, we formalize the specific multiobjective risk-equity constrained routing problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).
+Section~\ref{sec:experiments} is then devoted to the experimental analysis of this algorithm and its application to realistic instances in the Lazio region of Italy before Sec.~\ref{S_FW} concludes the paper.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Experimental analysis}
+\section{Experimental analysis}\label{sec:experiments}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Overall, it is to say that the approach is very general and can easily be adapted if the problem formulationi is (slightly) changed as it is often the case in practice, e.g., if different numbers of trucks are available.
-Also in case of different truck loads, it is only the objective function code that needs to be changed---a big advantage of blackbox optimizers such as the proposed EMO algorithm.
+Overall, it is to say that the approach is very general and can easily be adapted if the problem formulation is (slightly) changed as it is often the case in practice, e.g., if different numbers of trucks or different types of vehicles with different loads are available.
+Also in the case when the definition of total and equity risk changes, it is only the objective function code that needs to be changed---a big advantage of blackbox optimizers such as for the proposed EMO algorithm.
 
@@
 With the three objectives cost, total and equity risk, such cycles can be easily prevented within the variation operator which might eventually result in a faster convergence of the algorithm to the Pareto front.
+Another path for future research is the consideration of more realistic problem instances.
+Stochastic risks or dynamically changing instances (for example due to road closures or traffic jams) would be interesting to address---challenges that have been dealt with by EMO algorithms in other application domains already before \cite{deb2001,cvl2007a}.
 

### 2011-11-30 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added missing BibTeX entries

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \cite{cd2008a}.
+The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \citep{etv2007a,cd2008a}.
 The main target of this problem is to select routes from a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized---without producing excessive economic costs.
-%
-When solving such a problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions are not affected.
+
+More specifically, one can distinguish between two main problems of different scale \citep{etv2007a}.
+The \emph{local} hazmat routing problem aims at finding the route for a single shipment from a source to a destination node in a given network; the \emph{global} hazmat routing problem, on the other hand, looks as the macro level of shipping several hazmat types simultaneously from multiple sources to multiple destinations---with much less studies on the latter \citep{etv2007a}.
+When solving such a global hazmat routing problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions might not be affected at all.
 In this case, one may wish to distribute the risk in an equitable way over the population and the environment.
-Several studies consider this additional minimization of the equity risk, but most of them consist of a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{aeb2000a, cd2008a}).
-% 
-% 
-% and can be classified into the \textit{resolution-equity-based methods} and the \textit{model-equity-based methods}.
-In resolution-equity-based methods, the equity constraints are taken into account only indirectly via a dissimilarity index between paths which is to be maximized \cite{JOH92, LOM93, KUB97, aeb2000a}.
-Model-equity-based methods, on the other hand, formalize the risk equity as constraints directly within the model.
-In \cite{GOP90a, GOP90b}, for example, the authors propose an equity shortest path model that minimizes the total risk of travel while the difference between the risks imposed on any two arbitrary zones does not exceed a given threshold.
-%
-A more realistic \emph{multi}-commodity flow model was proposed in \citep{cd2008a} where each commodity is considered as one hazmat type.
-The objective function is formulated as the sum of the economical cost and the cost related to the consequences of an incident for each material.
-To deal with risk equity, the costs are defined as functions of the flow traversing the arcs which imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc.
+Although several studies consider this additional minimization of the equity risk, most of them restrict the model to a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{aeb2000a}).
+More realistic \emph{multi}-commodity flow model have been proposed where each commodity is considered as one hazmat type \citep{cd2008a} but the majority of all hazmat routing studies do not consider the multi-objective nature of the hazmat routing problem but optimize only the risk \cite{gkbk1990a} or an aggregation or scalarization of the objective functions instead \citep{cd2008a}.
+An exception is \citep{dgs2005a} where the focus lies on finding a set of dissimilar paths that are Pareto-optimal with respect to cost and risk instead of solving the multiobjective hazmat routing problem directly.
+Moreover, only one origin-destination pair is considered.
 
-The majority of all hazmat routing studies deal with a single-objective scenario although the problem itself is multiobjective in nature and it is important to study the trade-offs among the objectives.
-Evolutionary Multiobjective Optimization (EMO) algorithms are able to compute a set of solutions showing these trade-offs within a single algorithm run which is the reason why we propose to use them for the problem of hazmat routing in this study (Sec.~\ref{S_MEO}).
-Before, we formalize the routing of hazmat problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).
+As it is important for a decision maker to study the trade-offs among the objectives, we treat the global hazmat routing problem of finding a set of routes minimizing total costs, total risk, and equity risk, from a multiobjective perspective.
+Evolutionary Multiobjective Optimization (EMO) algorithms have been shown to be able to compute a set of solutions showing the trade-offs between multiple objectives within a single algorithm run in particular when other (exact) algorithms have difficulties to approach the Pareto front, e.g, if the front is too large.
+As general black-box algorithms, they also allow to cope with changes in the problem formulation which occurs frequently in real-world applications.
 
-% see for example \cite{cd2008a, cgr2007, gkbk1990, za2004}, but all problem formulations include the equity risk as constraints and not as an additional objective.
-\citet{gian1998} on the other hand considers the equity risk as separate objective but uses a weighted sum to aggregate his four objectives into a single-objective problem.
-The main reason why, to the best of our knowledge, no study considered a fully multiobjective model with the equity risk as objective might be the resulting non-linearity of the problem which make classical optimization algorithms such as branch-and-cut \TODON{equity is an objective in \cite{{gian1998}} ...
-We solve the broblem with a branch and price algorithm because the model is a multicommodity flow model, and this model is the more adapted one for this problem because it permit to manage simultaneously many hazmat types.
-The min-max formulation of the objective is easily linearized.
-I give you the major axes of the work: (1) Solve the presented problem (optimize cost, risk and equity) (2) what are models and resolution methods proposed in the litterature (3) what is the model retained and why (in our case, the multicommodity model) (4) how to handle equity with this model ?
-(5) how to solve it ?
-(branch and price is adapted for these models, but can not hadle many objectives $->$ evolutionnary algorithms) } infeasible and led us to our choice of proposing an evolutionary multiobjective algorithm as heuristic here.
-
-%The computation of routes with a fairly distributed risk consists in generating dissimilar origin-destination paths, i.e paths which relatively don't impact the same zones. Solution approaches
-
-%Section~\ref{S_RECRP} defines the problem of transporting hazardous materials with equity risk as a three-objective optimization problem with the minimization of the total routing cost, of the total routing risk {\em and} of the risk equity as the objectives---the latter broadly defined as the risk shared by a set of regions that compose the geographical area under consideration.
-Based on this formulation, Sec.~\ref{S_MEO} presents an evolutionary multiobjective optimization algorithm before Sec.~\ref{S_FW} concludes the abstract.
-
-%In \cite{cd2008a} was proposed a multi-commodity flow model for this problem, where each commodity is considered as one hazmat type.
-To deal with risk equity, the costs are defined as functions of the flow traversing the arcs, this imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc.
-The major drawback of this model is that it can lead to the saturation of the arcs of shortest paths and, therefore, does not guarantee the equity target.
-
-%In \cite{BEL10}, the authors consider a similar model but with additional equity constraints and objectives.
-The objective is the minimization of the maximum risk equity imposed on the population and the environment and the model is solved using a pranch-and-price algorithm.
-
-%As hazmat routing is multiobjective in nature, we introduce in this work a multiobjective version of this problem.
-Our goal is the minimization of the total routing cost, of the total routing risk {\em and} of the risk equity, the latter broadly defined as the risk shared by a set of regions that compose the geographical area under consideration.
-
-%This paper is organized as follows.
-We describe the problem in section \ref{S_RECRP}, we give its mathematical formulation in section \ref{S_AOM}, in section \ref{S_MEO} we show how multiobjective evolutionary methods can be applied to this problem and in section \ref{S_FW} we exopse our further work.
+In this study, we therefore propose an EMO algorithm for the global hazmat routing problem (Sec.~\ref{S_MEO}).
+Before, we formalize the specific multiobjective risk-equity constrained routing problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).
+Section~\ref{sec:experiments} is then devoted to the experimental analysis of this algorithm and its application to realistic instances in the Lazio region of Italy before Sec.~\ref{S_FW} concludes the paper.
 
@@
 With the three objectives cost, total and equity risk, such cycles can be easily prevented within the variation operator which might eventually result in a faster convergence of the algorithm to the Pareto front.
+Another path for future research is the consideration of more realistic problem instances.
+Stochastic risks or dynamically changing instances (for example due to road closures or traffic jams) would be interesting to address---challenges that have been dealt with by EMO algorithms in other application domains already before \cite{deb2001a,cvl2007a}.
 

### 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
final revision of manuscript including spell-check

@@
 \begin{abstract}
-Nowadays, various hazardous materials, or \emph{hazmats}, are transported on our streets, via trains, planes or ships.
-In paricular when routing them over the street network, where typically several possible routes exist, one would like to minimize both the costs and the total risks.
+Nowadays, various hazardous materials, or \emph{hazmats}, are transported on our streets, via trains, planes or by ships.
+In particular when routing them over the street network, where typically several possible discrete routes exist, one would like to minimize both the costs and the total risks.
 However, sometimes one aims at minimizing an additional type of risk, which in most optimization models is resulting in a non-linear objective: the equity risk, or the \emph{maximum} risk over all regions, streets, etc.
 In other words, one aims at distributing the overall risks as much as possible while the costs and the total risk should be still minimized simultaneously.
-In this paper, this three-objective optimization problem is tackled with a novel randomized algorithm, based on a state-of-the-art Evolutionary Multiobjective Optimization algorithm and taylored to the needs of finding Pareto-optimal truck routes in a given network by proposing a variable-length representation and problem-specific operators.
+In this paper, this three-objective optimization problem is tackled with a novel randomized algorithm, based on a state-of-the-art Evolutionary Multiobjective Optimization algorithm and tailored to the needs of finding Pareto-optimal truck routes in a given network by proposing a variable-length representation and problem-specific operators.
 Experimental results on realistic instances for the street network of Italy's Lazio region show that the proposed approach efficiently finds a variety of good solutions in reasonable time.
@@
 The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \citep{etv2007a,cd2008a}.
-The main target of this problem is to select routes from a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized---without producing excessive economic costs.
+The main target of this problem is to select routes for a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized when the hazmat is transported over the selected routes---without producing excessive economic costs.
 
-More specifically, one can distinguish between two main problems of different scale \citep{etv2007a}.
-The \emph{local} hazmat routing problem aims at finding the route for a single shipment from a source to a destination node in a given network; the \emph{global} hazmat routing problem, on the other hand, looks as the macro level of shipping several hazmat types simultaneously from multiple sources to multiple destinations---with much less studies on the latter \citep{etc2007a}.
-When solving such a global hazmat routing problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions might not be affected at all.
+More specifically, one can distinguish between two main problem types of different scale \citep{etv2007a}.
+The \emph{local} hazmat routing problem aims at finding the route for a single shipment from a source to a destination node in a given network; the \emph{global} hazmat routing problem, on the other hand, looks at the macro level of shipping several hazmat types simultaneously from multiple sources to multiple destinations---with much less studies on the latter \citep{etv2007a}.
+
+When solving a global hazmat routing problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions might not be affected at all.
 In this case, one may wish to distribute the risk in an equitable way over the population and the environment.
@@
 More realistic \emph{multi}-commodity flow model have been proposed where each commodity is considered as one hazmat type \citep{cd2008a} but the majority of all hazmat routing studies do not consider the multi-objective nature of the hazmat routing problem but optimize only the risk \cite{gkbk1990a} or an aggregation or scalarization of the objective functions instead \citep{cd2008a}.
-An exception is \citep{dgs2005a} where the focus lies on finding a set of dissimilar paths that are Pareto-optimal with respect to cost and risk instead of solving the multiobjective hazmat routing problem directly.
-Moreover, only one origin-destination pair is considered.
+An exception is \citep{dgs2005a} where the focus lies on finding a set of dissimilar paths that are Pareto-optimal with respect to cost and risk but the multiobjective hazmat routing problem is only solved indirectly.
+In addition, also only one origin-destination pair is considered in \citep{dgs2005a}.
 
-As it is important for a decision maker to study the trade-offs among the objectives, we treat the global hazmat routing problem of finding a set of routes minimizing total costs, total risk, and equity risk, from a multiobjective perspective.
-Evolutionary Multiobjective Optimization (EMO) algorithms have been shown to be able to compute a set of solutions showing the trade-offs between multiple objectives within a single algorithm run in particular when other (exact) algorithms have difficulties to approach the Pareto front, e.g, if the front is too large.
+As it is important for a decision maker to study the trade-offs among the objectives, we treat the global hazmat routing problem of finding a set of routes minimizing total costs, total risk, and equity risk, from a true multiobjective perspective.
+Evolutionary Multiobjective Optimization (EMO) algorithms have been shown to be able to compute a set of solutions showing those trade-offs between multiple objectives within a single algorithm run in particular when other (exact) algorithms have difficulties to approach the Pareto front, e.g, if the front is too large \citep{deb2001a,cvl2007a}.
 As general black-box algorithms, they also allow to cope with changes in the problem formulation which occurs frequently in real-world applications.
@@
 In this study, we therefore propose an EMO algorithm for the global hazmat routing problem (Sec.~\ref{S_MEO}).
-Before, we formalize the specific multiobjective risk-equity constrained routing problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).
+Before, we formalize the specific multiobjective risk-equity constrained routing problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) based on the multicommodity flow model of \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).
 Section~\ref{sec:experiments} is then devoted to the experimental analysis of this algorithm and its application to realistic instances in the Lazio region of Italy before Sec.~\ref{S_FW} concludes the paper.
@@
 We assume that the risk is computed on each arc of the network and is proportional to the number of trucks traversing such an arc.
-Although such a proportionality is not always given in pracice, our assumption is a good approximation if the trucks are of the same type and/or the load is very similar as it is the case for standardized castor containers for radioactive material\footnote{Note that changing the proposed model towards different types of trucks with different types of loads and therefore different resulting risks is straightforward and is expected to cause no additional difficulties for the proposed evolutionary algorithm.}.
+Although such a proportionality is not always given in practice, our assumption is a good approximation if the trucks are of the same type and/or the load is very similar as it is the case for standardized castor containers for radioactive material\footnote{Note that changing the proposed model towards different types of trucks with different types of loads and therefore different resulting risks is straightforward and is expected to cause no additional difficulties for the proposed EMO algorithm.}.
 With each arc $(i,j) \in A$, we also associate a cost $c_{ij}$, induced by the travel of a truck on this arc.
@@
 
-The first objective in (\ref{eq:obj1}) is a cost function and is to be minimized.
+The first objective in (\ref{eq:obj1}) is the cost function and is to be minimized.
 The second objective, given by (\ref{eq:obj2}), minimizes the total risk over all arcs and objective (\ref{eq:obj3}) minimizes the maximum risk imposed on all arcs.
 The constraints in (\ref{eq:flow}) are conservation constraints, where $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$ are the direct successors and predecessors of node $i$, and $q_i^c=V^c$ if  $i=s^c$, $q_i^c=-V^c$ if  $i=t^c$ and $q_i^c=0$ otherwise.
-Note that while the first two objectives are linear in the variables, the maximum in the third objective makes it highly non-linear---the main reason why we propose to use an evolutionary algorithm to deal with the above hazmat routing problem.
+Note that while the first two objectives are linear in the variables, the maximum in the third objective makes it highly non-linear---another reason to use an EMO algorithm to deal with the above hazmat routing problem.
 
@@
 We choose a variable length representation as it has been theoretically shown to be a good choice for multiobjective shortest paths problems \citep{horo2009a}: A solution is thereby represented by a list of paths of variable lengths with one path per truck.
-We decided to consider a fixed amount of trucks for each commodity and therefore a fixed number of paths through the network although, in principle, the number of trucks could change during the algorithm run if an appropriate mutation operator is designed.
+We decided to consider a fixed amount of trucks for each commodity and therefore a fixed number of paths through the network, although, in principle, the number of trucks could change during the algorithm run if an appropriate mutation operator is designed.
 The left hand side of Fig.~\ref{fig:paths} shows an exemplary solution with two trucks for commodity 1 and one truck for commodity 2.
@@
 Note that this initialization scheme, though random, does not produce fully random paths.
-Instead, the initialization results in truck paths which are constructed by concatenating two shortest paths---the shortest path from the source node to the one randomly selected and the shortest path from the random node to the destination, which is added due to the representation described above.
+Instead, the initialization creates truck paths which are constructed by concatenating two shortest paths---the shortest path from the source node to the one randomly selected and the shortest path from the random node to the destination, which is added due to the representation described above.
 
@@
 Nevertheless, since no other solution can have a smaller cost, the initial solution is already (weakly) Pareto-optimal and is expected to be a good starting point for the algorithm.
-In the following experiments, we will see when each of the two initialization schemes is favourable.
+In the following experiments, we will see when each of the two initialization schemes is favorable.
 
@@
 
-After the initialization of the population with $\alpha$ individuals by the variator, their evaluation and sending the objective values to the selector module, a standard PISA run always continues in the same way.
+After the initialization of the population with $\alpha$ solutions by the variator, the new solutions are evaluated and their objective vectors send to the selector module.
+Afterwards, a PISA run always continues in the same way.
 The selector decides which $\mu$ solutions to variate and the variator uses those $\mu$ solutions as starting points for the generation of $\lambda$ new search points which are again evaluated and send to the selector.
 In addition, the selector can decide whether solutions are stored in an archive or whether they are deleted.
-In the case of the selectors used in this study, no archive is maintained.
-The search stops after a predefined number of variator/selector iterations, also called generations, are performed.
+In the case of the selectors, used in this study, no archive is maintained.
+The search stops after a predefined number of variator/selector iterations, also called generations.
 In our case, $\alpha=\mu=\lambda$, which will be also called population size hereafter.
@@
 The main purpose of an EMO algorithm's environmental selection is to ensure convergence of the population towards the Pareto front as well as a good spread of it along the front \citep{deb2001a}.
-HypE persues these two goals by a two-step approach based on the hypervolume indicator \citep[see][]{ztlf2003a}.
+HypE pursues these two goals by a two-step approach based on the hypervolume indicator (see \citep{ztlf2003a} for a definition of the hypervolume indicator).
 
@@
 The solutions that are not dominated in the entire set get a rank of 1, the solutions of rank 2 are those which are non-dominated among the remaining solutions, and so forth.
-As long as the number of solutions is not exceeding $\mu$, entire fronts are chosen for the next iteration starting from the front with rank 1.
+As long as the number of solutions is not exceeding $\mu$, entire fronts are chosen to be introduced in the next generation's population; starting from the front with rank 1.
 For the first front for which the overall number of solutions exceeds $\mu$, a second criterion is used to determine the solutions to be kept for the next iteration.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-In the remainder of this paper, we investigate how the above proposed evolutionary multiobjective optimization algorithm can cope with realistic instances of the multiobjective risk-equity constrained routing problem.
+In the remainder of this paper, we investigate how the above proposed EMO algorithm can cope with realistic instances of the multiobjective risk-equity constrained routing problem.
 The experimental analyses have two main goals.
 On the one hand, we show that the algorithm is efficient and can find multiple non-dominated solutions on practical examples in reasonable time.
-On the other hand, we investigate which of the two proposed initialization schemes is favourable and when this is the case.
+On the other hand, we investigate which of the two proposed initialization schemes is favorable and when this is the case.
 In addition, a few more experiments are conducted to show that the standard parameters of the algorithm behave as with other test problems.
@@
 
-As a test case for all the experiments, we chose to use a real-world case study based on the work of \citet{bcg2009a}.
+As a test case for all the experiments, we choose to use a real-world case study based on the work of \citet{bcg2009a}.
 The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome.
@@
 Whereas the number of trucks for lazio\_3 and lazio\_4\_1 have been chosen randomly, lazio\_2 has the same number of trucks than the lazio\_3 instance for the first two commodities.
-The lazio\_4\_2 instance has only 1 truck more for commodity 4 than lazio\_4\_1 and overall three times the number of trucks for each commodity than the instance lazio\_4\_3.
+The lazio\_4\_2 instance has only 1 truck more for commodity 4 than lazio\_4\_1 and overall three times the number of trucks for each commodity than the instance lazio\_4\_3, cp.\ Table~\ref{tab:instances}.
 
@@
 However and thirdly, this relationship is not linear in the plots of Fig.~\ref{fig:compareAll} as can be seen, for example, between the lazio\_4\_2 and the lazio\_4\_3 instance.
-Though the number of trucks per commodity differs by a factor of 3 between them, the objective function values of the resulting non-dominated solutions after 1,000 generations are slightly further away.
-The reason for that is the much smaller search space for the lazio\_4\_3 instance which can be searched more exhaustively by the algorithm with the same number of function evalutions---resulting in a better convergence towards the Pareto front.
+Though the number of trucks per commodity differs by a factor of 3 between them, the objective function values of the resulting non-dominated solutions after 2,000 generations are slightly further away.
+The reason for that is the much smaller search space for the lazio\_4\_3 instance which can be searched more exhaustively by the algorithm with the same number of function evaluations---resulting in a better convergence towards the Pareto front.
 
@@
 Figure~\ref{fig:hypervolumeOverTime} shows the boxplots of the hypervolume values for HypE with population size 50 and 100 for different number of generations.
-What can be seen is that, in general, the hypervolume indicator is mazimized over time and that for the algorithm with population size 50, the hypervolume value starts to stagnate starting from around generation 1500 with the mean still going up but the variance staying high.
-The fact that all runs found a similar hypervolume value which is significantly higher than the one of the random population in the beginning (around $9.62\cdot 10^26$ for both algorithms) indicates that the algorithm found solution sets which are close to the Pareto front.
+What can be seen is that, in general, the hypervolume indicator is maximized over time and that for the algorithm with population size 50, the hypervolume value starts to stagnate starting from around generation 1500 with the mean still going up but the variance staying high.
+The fact that all runs found a similar hypervolume value which is significantly higher than the one of the random population in the beginning (around $9.62\cdot 10^{26}$ for both algorithms) indicates that the algorithm found solution sets which are close to the Pareto front.
 
@@
 Surprisingly, the cost-optimal initialization could not produce other solutions than the cost-optimal one itself.
-As we will see below, the reason is a high correlation between the risk and costs of each arc, which seemingly result in the fact that no non-dominated solution can be reached from the initial solution by the applied mutation operator.
-However, the comparison between the both initialization schemes on the lazio\_4\_1 instance shows that also the algorithm employing the random initialization can find solutions which are close to the Pareto front---improving over the hypervolume of the single cost-optimal solution in later generations, see the left-hand plot of Fig.~\ref{fig:costVsRandBoxPlots}.
+As we will see below, the reason is a high correlation between the risk and costs of each arc, which seemingly results in the fact that no non-dominated solution can be reached from the initial solution by the applied mutation operator.
+However, the comparison between the two initialization schemes on the lazio\_4\_1 instance shows that also the algorithm employing the random initialization can find solutions which are close to the Pareto front---improving over the hypervolume of the single cost-optimal solution in later generations, see the left-hand plot of Fig.~\ref{fig:costVsRandBoxPlots}.
 
 To show that the cost-optimal initialization is not useless, we run the algorithms again on the instance uncorrelated\_4 where the cost and risk values are fully uncorrelated.
-In this case, there are more local changes of the truck paths that result in non-dominated solutions and eventually in a larger variaty of solutions found by the algorithm employing the cost-optimal initialization (see Fig.~\ref{fig:randVsCostNew}).
+In this case, there are more local changes of the truck paths that result in non-dominated solutions and eventually in a larger variety of solutions found by the algorithm employing the cost-optimal initialization (see Fig.~\ref{fig:randVsCostNew}).
 Due to the fact that the cost-optimal initialization starts already on the Pareto-front, it also results in higher hypervolume values than for the random initialization variant---both for a low and a high number of generations, see the right-hand side plot of Fig.~\ref{fig:costVsRandBoxPlots}.
+All algorithms of Fig.~\ref{fig:costVsRandBoxPlots} are pairwisely significantly different with respect to their hypervolume values according to Wilcoxon rank sum tests with p-values below 0.01.
 
@@
 To this end, Fig.~\ref{fig:popsizes} shows the nondominated solutions of 10 independent runs after 100,000 function evaluations for three different versions of HypE where the population size changes from 50 over 100 to 200 on the largest instance lazio\_4\_1 with 4 commodities.
-From the visual inspection of the resulting solutions, we cannot make a statement about which population size to favour.
+From the visual inspection of the resulting solutions, we cannot make a statement about which population size to favor.
 When comparing the hypervolume values of the resulting populations, however, an effect is visible.
-Figure~\ref{fig:algoComparison} shows the corresponding box plots of the hypervolume values and one-sided Wilcoxon rank sum tests on the populations' hypervolume indicator values reveil statistical significant differences between two of the three possible algorithm pairs.
+Figure~\ref{fig:algoComparison} shows the corresponding box plots of the hypervolume values and one-sided Wilcoxon rank sum tests on the populations' hypervolume indicator values reveal statistical significant differences between two of the three possible algorithm pairs.
 With a $p$-value of $0.02163$, the null hypothesis of no differences between a population size of 50 and a population size of 100 has to be rejected in favor of stating that the algorithm variant with population size 50 is superior to the one with population size 100.
@@
 The Wilcoxon rank sum test comparing population sizes 25 and 50 results in a $p$-value of $0.004465$ whereas the other comparisons are not statistically significantly different.
-To conclude, among the tested population sizes, 50 seems to be the best choice for budgets around 100,000 function evaluations.
-Also with longer runs, we recommend to keep the population size of the algorithm around 50 as the HypE implementation in PISA becomes too costly if the population size is much larger.
+To conclude, among the tested population sizes, 50 seems to be the best choice for budgets around 100,000 function evaluations but the population size is not a parameter that is difficult to adjust for the user.
+According to the runtime of the algorithm, we recommend to keep the population size of the algorithm around 50 as the HypE implementation in PISA becomes too costly if the population size is much larger.
 Ten runs with 2000 iterations and population size 200, for example, need already around 20 hours to complete on an Intel Core2 Duo with 2.8GHz.
@@
 
-Instead of the environmental selection of HypE \citep{bz2011a}, other selection schemes can be easily used with the PISA framework and we show the resutls of two other algorithms exemplary.
+Instead of the environmental selection of HypE \citep{bz2011a}, other selection schemes can be easily used with the PISA framework and we show the results of two other algorithms exemplary.
 On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}.
 The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations for NSGA-II and population size of 25 and 4,000 generations for W-HypE) are shown in Fig.~\ref{fig:algoComparison}.
-What can be seen is that both algorithms show a reduced performance regarding the standard hypervolume value with W-HypE being much worse than NSGA-II.
-For NSGA-II, this decrease  in hypervolume is expected as its environmental selection schemes does not optimize the hypervolume indicator as does HypE.
+What can be seen is that both algorithms show a reduced performance regarding the standard hypervolume value, with W-HypE being much worse than NSGA-II.
+For NSGA-II, this decrease  in hypervolume is expected as its environmental selection schemes does not optimize the hypervolume indicator directly as HypE does.
 Moreover, NSGA-II is known to become less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}.
-This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing quality measures that do not comply with the Pareto dominance relation such as the crwoding distance of NSGA-II.
+This is less the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing quality measures that do not comply with the Pareto dominance relation such as the crowding distance of NSGA-II.
 
@@
 W-HypE and the underlying weighted hypervolume indicator was developed to steer the search towards regions of the objective space that have been specified by the user.
-To this end, instead of the standard hypervolume indicator, the weighted hypervolume indicator of \citep{zbt2007a} is optimized.
+To this end, instead of the standard hypervolume indicator, the weighted hypervolume indicator of \citep{zbt2007a} is optimized in W-Hype \citep{abbz2009b,bbtz2001a}.
 The idea behind W-HypE is that solutions in objective space regions with a higher weight will get a higher fitness due to the \emph{weighted} hypervolume they dominate.
@@
 While HypE originally finds 6 non-dominated solutions within the box, W-HypE finds 14 within the same budget of function evaluations.
-When both algorithm outcomes are combined, it turns out that 13 of W-HypE's solutions are still non-dominated while 5 out of the 6 solutions found by HypE are dominated by W-HypE solutions.
+When both algorithm outcomes are combined, it turns out that 13 of W-HypE's solutions within the box are still non-dominated while 5 out of the 6 solutions found by HypE are dominated by W-HypE solutions.
 It has to be noted that the advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations (results not shown here).
@@
 
-Regarding the two initialization schemes, it turned out that in combination with the (local and quite simple) variation operator, the cost-optimal initialization does not allow the algorithm to find additional solutions besides the initial one.
-However, if the cost and total risk objectives are less correlated, this cost-optimal initialization has advantages over the random initialization in the early stage of the search when compared with respect to the hypervolume indicator.
-In general, however, we suggest to use the random initialization if the number of function evaluations is not a limiting factor.
+Regarding the two initialization schemes, it turns out that in combination with the (local and quite simple) variation operator, the cost-optimal initialization does not allow the algorithm to find additional solutions besides the initial one on the Lazio instances.
+However, if the cost and total risk objectives are uncorrelated, this cost-optimal initialization has advantages over the random initialization in all stages of the search when compared with respect to the hypervolume indicator.
+In general, however, we suggest to use the random initialization\footnote{or a combination with a more exploring mutation operator} if the costs and risks are highly correlated.
 
 Using other initializations as well as additional variation operators from the literature is easily possible with the proposed algorithm but their investigation is left for future research.
-For example, one can think of a ``more random'' initialization where the initial truck paths are generated by iteratively applying the mutation operator on the empty path and also the multiple application of the described mutation operator within one single mutation might be beneficial as the explored space of neighbored paths is larger in comparison to the current mutation.
-Also well-designed crossover operators are expected to be beneficial to the problem when, for example, good sub paths can be exchanged between solution pairs.
+For example, one can think of a ``more random'' initialization where the initial truck paths are generated by iteratively applying the mutation operator on the empty path.
+Also the multiple application of the described mutation operator within one single mutation might be beneficial as the explored space of neighbored paths is larger in comparison to the current mutation.
+Well-designed crossover operators can also be expected to be beneficial to the problem when, for example, good sub-paths can be exchanged between solution pairs.
 
 Overall, it is to say that the approach is very general and can easily be adapted if the problem formulation is (slightly) changed as it is often the case in practice, e.g., if different numbers of trucks or different types of vehicles with different loads are available.
-Also in the case when the definition of total and equity risk changes, it is only the objective function code that needs to be changed---a big advantage of blackbox optimizers such as for the proposed EMO algorithm.
+Also in the case when the definition of total and equity risk changes, it is only the objective function code that needs to be changed---a big advantage of blackbox optimizers such as the proposed EMO algorithm.
 
@@
 The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest.
-Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary multiobjective optimization (EMO) algorithm to cope with the non-linear equity risk objective.
+Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary multiobjective optimization (EMO) algorithm to cope with the non-linear equity risk objective in a true multiobjective sense.
 
@@
 With its random initialization, the algorithm allows to find sets of non-dominated solutions in reasonable time which are close to the Pareto-optimal front.
-For the tested real-world instances with the road network of the Lazio region and dozens of trucks for up to 4 hazmat commodities, the current Java implementation allows to finish the 100,000 function evaluations of our experiments within less than 15 minutes on a standard Intel Core i5 laptop.
+For the tested real-world instances with the road network of the Lazio region and dozens of trucks for up to 4 hazmat commodities, the current Java implementation allows to finish the 100,000 function evaluations of a single experiment within less than 15 minutes on a standard Intel Core i5 laptop.
+In addition, the fact that EMO algorithms are \emph{anytime} algorithms is another advantage which allows the user to interrupt the algorithm at anytime while the algorithm provides a feasible set of solutions.
 
 In future work, one might want to look further into adding (simple) heuristics within the EMO algorithm in order to improve the convergence of the algorithm.
-For example, one can observe that the solutions produced by the algorithm sometimes uses one and the same arc within a truck path, hence introducing cylces.
+For example, one can observe that the solutions produced by the algorithm sometimes uses one and the same arc within a truck path, hence introducing cycles.
 With the three objectives cost, total and equity risk, such cycles can be easily prevented within the variation operator which might eventually result in a faster convergence of the algorithm to the Pareto front.
 Another path for future research is the consideration of more realistic problem instances.
-Stochastic risks or dynamically changing instances (for example due to road closures or traffic jams) would be interesting to address---challenges that have been dealt with by EMO algorithms in other application domains already before \cite{deb2001,cvl2007a}.
+Stochastic risks or dynamically changing instances (for example due to road closures or traffic jams) would be interesting to address---challenges that have been dealt with by EMO algorithms in other application domains already before \cite{deb2001a,cvl2007a}.
 
@@
 The authors would like to thank Stefano Giordani for providing us with the instances of the Lazio network.
-Moreover, Dimo Brockhoff would like to acknowledge support from the CNRS-Microsoft chair "`Optimization for Sustainable Development"' during his time at Ecole Polytechnique, Palaiseau, France.
+Moreover, Dimo Brockhoff would like to acknowledge support from the CNRS-Microsoft chair \emph{Optimization for Sustainable Development} during his time at Ecole Polytechnique, Palaiseau, France.
 

### 2011-11-30 DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/hypervolumes/hypervolumeAlgoComparisons.m
[tag] version as submitted to Discrete Applied Mathematics on November 30, 2011

@@ -1,0 +1,33 @@
+addpath('../');
+
+hypervolumes = zeros(10, 6);
+
+% HypE popsize 25
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize25_randInit.4000';
+hypervolumes(:,1) = -1 * dlmread(filename);
+% HypE popsize 50
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize50_randInit.2000';
+hypervolumes(:,2) = -1 * dlmread(filename);
+% HypE popsize 100
+filename = 'HYP_ns4_1_OriginalCosts_randInit.1000';
+hypervolumes(:,3) = -1 * dlmread(filename);
+% HypE popsize 200
+filename = 'HYP_ns4_1_OriginalCosts_HypE_popsize200_randInit.500';
+hypervolumes(:,4) = -1 * dlmread(filename);
+% NSGA-II
+filename = 'HYP_ns4_1_OriginalCosts_nsga2_randInit.1000';
+hypervolumes(:,5) = -1 * dlmread(filename);
+% W-HypE
+filename = 'HYP_ns4_1_OriginalCosts_whype_box2_popsize25_randInit.4000';
+hypervolumes(:,6) = -1 * dlmread(filename);
+
+
+
+
+boxplot(hypervolumes, 'Labels', {'HypE_25'; 'HypE_50'; 'HypE_100'; 'HypE_200'; 'NSGA-II'; 'W-HypE'});
+
+
+
+% save plot as eps:
+exportfig(gcf, 'algoComparison.eps');
+

### 2011-11-30 DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/interactivemonitor/interaction.py

@@ -1,0 +1,374 @@
+import matplotlib.pyplot as plt
+from Pycluster import *
+from scipy.spatial.distance import *
+from scipy import stats
+import numpy as np
+import AnnoteFinder
+from pylab import connect
+from intmon_internal import intMonIO as io
+from intmon import wait
+import math
+import onClick
+from comparativePrefStatements import *
+
+def parallelCoordinatesPlot(objVectors, generation, run, params):
+    color = ['b', 'g', 'r', 'c', 'm', 'y', 'k']
+    
+    if (params.interaction == 1):
+        plt.clf()
+        plt.hold('on')
+        
+        # sort objective vectors wrt last objective for nicer plotting
+        I = np.argsort(objVectors[:,objVectors.shape[1]-1])
+        objVectors = objVectors[I,:]
+        
+        # add (kind of) id of solution as last column for nicer printing:
+        maximum = objVectors.max(0).max(0)
+        minimum = objVectors.min(0).min(0)
+        objVectors = np.column_stack((objVectors,
+            np.linspace(minimum, maximum, objVectors.shape[0]).transpose()))
+
+        i = 0
+        for objVector in objVectors:
+            i = i + 1
+            l = len(objVector)
+            
+            x = np.hstack((range(l-1), len(objVector)-1.5))
+            plt.plot(x, objVector, color[i % len(color) - 1])
+        
+        # add ids on the right:
+        i = 0
+        for s in objVectors:
+            plt.text(len(s)-1.45, s[len(s)-1], i, fontsize=12)
+            i = i + 1
+            
+        plt.tick_params(axis='both', which='major', direction='out')
+        plt.grid('on', 'major')
+        plt.xlabel('objectives')
+        plt.ylabel('values')
+        
+        # create and place xtick labels
+        ticksOnXAxis = []
+        i = 1
+        while i < objVectors.shape[1]:
+            ticksOnXAxis.append('f' + str(i))
+            i = i + 1
+        plt.xticks( np.arange(objVectors.shape[1]-1), ticksOnXAxis )
+        
+        # make room at the top for placing some interactive boxes:
+        ymin, ymax = plt.ylim()   # return the current ylim
+        plt.ylim( ( ymin, ymax + (ymax-ymin)*0.1 ) )
+
+        # draw first box for using preference statements
+        plt.fill([0, 0, 1, 1], [ymax, ymax + (ymax-ymin)*0.1, ymax + (ymax-ymin)*0.1, ymax], 'b' )
[…]

### 2011-11-30 DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/interactivemonitor/intmon_internal.py

@@ -1,0 +1,675 @@
+# ========================================================================
+#  PISA  (www.tik.ee.ethz.ch/pisa/)
+# 
+#  ========================================================================
+#  Computer Engineering Lab (TIK)
+#  ETH Zurich
+# 
+#  ========================================================================
+#  Interactive Monitor 
+#  
+#  internal classes and functions such as IntMonParameters and IntMonState
+#  
+#  python implementation
+#  
+#  file: intmon_internal.py
+#  author: Dimo Brockhoff  <brockho@lix.polytechnique.fr>
+#
+#  last change: April 2011
+#  
+#  ========================================================================
+
+from collections import deque
+import sys
+import numpy as np
+#import matplotlib.pyplot as plt
+#import time
+
+
+class IntMonParameters:
+    """ stores all parameters of the interactive PISA monitor and provides\n the corresponding file input/output methods """
+    
+    def read_common_parameters(self,filename):
+        """ reads common parameters alpha, mu, lam, and dimension from 'cfg' file and tests for inconsistencies """
+
+        try:
+            f = open(filename, 'r')
+        except IOError:
+            print 'Problem reading file ' + filename
+
+        for line in f:
+            if 'alpha' in line:
+                newalpha = int(line[6:len(line)])
+            if 'mu' in line:
+                newmu = int(line[3:len(line)])
+            if 'lambda' in line:
+                newlam = int(line[7:len(line)])
+            if 'dim' in line:
+                newdim = int(line[4:len(line)])
+
+        f.closed
+
+        # test for inconsistency:
+        if self.alpha == 0:
+            self.alpha = newalpha
+            self.mu = newmu
+            self.lam = newlam
+            self.dimension = newdim
+        else:
+            assert newalpha == self.alpha
+            assert newmu == self.mu
+            assert newlam == self.lam
+            assert newdim == self.dimension
+
+        
+        
+    # pre-condition: args contains exactly 7 entries
+    def read_selectorvariator_parameters(self,args):
+        """ reads parameters from files """
+        self.poll = args.pop()
+        self.filenamebase_monitor = args.pop()
+        self.parameter_file_monitor = args.pop()
+        self.filenamebase_selector = args.pop()
+        self.parameter_file_selector = args.pop()
+        self.filenamebase_variator = args.pop()
+        self.parameter_file_variator = args.pop()
+        
+        
+        # generate file names based on 'filenamebase's
[…]

### 2011-11-30 DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/mysubplot3.m

@@ -1,0 +1,60 @@
+function [] = mysubplot3(A, B, C, A_label, B_label, C_label, titlename, filename)
+figure()
+
+subplot(2,2,1)
+plot3(A(:,1),A(:,2),A(:,3), 'x');
+hold all; grid on;
+plot3(B(:,1),B(:,2),B(:,3), 'o');
+plot3(C(:,1),C(:,2),C(:,3), 's');
+xlabel('cost')
+ylabel('total risk')
+zlabel('equity risk')
+view(45,30);
+
+subplot(2,2,2)
+plot(A(:,1),A(:,2), 'x');
+hold all; grid on;
+plot(B(:,1),B(:,2), 'o');
+plot(C(:,1),C(:,2), 's');
+%ylim([4,6]*1e6);
+h_legend = legend(A_label, B_label, C_label);
+set(h_legend,'FontSize',7);
+xlabel('cost')
+ylabel('total risk')
+
+
+subplot(2,2,3)
+plot(A(:,1),A(:,3), 'x');
+hold all; grid on;
+plot(B(:,1),B(:,3), 'o');
+plot(C(:,1),C(:,3), 's');
+%ylim([2,5]*1e5);
+h_legend = legend(A_label, B_label, C_label);
+set(h_legend,'FontSize',7);
+xlabel('cost')
+ylabel('equity risk')
+
+
+subplot(2,2,4)
+plot(A(:,2),A(:,3), 'x');
+hold all; grid on;
+plot(B(:,2),B(:,3), 'o');
+plot(C(:,2),C(:,3), 's');
+%ylim([2,5]*1e5);
+h_legend = legend(A_label, B_label, C_label);
+set(h_legend,'FontSize',7);
+xlabel('total risk')
+ylabel('equity risk')
+
+% save plot as eps:
+exportfig(gcf, filename);
+
+% add title above all for plot on screen:
+set(gcf,'NextPlot','add');
+axes;
+h = title(titlename);
+set(gca,'Visible','off');
+set(h,'Visible','on'); 
+
+
+end

### 2011-11-30 DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/mysubplotRandVsCost.m

@@ -1,0 +1,64 @@
+function [] = mysubplotRandVsCost(A, B, A_label, B_label, titlename, filename)
+figure()
+
+subplot(2,2,1)
+plot3(A(:,1),A(:,2),A(:,3), 'o');
+hold all; grid on;
+plot3(B(:,1),B(:,2),B(:,3), 'x');
+xlabel('cost')
+ylabel('total risk')
+zlabel('equity risk')
+if strncmp(filename, 'randVsCost_ns4_1_New_', 20) 
+    view([19.5 22]);
+    %xlim([0.5e4 1.2e4]);
+else
+    view([14.5 10]);
+    xlim([0.5e4 1.2e4]);
+end
+    
+subplot(2,2,2)
+plot(A(:,1),A(:,2), 'o');
+hold all; grid on;
+plot(B(:,1),B(:,2), 'x');
+%ylim([4,6]*1e6);
+h_legend = legend(A_label, B_label);
+set(h_legend,'FontSize',7);
+xlabel('cost')
+ylabel('total risk')
+if strncmp(filename, 'randVsCost_ns4_1_New_', 20) 
+    ylim([4.8 8]*1e7);
+end
+
+subplot(2,2,3)
+plot(A(:,1),A(:,3), 'o');
+hold all; grid on;
+plot(B(:,1),B(:,3), 'x');
+%ylim([2,5]*1e5);
+h_legend = legend(A_label, B_label);
+set(h_legend,'FontSize',7);
+xlabel('cost')
+ylabel('equity risk')
+
+
+subplot(2,2,4)
+plot(A(:,2),A(:,3), 'o');
+hold all; grid on;
+plot(B(:,2),B(:,3), 'x');
+%ylim([2,5]*1e5);
+h_legend = legend(A_label, B_label);
+set(h_legend,'FontSize',7);
+xlabel('total risk')
+ylabel('equity risk')
+
+% save plot as eps:
+exportfig(gcf, filename);
+
+% add title above all for plot on screen:
+set(gcf,'NextPlot','add');
+axes;
+h = title(titlename);
+set(gca,'Visible','off');
+set(h,'Visible','on'); 
+
+
+end

### 2011-11-30 DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/plot3d_after1000gen.m

@@ -1,0 +1,25 @@
+% plot resulting populations after 1000 generations
+
+C = dlmread('results/graph_ns2_1_New_costInit.1000');
+C_nondom = C(find(paretoFront(C(:,1:3))),:);
+%plot3(C(:,1),C(:,2),C(:,3), 'x');
+%hold all; grid on;
+%plot3(C_nondom(:,1),C_nondom(:,2),C_nondom(:,3), 'o');
+
+
+R = dlmread('results/graph_ns2_1_New_randInit.1000');
+R_nondom = R(find(paretoFront(R(:,1:3))),:);
+%figure()
+%plot3(R(:,1),R(:,2),R(:,3), 'x');
+%hold all; grid on;
+%plot3(R_nondom(:,1),R_nondom(:,2),R_nondom(:,3), 'o');
+
+
+
+%mysubplot(C,C_nondom, 'C', 'C_nondom', 'C only', 'filename');
+%mysubplot(R,R_nondom, 'R', 'R_nondom', 'R only', 'filename');
+
+mysubplot(C,R, 'shortest path initialization', 'random initialization', 'all solutions', 'allsolutions.eps')
+mysubplot(C_nondom,R_nondom, 'shortest path initialization', 'random initialization', 'nondominated', 'onlynondominated.eps')
+
+hold off;

### 2011-11-30 DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/plotCompareHypEAndNSGAII.m

@@ -1,0 +1,29 @@
+% plot resulting populations of HypE and NSGA-II runs
+% after certain number of generations
+
+%%%%%%%%%%%%%%%%%%%%%%%
+generation = 250;
+numOfCommodities = 4;
+instance = 1;
+%%%%%%%%%%%%%%%%%%%%%%%
+
+filename = ['results/graph_ns' int2str(numOfCommodities) '_' int2str(instance) '_OriginalCosts_randInit.' int2str(generation)];
+H = dlmread(filename);
+H_nondom = H(find(paretofront(H(:,1:3))),:);
+figure(77);
+plot3(H_nondom(:,1),H_nondom(:,2),H_nondom(:,3), 'o');
+hold all; grid on;
+
+filename = ['results/graph_ns' int2str(numOfCommodities) '_' int2str(instance) '_OriginalCosts_nsga2_randInit.' int2str(generation)];
+N = dlmread(filename);
+N_nondom = N(find(paretofront(N(:,1:3))),:);
+plot3(N_nondom(:,1),N_nondom(:,2),N_nondom(:,3), 'o');
+hold off;
+
+
+title = ['all solutions after ' int2str(generation) ' generations'];
+mysubplot(H,N, 'HypE\_randinit', 'NSGA-II\_randinit', title, 'allsolutions.eps');
+title = ['nondominated solutions after ' int2str(generation) ' generations'];
+mysubplot(H_nondom,N_nondom, 'HypE\_randinit', 'NSGA-II\_randinit', title, 'onlynondominated.eps');
+
+hold off;

### 2011-11-30 DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/plotOverTime.m

@@ -1,0 +1,33 @@
+% plot resulting populations of HypE_randinit over time
+
+%%%%%%%%%%%%%%%%%%%%%%%
+generation1 = 0;
+generation2 = 250;
+generation3 = 2000;
+numOfCommodities = 4;
+instance = 1;
+%%%%%%%%%%%%%%%%%%%%%%%
+
+title = 'HypE population after different number of generations';
+
+filename = ['results/graph_ns' int2str(numOfCommodities) '_' int2str(instance) '_OriginalCosts_randInit.' int2str(generation1)];
+H1 = dlmread(filename);
+H1_nondom = H1(find(paretofront(H1(:,1:3))),:);
+
+filename = ['results/graph_ns' int2str(numOfCommodities) '_' int2str(instance) '_OriginalCosts_randInit.' int2str(generation2)];
+H2 = dlmread(filename);
+H2_nondom = H2(find(paretofront(H2(:,1:3))),:);
+
+filename = ['results/graph_ns' int2str(numOfCommodities) '_' int2str(instance) '_OriginalCosts_randInit.' int2str(generation3)];
+H3 = dlmread(filename);
+H3_nondom = H3(find(paretofront(H3(:,1:3))),:);
+
+
+mysubplot3(H1_nondom,H2_nondom,H3_nondom, ['HypE\_randinit@generation ' int2str(generation1)], ...
+            ['HypE\_randinit@generation ' int2str(generation2)], ...
+            ['HypE\_randinit@generation ' int2str(generation3)], title, 'onlynondominated.eps');
+
+subplot(2,2,4)
+plot3(1,2,3, 'dr');
+
+hold off;

### 2011-11-30 trunk/DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/UniformSampling.m
[tag] version as submitted to Discrete Applied Mathematics on November 30, 2011

@@ -1,0 +1,7 @@
+function S = UniformSampling( nrOfSamples, lower, upper )
+    dim = length(upper);
+
+    ex = upper - lower;    
+    S = repmat( lower, nrOfSamples, 1 ) + ...
+        rand(nrOfSamples, dim )*diag(ex);    
+

### 2011-11-30 trunk/DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/interactivemonitor/comparativePrefStatements.py

@@ -1,0 +1,492 @@
+
+# preference statements should be given as a string of the following form:
+# - PREFER s OVER t
+# - where s and t can be simple statements such as f{1|...|k} {<|<=|==|>=|>} value
+# - and the statements themselves can be combined with 'AND'
+# Example: 'PREFER f1<4 AND f2>2 AND f3==4.2 OVER f1>4 AND f2<2'
+
+import sys
+import numpy as np
+import string
+from comparativePrefStatements import *
+
+class comparativePrefStatements:
+    
+    ps = [] # preference statements of the form 'PREFER p OVER q'
+    es = [] # equality statements of the form 'PREFER p EQUALLYTO q'
+
+    # constructor for a population 'solutionSet' given as numpy array with
+    # solutions as rows and objective values as columns while
+    # the preference statements are read from a file with filename
+    # 'fileName'
+    def __init__(self, solutionSet, fileName=None):
+        self.pop = solutionSet
+        if (fileName != None):
+            self.readPreferenceStatementsFromFile(fileName)
+
+    
+    def readPreferenceStatementsFromFile(self, fileName):
+        self.ps = []
+        self.es = []
+        
+        # read ids from 'fileName'
+        try:
+            file = open(fileName, 'r')
+            for line in file:                
+                if (line.find('OVER') >= 0):
+                    self.ps.append(line)
+                elif (line.find('EQUALLYTO') >= 0):
+                    self.es.append(line)
+            file.close()
+        except IOError:
+            print 'Problem reading file ' + fromFile
+
+    def getLofP(self):
+        return self.getAllLRSetPairs(self.ps)
+    
+    def getQofP(self):
+        return self.getAllLRSetPairs(self.es)
+
+    def getAllLRSetPairs(self, preferenceStatements):
+        allIDs = np.arange(self.pop.shape[0])
+        return self.getLRSetPairs(preferenceStatements, allIDs)
+
+    def getLRSetPairs(self, preferenceStatements, ids):
+        LRSetPairs = []
+        for ps in preferenceStatements:
+            LRSetPairs.append(self.getSetsLandRinAccordanceToPreference(ps, ids))
+        return LRSetPairs
+
+    def getSetsLandRinAccordanceToPreference(self, preferenceStatement,ids):
+        pandq = self.extractPandQ(preferenceStatement)
+        
+        L = self.getSolutionsSatisfyingPbutNotQ(pandq, ids)
[…]

### 2011-11-30 trunk/DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/mysubplot.m

@@ -1,0 +1,56 @@
+function [] = mysubplot(C, R, C_label, R_label, titlename, filename)
+figure()
+
+subplot(2,2,1)
+plot3(C(:,1),C(:,2),C(:,3), 'x');
+hold all; grid on;
+plot3(R(:,1),R(:,2),R(:,3), 'o');
+xlabel('cost')
+ylabel('total risk')
+zlabel('equity risk')
+
+
+subplot(2,2,2)
+plot(C(:,1),C(:,2), 'x');
+hold all; grid on;
+plot(R(:,1),R(:,2), 'o');
+%ylim([4,6]*1e6);
+h_legend = legend(C_label, R_label);
+set(h_legend,'FontSize',7);
+xlabel('cost')
+ylabel('total risk')
+
+
+subplot(2,2,3)
+plot(C(:,1),C(:,3), 'x');
+hold all; grid on;
+plot(R(:,1),R(:,3), 'o');
+%ylim([2,5]*1e5);
+h_legend = legend(C_label, R_label);
+set(h_legend,'FontSize',7);
+xlabel('cost')
+ylabel('equity risk')
+
+
+subplot(2,2,4)
+plot(C(:,2),C(:,3), 'x');
+hold all; grid on;
+plot(R(:,2),R(:,3), 'o');
+%ylim([2,5]*1e5);
+h_legend = legend(C_label, R_label);
+set(h_legend,'FontSize',7);
+xlabel('total risk')
+ylabel('equity risk')
+
+% save plot as eps:
+exportfig(gcf, filename);
+
+% add title above all for plot on screen:
+set(gcf,'NextPlot','add');
+axes;
+h = title(titlename);
+set(gca,'Visible','off');
+set(h,'Visible','on'); 
+
+
+end

### 2011-11-30 trunk/DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/mysubplot3sampling.m

@@ -1,0 +1,85 @@
+function [] = mysubplot3sampling(A, B, Cupperlower, A_label, B_label, Cupperlower_label, titlename, filename)
+figure()
+
+% create 2D and 3D polygon for sampling box from Cupperlower:
+X3 = [Cupperlower(1,1) Cupperlower(1,2) Cupperlower(1,3);
+    Cupperlower(2,1) Cupperlower(1,2) Cupperlower(1,3);
+    Cupperlower(2,1) Cupperlower(2,2) Cupperlower(1,3);
+    Cupperlower(1,1) Cupperlower(2,2) Cupperlower(1,3);
+    Cupperlower(1,1) Cupperlower(1,2) Cupperlower(2,3);
+    Cupperlower(2,1) Cupperlower(1,2) Cupperlower(2,3);
+    Cupperlower(2,1) Cupperlower(2,2) Cupperlower(2,3);
+    Cupperlower(1,1) Cupperlower(2,2) Cupperlower(2,3)];
+
+X3_d = X3(1:4, :);
+X3_l = [X3(1,:); X3(4,:); X3(8,:); X3(5,:)];
+X3_r = [X3(4,:); X3(3,:); X3(7:8,:)];
+
+X_12 = X3(1:4,:);
+X_13 = [X3(1:2,:); X3(6,:); X3(5,:)];
+X_23 = [X3(1,:); X3(4,:); X3(8,:); X3(5,:)];
+
+C = 'y';
+
+subplot(2,2,1)
+fill3(X3_d(:,1),X3_d(:,2),X3_d(:,3),C);
+hold all; grid on;
+fill3(X3_l(:,1),X3_l(:,2),X3_l(:,3),C);
+fill3(X3_r(:,1),X3_r(:,2),X3_r(:,3),C);
+plot3([X3(2,1); X3(6,1); X3(5,1)], [X3(2,2); X3(6,2); X3(5,2)], [X3(2,3); X3(6,3); X3(5,3)], '-k');
+plot3(X3(6:7,1), X3(6:7,2), X3(6:7,3), '-k');
+plot3(A(:,1),A(:,2),A(:,3), 'x');
+plot3(B(:,1),B(:,2),B(:,3), 'o');
+xlabel('cost')
+ylabel('total risk')
+zlabel('equity risk')
+view([14.5 10]);
+xlim([0.5e4 1.2e4]);
+
+subplot(2,2,2)
[…]

### 2011-11-30 trunk/DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/mysubplot5.m

@@ -1,0 +1,69 @@
+function [] = mysubplot5(A, B, C, D, E, A_label, B_label, C_label, D_label, E_label, titlename, filename)
+figure()
+
+subplot(2,2,1)
+plot3(A(:,1),A(:,2),A(:,3), 's');
+hold all; grid on;
+plot3(B(:,1),B(:,2),B(:,3), 'o');
+plot3(C(:,1),C(:,2),C(:,3), 'd');
+plot3(D(:,1),D(:,2),D(:,3), 'x');
+plot3(E(:,1),E(:,2),E(:,3), 'p');
+xlabel('cost')
+ylabel('total risk')
+zlabel('equity risk')
+view(15,30);
+xlim([0 15000]);
+
+subplot(2,2,2)
+plot(A(:,1),A(:,2), 's');
+hold all; grid on;
+plot(B(:,1),B(:,2), 'o');
+plot(C(:,1),C(:,2), 'd');
+plot(D(:,1),D(:,2), 'x');
+plot(E(:,1),E(:,2), 'p');
+%ylim([4,6]*1e6);
+h_legend = legend(A_label, B_label, C_label, D_label, E_label);
+set(h_legend,'FontSize',6, 'Location', 'SouthEast');
+xlabel('cost')
+ylabel('total risk')
+
+
+subplot(2,2,3)
+plot(A(:,1),A(:,3), 's');
+hold all; grid on;
+plot(B(:,1),B(:,3), 'o');
+plot(C(:,1),C(:,3), 'd');
+plot(D(:,1),D(:,3), 'x');
+plot(E(:,1),E(:,3), 'p');
+%ylim([2,5]*1e5);
+%h_legend = legend(A_label, B_label, C_label, D_label, E_label);
+%set(h_legend,'FontSize',7);
+xlabel('cost')
+ylabel('equity risk')
+
+
+subplot(2,2,4)
+plot(A(:,2),A(:,3), 's');
+hold all; grid on;
+plot(B(:,2),B(:,3), 'o');
+plot(C(:,2),C(:,3), 'd');
+plot(D(:,2),D(:,3), 'x', 'MarkerSize', 12);
+plot(E(:,2),E(:,3), 'p');
+xlim([0,5]*1e7);
+%h_legend = legend(A_label, B_label, C_label, D_label, E_label);
+%set(h_legend,'FontSize',7);
+xlabel('total risk')
+ylabel('equity risk')
+
+% save plot as eps:
+exportfig(gcf, filename);
+
[…]

### 2011-11-30 trunk/DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/plotCompareAll.m

@@ -1,0 +1,40 @@
+% plots the resulting populations of lazio_2 and lazio_3
+
+%%%%%%%%%%%%%%%%%%%%%%%
+generation = 2000;
+%%%%%%%%%%%%%%%%%%%%%%%
+
+filename = ['results/graph_ns2_1_OriginalCosts_HypE_popsize50_randInit.' int2str(generation)];
+Lazio2 = dlmread(filename);
+Lazio2_nondom = Lazio2(find(paretofront(Lazio2(:,1:3))),:);
+figure(88);
+plot3(Lazio2_nondom(:,1),Lazio2_nondom(:,2),Lazio2_nondom(:,3), 'x');
+hold all; grid on;
+
+filename = ['results/graph_ns3_1_OriginalCosts_HypE_popsize50_randInit.' int2str(generation)];
+Lazio3 = dlmread(filename);
+Lazio3_nondom = Lazio3(find(paretofront(Lazio3(:,1:3))),:);
+plot3(Lazio3_nondom(:,1),Lazio3_nondom(:,2),Lazio3_nondom(:,3), 'o');
+
+filename = ['results/graph_ns4_1_OriginalCosts_HypE_popsize50_randInit.' int2str(generation)];
+Lazio41 = dlmread(filename);
+Lazio41_nondom = Lazio41(find(paretofront(Lazio41(:,1:3))),:);
+plot3(Lazio41_nondom(:,1),Lazio41_nondom(:,2),Lazio41_nondom(:,3), 's');
+
+filename = ['results/graph_ns4_2_OriginalCosts_HypE_popsize50_randInit.' int2str(generation)];
+Lazio42 = dlmread(filename);
+Lazio42_nondom = Lazio42(find(paretofront(Lazio42(:,1:3))),:);
+plot3(Lazio42_nondom(:,1),Lazio42_nondom(:,2),Lazio42_nondom(:,3), 'd');
+
+filename = ['results/graph_ns4_3_OriginalCosts_HypE_popsize50_randInit.' int2str(generation)];
+Lazio43 = dlmread(filename);
+Lazio43_nondom = Lazio43(find(paretofront(Lazio43(:,1:3))),:);
+plot3(Lazio43_nondom(:,1),Lazio43_nondom(:,2),Lazio43_nondom(:,3), 'd');
+hold off;
+
+
+title = ['nondominated solutions after ' int2str(generation) ' generations'];
[…]

### 2011-11-30 trunk/DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/plotCompareRandAndCost.m

@@ -1,0 +1,49 @@
+% plot resulting populations of rand and cost initialization
+% after certain number of generations
+
+%%%%%%%%%%%%%%%%%%%%%%%
+generation = 2000;
+%%%%%%%%%%%%%%%%%%%%%%%
+
+% first ns4_1_New:
+%%%%%%%%%%%%%%%%%%
+filename = ['results/graph_ns4_1_New_HypE_popsize50_costInit.' int2str(generation)];
+C = dlmread(filename);
+C_nondom = C(find(paretofront(C(:,1:3))),:);
+figure(77);
+plot3(C_nondom(:,1),C_nondom(:,2),C_nondom(:,3), 'o');
+hold all; grid on;
+
+filename = ['results/graph_ns4_1_New_HypE_popsize50_randInit.' int2str(generation)];
+R = dlmread(filename);
+R_nondom = R(find(paretofront(R(:,1:3))),:);
+plot3(R_nondom(:,1),R_nondom(:,2),R_nondom(:,3), 'o');
+hold off;
+
+title = ['nondominated solutions after ' int2str(generation) ' generations'];
+filename = ['randVsCost_ns4_1_New_' int2str(generation) 'gens.eps'];
+mysubplotRandVsCost(R_nondom, C_nondom, 'random', 'shortest path', title, filename);
+
+hold off;
+
+
+% now ns4_1_OriginalCost:
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+filename = ['results/graph_ns4_1_OriginalCosts_HypE_popsize50_costInit.' int2str(generation)];
+C = dlmread(filename);
+C_nondom = C(find(paretofront(C(:,1:3))),:);
+figure(78);
+plot3(C_nondom(:,1),C_nondom(:,2),C_nondom(:,3), 'o');
+hold all; grid on;
+
+filename = ['results/graph_ns4_1_OriginalCosts_HypE_popsize50_randInit.' int2str(generation)];
+R = dlmread(filename);
+R_nondom = R(find(paretofront(R(:,1:3))),:);
+plot3(R_nondom(:,1),R_nondom(:,2),R_nondom(:,3), 'o');
+hold off;
+
+title = ['nondominated solutions after ' int2str(generation) ' generations'];
+filename = ['randVsCost_ns4_1_OriginalCost_' int2str(generation) 'gens.eps'];
+mysubplotRandVsCost(R_nondom, C_nondom, 'random', 'shortest path', title, filename);
+
+hold off;

### 2011-11-30 trunk/DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/plotCompareWHypE_box2.m

@@ -1,0 +1,37 @@
+% plot resulting populations of W-HypE vs. HypE
+% after certain number of generations
+
+%%%%%%%%%%%%%%%%%%%%%%%
+generation = 4000;
+%%%%%%%%%%%%%%%%%%%%%%%
+
+% sampling box:
+Supperlower = [7e3 3e7 0.5e6; 1e4 4e7 1.1e6];
+
+figure(2345);
+S = UniformSampling(10000,Supperlower(1,:), Supperlower(2,:));
+plot3(S(:,1),S(:,2),S(:,3), 'y.', 'MarkerSize', 2);
+hold all; grid on;
+
+% first W-HypE:
+%%%%%%%%%%%%%%%
+filename = ['results/graph_ns4_1_OriginalCosts_WHypE_box2_popsize25_randInit.' int2str(generation)];
+W = dlmread(filename);
+W_nondom = W(find(paretofront(W(:,1:3))),:);
+plot3(W_nondom(:,1),W_nondom(:,2),W_nondom(:,3), 'o');
+
+filename = ['results/graph_ns4_1_OriginalCosts_HypE_popsize25_randInit.' int2str(generation)];
+H = dlmread(filename);
+H_nondom = H(find(paretofront(H(:,1:3))),:);
+plot3(H_nondom(:,1),H_nondom(:,2),H_nondom(:,3), 'o');
+
+
+all = [W_nondom; H_nondom];
+all = all(find(paretofront(all(:,1:3))),:);
+plot3(all(:,1),all(:,2),all(:,3), 'sr');
+hold off;
+
+title = ['nondominated solutions after ' int2str(generation) ' generations'];
+mysubplot3sampling(W_nondom, H_nondom, Supperlower, 'W-HypE', 'HypE', 'sample box', title, 'WHypEVsHypE_ns4_1_OriginalCosts2.eps');
+
+hold off;

### 2011-11-30 trunk/DAW_hazmatHypE/tags/submittedToDAM/experiments/randVsCost/plotDiffPopsizes.m

@@ -1,0 +1,43 @@
+% plot resulting populations of HypE with rand initialization
+% after 100,000 funevals to compare different population sizes
+
+figure(77);
+hold all; grid on;
+
+filename = 'results/graph_ns4_1_OriginalCosts_HypE_popsize25_randInit.4000';
+H25 = dlmread(filename);
+H25_nondom = H25(find(paretofront(H25(:,1:3))),:);
+plot3(H25_nondom(:,1),H25_nondom(:,2),H25_nondom(:,3), 'x');
+
+filename = 'results/graph_ns4_1_OriginalCosts_HypE_popsize50_randInit.2000';
+H50 = dlmread(filename);
+H50_nondom = H50(find(paretofront(H50(:,1:3))),:);
+plot3(H50_nondom(:,1),H50_nondom(:,2),H50_nondom(:,3), 'o');
+
+filename = 'results/graph_ns4_1_OriginalCosts_randInit.1000';
+H100 = dlmread(filename);
+H100_nondom = H100(find(paretofront(H100(:,1:3))),:);
+plot3(H100_nondom(:,1),H100_nondom(:,2),H100_nondom(:,3), 's');
+
+filename = 'results/graph_ns4_1_OriginalCosts_HypE_popsize200_randInit.500';
+H200 = dlmread(filename);
+H200_nondom = H200(find(paretofront(H200(:,1:3))),:);
+plot3(H200_nondom(:,1),H200_nondom(:,2),H200_nondom(:,3), 'd');
+
+allnondom = [H25; H50; H100; H200];
+allnondom = allnondom(find(paretofront(allnondom(:,1:3))),:);
+plot3(allnondom(:,1),allnondom(:,2),allnondom(:,3), 'pk');
+xlabel('f_1')
+ylabel('f_2')
+zlabel('f_3')
+view(45,30);
+
+
+
+mysubplot3(H50_nondom, H100_nondom, H200_nondom, 'HypE, popsize 50', 'HypE, popsize 100', 'HypE, popsize 200', 'Nondominated solutions of 10 runs after 100,000 funevals', 'diffPopsizes.eps');
+
+
+hold off;
+
+
+

### 2011-11-30 DAW_hazmatHypE/tags/submittedToDAM/paper/submissionDAM/hazmatRouting_EvolOptim.tex
renamed misspelled subfolder

--- 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
@@
 \begin{figure}%
-\includegraphics[width=\columnwidth]{figs/truckpaths}%
+\includegraphics[width=\columnwidth]{truckpaths}%
 \caption{\label{fig:paths} Illustration of the variable-length representation of a solution (left) with its two paths for commodity 1 and one path for commodity 2 in opposition to the resulting full source--destination paths where the last part is the shortest path between the variable-length path's end node and the destination node for this commodity (right).
@@
 	\centering
-	\includegraphics[width=0.48\columnwidth]{figs/correlationLazio41.eps}
+	\includegraphics[width=0.48\columnwidth]{correlationLazio41.eps}
 	\hfill
-	\includegraphics[width=0.48\columnwidth]{figs/correlationUncorrelated4.eps}
+	\includegraphics[width=0.48\columnwidth]{correlationUncorrelated4.eps}
 	\caption{\label{fig:correlation} Correlation between the costs and the risks of the first commodity for both the lazio\_4\_1 instance (left) and the uncorrelated\_4 instance (right).
@@
 	\centering
-	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/compAll_2000gens.eps}
+	\includegraphics[width=0.75\columnwidth]{compAll_2000gens.eps}
 	\caption{\label{fig:compareAll} Comparison between the different instances lazio\_2 (squares), lazio\_3 (circles), lazio\_4\_1 (diamonds), lazio\_4\_2 (crosses), and lazio\_4\_3 (stars).
@@
 	\centering
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}
+	\includegraphics[width=0.48\columnwidth]{hypervolumeOverTime_50.eps}
 	\hfill
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}
+	\includegraphics[width=0.48\columnwidth]{hypervolumeOverTime_100.eps}
 	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
@@
 	\centering
-	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/randVsCost_ns4_1_OriginalCost_2000gens.eps}
+	\includegraphics[width=0.75\columnwidth]{randVsCost_ns4_1_OriginalCost_2000gens.eps}
 	\vspace{-1em}
@@
 	\centering
-	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/randVsCost_ns4_1_New_2000gens.eps}
+	\includegraphics[width=0.75\columnwidth]{randVsCost_ns4_1_New_2000gens.eps}
 	\vspace{-1em}
@@
 	\centering
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/initBoxPlots_OriginalCosts.eps}
+	\includegraphics[width=0.48\columnwidth]{initBoxPlots_OriginalCosts.eps}
 	\hfill
-	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/initBoxPlots_New.eps}
+	\includegraphics[width=0.48\columnwidth]{initBoxPlots_New.eps}
 	\vspace{-1em}
@@
 	\centering
-	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/diffPopsizes}%
+	\includegraphics[width=0.75\columnwidth]{diffPopsizes}%
 	\vspace{-1em}
@@
 	\centering
-	\includegraphics[width=0.5\columnwidth]{../experiments/randVsCost/hypervolumes/algoComparison}%
+	\includegraphics[width=0.5\columnwidth]{algoComparison}%
 	\caption{\label{fig:algoComparison} Boxplots of the hypervolume values of 10 independent runs after 100,000 function evaluations of HypE with a population size of 25, 50, 100, and 200, as well as for NSGA-II.}
@@
 	\centering
-	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/WHypEVsHypE_ns4_1_OriginalCosts2.eps}
+	\includegraphics[width=0.75\columnwidth]{WHypEVsHypE_ns4_1_OriginalCosts2.eps}
 	\vspace{-1em}
@@
 
-% 
- \bibliographystyle{plainnat}
- \footnotesize
- \bibliography{all}
+\begin{thebibliography}{20}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\url}[1]{\texttt{#1}}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[Akgun et~al.(2000)Akgun, Erkut, and Batta]{aeb2000a}
+V.~Akgun, E.~Erkut, and R.~Batta.
+\newblock {On Finding Dissimilar Paths}.
+\newblock \emph{European Journal of Operational Research}, 121\penalty0
+  (2):\penalty0 232--246, 2000.
+\newblock \doi{10.1016/S0377-2217(99)00214-3}.
+
+\bibitem[Auger et~al.(2009)Auger, Bader, Brockhoff, and Zitzler]{abbz2009b}
+A.~Auger, J.~Bader, D.~Brockhoff, and E.~Zitzler.
+\newblock {Articulating User Preferences in Many-Objective Problems by Sampling
+  the Weighted Hypervolume}.
+\newblock In G.~Raidl et~al., editors, \emph{Genetic and Evolutionary
+  Computation Conference (GECCO 2009)}, pages 555--562, New York, NY, USA,
+  2009. ACM.
+
+\bibitem[Bader and Zitzler(2011)]{bz2011a}
+J.~Bader and E.~Zitzler.
+\newblock {HypE: An Algorithm for Fast Hypervolume-Based Many-Objective
+  Optimization}.
+\newblock \emph{Evolutionary Computation}, 19\penalty0 (1):\penalty0 45–--76,
+  2011.
+
+\bibitem[Bader et~al.(2011)Bader, Brockhoff, Thiele, and Zitzler]{bbtz2011a}
+J.~Bader, D.~Brockhoff, L.~Thiele, and E.~Zitzler.
+\newblock {Directed Multiobjective Optimization Based on the Hypervolume
+  Indicator}.
+\newblock \emph{Journal of Multi-Criteria Decision Analysis}, 2011.
+\newblock submitted.
+
+\bibitem[Bianco et~al.(2009)Bianco, Caramia, and Giordani]{bcg2009a}
+L.~Bianco, M.~Caramia, and S.~Giordani.
+\newblock {A Bilevel Flow Model for Hazmat Transportation Network Design}.
+\newblock \emph{Transportation research Part C: Emerging technologies},
+  17\penalty0 (2):\penalty0 175--196, 2009.
+\newblock \doi{10.1016/j.trc.2008.10.001}.
+
+\bibitem[Bleuler et~al.(2001)Bleuler, Brack, Thiele, and Zitzler]{bbtz2001a}
+S.~Bleuler, M.~Brack, L.~Thiele, and E.~Zitzler.
+\newblock {Multiobjective Genetic Programming: Reducing Bloat by Using
+  {SPEA2}}.
+\newblock In \emph{Congress on Evolutionary Computation (CEC~2001)}, pages
+  536--543, Piscataway, NJ, 2001. IEEE Press.
+
+\bibitem[Bleuler et~al.(2003)Bleuler, Laumanns, Thiele, and Zitzler]{bltz2003a}
+S.~Bleuler, M.~Laumanns, L.~Thiele, and E.~Zitzler.
+\newblock {PISA---A Platform and Programming Language Independent Interface for
+  Search Algorithms}.
+\newblock In C.~M. Fonseca et~al., editors, \emph{Conference on Evolutionary
+  Multi-Criterion Optimization {(EMO~2003)}}, volume 2632 of \emph{LNCS}, pages
+  494--508, Berlin, 2003. Springer.
+
+\bibitem[Caramia and Dell'Olmo(2008)]{cd2008a}
+M.~Caramia and P.~Dell'Olmo.
+\newblock \emph{{Multi-objective Management in Freight Logistics: Increasing
+  Capacity, Service Level and Safety With Optimization Algorithms}}.
+\newblock Springer, 2008.
+
+\bibitem[{Coello Coello} et~al.(2007){Coello Coello}, {Lamont}, and {Van
+  Veldhuizen}]{cvl2007a}
+C.~A. {Coello Coello}, G.~B. {Lamont}, and D.~A. {Van Veldhuizen}.
+\newblock \emph{{Evolutionary Algorithms for Solving Multi-Objective
+  Problems}}.
+\newblock Springer, Berlin, Germany, 2007.
+
+\bibitem[Deb(2001)]{deb2001a}
+K.~Deb.
+\newblock \emph{{Multi-Objective Optimization Using Evolutionary Algorithms}}.
+\newblock Wiley, Chichester, UK, 2001.
+
+\bibitem[Deb et~al.(2002)Deb, Pratap, Agarwal, and Meyarivan]{dapm2002a}
+K.~Deb, A.~Pratap, S.~Agarwal, and T.~Meyarivan.
+\newblock {A Fast and Elitist Multiobjective Genetic Algorithm: NSGA-II}.
+\newblock \emph{IEEE Transactions on Evolutionary Computation}, 6\penalty0
+  (2):\penalty0 182--197, 2002.
+
+\bibitem[Dell'Olmo et~al.(2005)Dell'Olmo, Gentili, and Scozzari]{dgs2005a}
+P.~Dell'Olmo, M.~Gentili, and A.~Scozzari.
+\newblock {On Finding Dissimilar Pareto-optimal Paths}.
+\newblock \emph{European Journal of Operational Research}, 162\penalty0
+  (1):\penalty0 70--82, 2005.
+
+\bibitem[Erkut et~al.(2007)Erkut, Tjandra, and Verter]{etv2007a}
+E.~Erkut, S.~A. Tjandra, and V.~Verter.
+\newblock {Chapter 9: Hazardeous Materials Transportation}.
+\newblock In C.~Barnhart and G.~Laporte, editors, \emph{Handbooks in Operations
+  Research and Management Science}, volume~14, pages 539--621. Elsevier, 2007.
+
+\bibitem[Goldberg(1989)]{gold1989a}
+D.~E. Goldberg.
+\newblock \emph{{Genetic Algorithms in Search, Optimization, and Machine
+  Learning}}.
+\newblock Addison-Wesley, Reading, Massachusetts, 1989.
+
+\bibitem[Gopalan et~al.(1990)Gopalan, Kolluri, Batta, and Karwan]{gkbk1990a}
+R.~Gopalan, K.~S. Kolluri, R.~Batta, and M.~H. Karwan.
+\newblock {Modeling Equity of Risk in the Transportation of Hazardous
+  Materials}.
+\newblock \emph{Operations Research}, 38\penalty0 (6):\penalty0 961--975, 1990.
+
+\bibitem[Horoba(2009)]{horo2009a}
+C.~Horoba.
+\newblock {Analysis of a Simple Evolutionary Algorithm for the Multiobjective
+  Shortest Path Problem}.
+\newblock In \emph{Foundations of Genetic Algorithms (FOGA 2009)}, pages
+  113--120. ACM, 2009.
+
+\bibitem[Wagner et~al.(2007)Wagner, Beume, and Naujoks]{wbn2007a}
+T.~Wagner, N.~Beume, and B.~Naujoks.
+\newblock {Pareto-\nolinebreak[4], Aggregation\mbox{-}, and Indicator-based
+  Methods in Many-objective Optimization}.
+\newblock In S.~Obayashi et~al., editors, \emph{Conference on Evolutionary
+  Multi-Criterion Optimization (EMO~2007)}, volume 4403 of \emph{LNCS}, pages
+  742--756, Berlin Heidelberg, Germany, 2007. Springer.
+
+\bibitem[Zitzler et~al.(2003)Zitzler, Thiele, Laumanns, Fonseca, and {Grunert
+  da Fonseca}]{ztlf2003a}
+E.~Zitzler, L.~Thiele, M.~Laumanns, C.~M. Fonseca, and V.~{Grunert da Fonseca}.
+\newblock {Performance Assessment of Multiobjective Optimizers: An Analysis and
+  Review}.
+\newblock \emph{IEEE Transactions on Evolutionary Computation}, 7\penalty0
+  (2):\penalty0 117--132, 2003.
+
+\bibitem[Zitzler et~al.(2007)Zitzler, Brockhoff, and Thiele]{zbt2007a}
+E.~Zitzler, D.~Brockhoff, and L.~Thiele.
+\newblock {The Hypervolume Indicator Revisited: On the Design of
+  Pareto-compliant Indicators Via Weighted Integration}.
+\newblock In S.~Obayashi et~al., editors, \emph{Conference on Evolutionary
+  Multi-Criterion Optimization (EMO~2007)}, volume 4403 of \emph{LNCS}, pages
+  862--876, Berlin, 2007. Springer.
+
+\bibitem[Zitzler et~al.(2010)Zitzler, Thiele, and Bader]{ztb2010a}
+E.~Zitzler, L.~Thiele, and J.~Bader.
+\newblock {On Set-Based Multiobjective Optimization}.
+\newblock \emph{IEEE Transactions on Evolutionary Computation}, 14\penalty0
+  (1):\penalty0 58--79, 2010.
+
+\end{thebibliography}
 

### 2012-08-08 trunk/DAW_hazmatHypE/trunk/paper/reviewsAndAnswersRound1.tex
added LaTeX file for replying to reviewers' comments

%% This is file `elsarticle-template-1a-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
%%
\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% our packages and definitions: %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{url} 
\usepackage{color}
% \usepackage{dsfont}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{multirow}

\newcommand{\TODO}[1]{{\color{red}TODO: #1}}
\newcommand{\TODON}[1]{{\color{blue}TODO: #1}}
\newcommand{\COMMENTD}[1]{{\color{green}Dimo: #1}}
\newcommand{\COMMENTN}[1]{{\color{green}Nora: #1}}

\renewcommand{\topfraction}{.70}
\renewcommand{\textfraction}{.20}

\newcommand{\reviewer}[1]{\begin{flushleft}\noindent{\sloppy {\bf Comment: } #1}\end{flushleft}}
\newcommand{\answer}[1]{\noindent{\em {\bf Reply: } #1}\\[1em]}

\journal{DAM}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem\\ Responses to the Reviewers' Comments}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author{Dimo Brockhoff}

\author{Nora Touati-Moungla}

\begin{abstract}
tbd: thank reviewers for careful comments, etc. 
\end{abstract}

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Reviewer 1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\reviewer{
It is not very clear what the authors mean for ``minimization of the equity risk'' (see, e.g., abstract,
and Introduction (three times) ) or ``minimizing the distributed risk'' (see, e.g., Section 2.1). As far I
can understand, one wishes to maximize risk equity.
}
\answer{\TODO{tbd}}

\reviewer{
It is not clear why the authors talk about ``risk-equity constrained routing problem'' (see, e.g.,
Title, Introduction, Sections 2, 3, 3.5, and 4). In their multi-objective model, risk equity does not
appear as a constraint, but as one of the objectives of the problem.
}
\answer{\TODO{tbd}}

\reviewer{
There is a lack in reviewing existing approaches:\\
a) Besides the cited paper of Dell'Olmo et al. (2005), more recently other papers propose
approaches for finding Pareto-optimal dissimilar paths for hazmat routing. See, e.g.:

M. Caramia and S. Giordani. On the selection of k efficient paths by clustering techniques.
International Journal on Data Mining, Modelling and Management, 1(3):237.260, 2009.\\
M. Caramia, S. Giordani, and A. Iovanella. On the selection of k routes in multi-objective
hazmat route planning. IMA Journal of Management Mathematics, 21(3):239.251, 2010.

b) The model studied by the authors is multi-objective, but it considers hazmat routing from the
perspective of a single decision maker. Indeed global hazmat route planning involves in general
at least two stakeholders: the government authority from one side and the carriers from the other
side. The main concern for the government authority is to control the risk induced by hazmat
transportation over the population and the environment, while the main concern from the
carriersÅf point of view is minimizing transportation cost. Since, typically, the government
authority does not have the right to impose specific routes to hazmat carriers (contrarily to the
assumption made by single decision maker hazmat routing models like the one proposed by the
authors), it can only achieve hazmat transportation risk mitigation adopting policies that induce
hazmat carriers to select low risk routes. This has led researchers to study multi-decision-makers
models for hazmat routing. In particular, two classes of models are investigated in recent papers:
``hazmat transportation network design'''' and ``toll setting approaches''. The authors should
therefore recall also these classes of models for global hazmat route planning and cite the related
papers, e.g.:

B.Y. Kara and V. Verter. Designing a road network for hazardous materials transportation.
Transportation Science, 38(2):188--196, 2004.\\
E. Erkut and O. Alp. Designing a road network for dangerous goods shipments. Computers \&
Operations Research, 34(5):1389--1405, 2007.\\
E. Erkut and F. Gzara. Solving the hazmat transport network design problem. Computers \&
Operations Research, 35(7):2234--2247, 2008.\\
V. Verter and B.Y. Kara. A path-based approach for the hazardous network design problem.
Management Science, 54(1):29--40, 2008.\\
L. Bianco, M. Caramia, and S. Giordani. A bilevel flow model for hazmat transportation
network design. Transportation Research Part C, 17:175--196, 2009.\\
P. Marcotte, A. Mercier, G. Savard, and V. Verter. Toll policies for mitigating hazardous
materials transport risk. Transportation Science, 43(2): 228--243, 2009.
}
\answer{\TODO{tbd}}

\reviewer{
Model's parameter $r^c_{ij}$ is not defined.
}
\answer{\TODO{tbd}}

\reviewer{
Please clarify that with o.f. (3) one achieves the maximization of the distribution of the risk, i.e.
risk equity.
}
\answer{\TODO{tbd}}

\reviewer{
Note that (3) can be linearized by introducing a new variable (e.g., variable $\gamma$), the constraints $\sum_{c\in C} r^c_{ij}y^c_{ij} \leq \gamma$, for all $(i, j)\in A$, and by replacing (3) with min $\gamma$.
}
\answer{\TODO{tbd}}

\reviewer{
In Section 3.1, some details explaining the meaning of ``variable length representation'' should be
added.
}
\answer{\TODO{tbd}}

\reviewer{
In Section 3.3, it is not clear the meaning of ``length(p)''. As far I can understand it should be
intended as the number of vertices of path p.
}
\answer{\TODO{tbd}}

\reviewer{
The authors provide sufficient insights about initialization and variation of the solutions'
population, but not how the population is updated and the stopping criteria used. Please provide a
pseudo-code description of the whole (HypE) algorithm, in order to make the paper self-contained.
}
\answer{\TODO{tbd}}

\reviewer{
Concepts as hypervolume and hypervolume indicator (e.g., see Section 3.5) should be defined in
order to make the paper self-contained. This will allow to better understand the meaning and the
values specifying the reference points for the hypervolume indicator (e.g., see Section 4.1). In
particular, without providing additional details in Section 3.5, it is very difficult for the reader to
understand the implication of the choices made in Section 4.1.
}
\answer{\TODO{tbd}}

\reviewer{
Provide some details to clarify the statements at the end of Section 4.1.
}
\answer{\TODO{tbd}}

\reviewer{
Quantifying risk equity is not clear. Are risk equity values computed according to o.f. (3), i.e., in
terms of maximum link total risk? Please clarify.
}
\answer{\TODO{tbd}}

\reviewer{
Figure 3 is too small to appreciate the depicted points. Moreover, and again, the definition of
equity risk is misleading: one wishes to maximize risk equity, while from the figure it seems that it
is preferable to have small risk equity. Of course, if in the figure ``risk equity'' is replaced with
``maximum link total risk'' the figure will make sense. The same applies to Figures 5, 6, 8, and 10.
}
\answer{\TODO{tbd}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Reviewer 2}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\reviewer{
First of all, the reviewer believes that the paper is in
general written well.  The research idea is clearly delivered and
the method and computation results are well described.

The paper presents a specific implementation of EMO algorithms,
namely, HypE, for tackling the multiobjective routing problem, in
which one of the objectives is the min-max type.

A major deficiency of this paper is that the literature review is far
from sufficiency, either for multiobjective routing problems or for
hazardous materials transportation problems.  A quite complete
review of multiobjective shortest path problems is contained in Xie
and Waller (2012), which provides a list of multiobjective routing
papers and might give the authors some clues, for example.  An
example for hazardous materials transportation is Lindner-Dutton et
al (1991), who presents a multiobjective hazardous materials
transportation problem, which includes a risk equity objective. Here
are just two examples the reviewer randomly found. Many other papers
should be consulted and included appropriately in the literature
review.  The authors should give a more comprehensive and enhanced
discussion on the relevant literature in the revised version of the
paper.

Lindner-Dutton, L., Batta, R. and Karwan, M.H. (1991). "Equitable
sequencing of a given set of hazardous materials shipments."
Transportation Science, 25(2), 1991.

Xie, C. and Waller, S.T. (2012). Parametric search and problem
decomposition for approximating Pareto-optimal paths. Transportation
Research Part B. (In press)
}
\answer{\TODO{tbd}}

\reviewer{ 
Another concern is about the EMO algorithm's performance.  The
authors should give a comparative evaluation, with an exact method
as the benchmark, at least on a small example problem.  Otherwise,
readers have no idea about how much the Pareto-optimal solution set
is approximated.
}
\answer{tbd}

% 
 \bibliographystyle{plainnat}
 \footnotesize
 \bibliography{all}

\end{document}

%%
%% End of file `elsarticle-template-1a-num.tex'.
